### 2011-09-25 proposal/proposal.tex
Empty file for project proposal.

\documentclass{article}

\title{15-744 Project Proposal}
\author{Carlo Angiuli and Michael Sullivan}
\date{September 28, 2011}

\begin{document}
\maketitle

% Needs to include:
%  - problem statement
%  - state-of-the-art
%  - work plan / milestones

\end{document}

### 2011-09-26 proposal/proposal.tex
Started sketching out project proposal; added BibTeX file.

@@
 \documentclass{article}
+\usepackage{cite}
+\renewcommand{\t}{\texttt}
 
 \title{15-744 Project Proposal}
-\author{Carlo Angiuli and Michael Sullivan}
+\author{Carlo Angiuli \and Michael Sullivan}
 \date{September 28, 2011}
@@
 
+\section{Motivation}
+
+It's good for network applications to be able to support many simultaneous
+connections---this is known as the C10K (``10,000 client'') problem\cite{c10k}.
+
+Simultaneous client connections require an application to maintain multiple
+states at once. A classic way to do this is to spawn a thread for each client.
+However, this approach does not scale as the number of clients grows.
+
+(nonblocking?)
+(lightweight threads?)
+(cite Erlang)
+
+A lightweight method is to use \t{epoll} or \t{kquery}.
+This requires the application programmer to manually maintain a state machine
+with
+
+Requires a lot of complex but straightforward code which is difficult to debug.
+Unsurprisingly, there exist many thin C/C++ wrappers, but these still require
+...
+
+and provide no assurance that the program ...
+
+\section{Proposal}
+
+We present a domain-specific language which provides a simple
+
+easily extensible with more protocol definitions
+
+outputs C code to be linked with the rest of the server code.
+
+formal guarantees: given our abstract specification of \t{epoll}, our DSL
+generates code with the following properties...
+
+embedded in Haskell: natively allows functional metaprogramming,
+leverages native type system
+
+\section{Milestones}
+
+\bibliography{citations}{}
+\bibliographystyle{plain}
+
 \end{document}

### 2011-09-26 proposal/proposal.tex
Some tweaks to proposal draft.

@@
 
-A lightweight method is to use \t{epoll} or \t{kquery}.
-This requires the application programmer to manually maintain a state machine
-with
-
-Requires a lot of complex but straightforward code which is difficult to debug.
-Unsurprisingly, there exist many thin C/C++ wrappers, but these still require
-...
+A lightweight method is to use nonblocking IO and a multiplexing
+facility such as \t{select}, \t{poll}, \t{epoll} or \t{kquery}.  This
+requires the application programmer to structure the code as a state
+machine, which is tedious, time-consuming, and error-prone. 
+Unsurprisingly, there exist many thin
+C/C++ wrappers, but these still require ...
 
@@
 easily extensible with more protocol definitions
+
+generates an explicit state-machine representation of the application logic
 

### 2011-09-28 proposal/proposal.tex
Mostly fleshed out the proposal.

--- 2011-09-25 proposal/proposal.tex
@@
 
+\section{Motivation}
+
+Ideally, network applications should be able to support many simultaneous
+clients---this is known as the C10K (``10,000 client'') problem\cite{c10k}.
+Maintaining connections simultaneously requires an application to track each
+client's state independently. A straightforward way to do this is to spawn a
+thread for each client. However, this approach does not scale as the number of
+clients grows.
+
+A lightweight approach is to use kernel-supported nonblocking IO with
+multiplexing such as \t{select}, \t{poll}, \t{epoll}, or \t{kquery}. This
+require the application programmer to structure the code as a state machine,
+which is tedious and error-prone. Unsurprisingly, there exist many thin C/C++
+wrappers for these facilities, but these still require structuring the
+application code as a state machine.
+
+To maintain the illusion of (?) client-specific state (?) languages like Erlang
+and Haskell provide lightweight thread mechanisms built on a user-space
+scheduler with nonblocking IO. While this solution is significantly better, the
+full generality of this approach still imposes unnecessary overhead on
+networking applications which don't require preemptable, independently scheduled
+threads.
+(cite Erlang)
+
+and provide no assurance that the program ...
+
+\section{Proposal}
+
+Hence, we intend to produce an embedded domain-specific language in Haskell for
+describing multi-client network applications.
+
+This language will provide composable, high-level abstractions for network
+operations which permits programmers to write server logic naturally, as if the
+application were threaded, while automatically synthesizing a state machine
+representation for use with a nonblocking main loop (as necessary for
+\t{epoll}).
+
+We will additionally be able to output a graphical representation of the
+underlying state machine logic.
+
+The back-end of this system could be implemented via several different options
+which we are still evaluating. One option is to directly interface with the
+\t{epoll} hooks in the Haskell runtime system, bypassing the native lightweight
+threading mechanism. Another would be to generate C source code which itself
+implements the network application.
+
+optionally linkable with native C code.
+
+easily extensible with more protocol definitions
+
+embedded in Haskell: natively allows functional metaprogramming,
+leverages native type system
+
+\section{Milestones}
+
+We will first collect a representative sample of network applications and
+examine the range of high-level primitives required to support them. We will
+then decide on a specific back-end for our language, and begin implementing it.
+To evaluate the utility of this language, we will build a number of
+representative servers in our language (e.g., echo server, simple HTTP server)
+and compare them to traditional implementations with regard to code simplicity
+and performance.
+
+\bibliography{citations}{}
+\bibliographystyle{plain}
+
 \end{document}

### 2011-09-28 proposal/proposal.tex
Finished (tentatively) project proposal.

--- 2011-09-26 proposal/proposal.tex
@@
 
-It's good for network applications to be able to support many simultaneous
-connections---this is known as the C10K (``10,000 client'') problem\cite{c10k}.
+Modern network applications need to support many simultaneous clients---this is
+known as the C10K (``10,000 client'') problem\cite{Kegel}. But to maintain these
+simultaneous connections, an application must track each client's state
+independently. A straightforward way to do this is to spawn a thread for each
+client, but this approach does not scale as the number of clients grows.
 
-Simultaneous client connections require an application to maintain multiple
-states at once. A classic way to do this is to spawn a thread for each client.
-However, this approach does not scale as the number of clients grows.
+A lightweight approach is to use kernel mechanisms for multiplexed, non-blocking
+I/O, such as \t{select}, \t{poll}, \t{epoll}, or \t{kquery}; unfortunately, to
+maintain separate state for each client, application programmers must manually
+structure their code as state machines, which is tedious and error-prone.
+Unsurprisingly, there exist many thin C/C++ wrappers for these facilities, but
+none avoid major code refactoring.
 
-(nonblocking?)
-(lightweight threads?)
-(cite Erlang)
-
-A lightweight method is to use \t{epoll} or \t{kquery}.
-This requires the application programmer to manually maintain a state machine
-with
-
-Requires a lot of complex but straightforward code which is difficult to debug.
-Unsurprisingly, there exist many thin C/C++ wrappers, but these still require
-...
-
-and provide no assurance that the program ...
+To maintain the illusion of threads (in particular, client-specific state),
+languages like Erlang and Haskell provide lightweight threading via user-space
+schedulers and non-blocking IO. While this affords better performance, the full
+generality of this approach still imposes unnecessary overhead on networking
+applications which don't require preemptable, independently scheduled threads
+\cite{Vinoski}.
 
@@
 
-We present a domain-specific language which provides a simple
+Hence, we intend to produce an embedded domain-specific language in Haskell to
+describe concurrent network applications. This language will provide composable,
+high-level network operations, in which programmers may express their
+application logic naturally, as if using threads.
 
-easily extensible with more protocol definitions
+However, this high-level description will be automatically synthesized into a
+state machine representation usable in a C loop invoking \t{epoll}, using no
+threads whatsoever. We will additionally output a graphical representation of
+this underlying state machine.
 
-outputs C code to be linked with the rest of the server code.
-
-formal guarantees: given our abstract specification of \t{epoll}, our DSL
-generates code with the following properties...
-
-embedded in Haskell: natively allows functional metaprogramming,
-leverages native type system
+The back-end of this system can be implemented via several different options
+which we are still evaluating. One option is to directly interface with the
+\t{epoll} hooks in the Haskell runtime system, bypassing the native lightweight
+threads. Another option is to generate C source code which itself implements the
+network application, optionally linkable with native C code.
 
 \section{Milestones}
+
+We will first collect a representative sample of Linux network applications and
+examine the range of high-level primitives required to support them. We will
+then decide on a specific back-end for our language, and design the front- and
+back-end in tandem. To evaluate the utility of this language, we will build a
+number of representative applications in our language (e.g., an echo server,
+simple HTTP server) and compare them to traditional implementations, with
+respect to both code simplicity and performance.
 

### 2011-10-30 src/CLib.hs
Finished Language.C functionality.

@@ -29,7 +29,11 @@ cUnion id = cStructUnion CUnionTag id Nothing
 cStructUnion tag id decls = CTypeSpec $ CSUType
   (CStruct tag (Just $ ident id) decls [] undefNode) undefNode
 
--- TODO enums
+-- Enums
+cEnum id = cEnum' id Nothing
+
+cEnum' id enumls = CTypeSpec $ CEnumType
+  (CEnum (Just $ ident id) enumls [] undefNode) undefNode
 
 -- }}}
 -- Declarations {{{
@@ -58,8 +62,8 @@ cTypedef indirs typ id = CDeclExt $ cDecl'
 -- Structure declarations
 cStructDecl id decls = cTypeDecl [] $ cStructUnion CStructTag id (Just decls)
 cUnionDecl id decls = cTypeDecl [] $ cStructUnion CUnionTag id (Just decls)
-
--- Parameter declarations?
+cEnumDecl id idexps = cTypeDecl [] $ cEnum' id (Just ls)
+  where ls = map (\(id,expr) -> (ident id,expr)) idexps
 
 -- Declarators
 cDeclr id indirs = CDeclr (Just $ ident id) indirs Nothing [] undefNode
@@ -69,7 +73,6 @@ cAbstractDeclr indirs = CDeclr Nothing indirs Nothing [] undefNode
 cPtr = CPtrDeclr [] undefNode
 cArray Nothing = CArrDeclr [] (CNoArrSize True) undefNode
 cArray (Just expr) = CArrDeclr [] (CArrSize False expr) undefNode
-cFun params = CFunDeclr (Right (params,False)) [] undefNode
 
 -- Storage specifiers
 cStatic = CStorageSpec $ CStatic undefNode
@@ -77,7 +80,9 @@ cExtern = CStorageSpec $ CExtern undefNode
 
 -- Declarators
 cFunction id params retTyp body = CFDefExt $ CFunDef
-  [retTyp] (cDeclr id [cFun params]) [] body undefNode
+  [retTyp] (cDeclr id [cFun $ map f params]) [] body undefNode where
+  cFun params = CFunDeclr (Right (params,False)) [] undefNode
+  f (typ,id) = cDecl' [typ] [] id Nothing
 
 -- Initializers
 cInit expr = CInitExpr expr undefNode
@@ -124,7 +129,7 @@ cOr = cBinaryOp COrOp
 cLand = cBinaryOp CLndOp
[…]

### 2011-10-30 src/CLib.hs
Some cleanup.

@@ -1,4 +1,5 @@
 -- Wrappers for Language.C
+-- Only functions starting with 'c' should be used externally.
 
 module CLib where
 
@@ -7,7 +8,7 @@ import qualified Language.C
 
 -- Miscellaneous {{{
 
-cTopLevel ns = CTranslUnit ns undefNode
+cFile ns = CTranslUnit ns undefNode
 ident = internalIdent
 
 -- }}}
@@ -23,46 +24,43 @@ cUnsig = CTypeSpec $ CUnsigType undefNode
 cBool = CTypeSpec $ CBoolType undefNode
 
 -- Structs and unions
-cStruct id = cStructUnion CStructTag id Nothing
-cUnion id = cStructUnion CUnionTag id Nothing
+cStruct id = structUnion CStructTag id Nothing
+cUnion id = structUnion CUnionTag id Nothing
 
-cStructUnion tag id decls = CTypeSpec $ CSUType
+structUnion tag id decls = CTypeSpec $ CSUType
   (CStruct tag (Just $ ident id) decls [] undefNode) undefNode
 
 -- Enums
-cEnum id = cEnum' id Nothing
+cEnum id = enum id Nothing
 
-cEnum' id enumls = CTypeSpec $ CEnumType
+enum id enumls = CTypeSpec $ CEnumType
   (CEnum (Just $ ident id) enumls [] undefNode) undefNode
 
 -- }}}
 -- Declarations {{{
 
-cDecl typ = cDecl' [typ]
-cDeclStatic typ = cDecl' [cStatic,typ]
-cDeclExtern typ = cDecl' [cExtern,typ]
+cDecl typ = decl [typ]
+cDeclStatic typ = decl [cStatic,typ]
+cDeclExtern typ = decl [cExtern,typ]
 
-cDecl' typSpecs indirs id init = CDecl typSpecs
+cDeclExt = CDeclExt
+
+decl typSpecs indirs id init = CDecl typSpecs
   [(Just $ cDeclr id indirs,initExpr,Nothing)] undefNode
   where initExpr = init >>= (Just . cInit)
 
--- Top-level declarations
-cTopDecl typ indirs id init = CDeclExt $ cDecl typ indirs id init
-cTopDeclStatic typ indirs id init = CDeclExt $ cDeclStatic typ indirs id init
[…]

### 2011-11-13 src/Lang.hs
Started mocking up language.

@@ -1,0 +1,38 @@
+module Lang where
+
+{-
+Echo server:
+
+setup -> accept/spawn ->
+  thread_setup ->
+  read -> write | die
+  write -> write | read | die
+
+main = do
+  fd <- var FD "fd" (setup_listener port)
+  e <- var Event "e" (reg_event fd READ)
+  while true { do
+    fd' <- call accept fd
+    spawn child_code fd'
+  }
+
+child_code = Thread [(Int,fd)] $ do
+  e <- var Event "e" (setup_connection fd)
+  buf <- var Buffer "buf" (new_buf 4096)
+  while true { do
+    amount_read <- read e buf 4096
+    write e buf amount_read
+  }
+
+-}
+
+type Prgm = [Stmt]
+type Var = String
+data Type = Int | Bool | ThreadT | FD | Buffer | Event
+data Stmt = Decl Type Var Expr 
+          | While Expr [Stmt]
+          | If Expr [Stmt] [Stmt] [Stmt]
+          | Assign Var Expr
+data Expr = Call Prim 
+          | Thread [Stmt]
+data Prim = Unit

### 2011-11-13 src/rt/echo_handwritten.c

@@ -158,11 +158,11 @@ bool setup(struct thread_t *thread)
 	make_socket_non_blocking(e->fd);
 
 	memset(&servaddr, 0, sizeof(servaddr));
-    servaddr.sin_family      = AF_INET;
-    servaddr.sin_addr.s_addr = htonl(INADDR_ANY);
-    servaddr.sin_port        = htons(port);
+	servaddr.sin_family      = AF_INET;
+	servaddr.sin_addr.s_addr = htonl(INADDR_ANY);
+	servaddr.sin_port        = htons(port);
 	
-    if (bind(e->fd, (struct sockaddr *)&servaddr,
+	if (bind(e->fd, (struct sockaddr *)&servaddr,
 	         sizeof(servaddr)) < 0)
 		fail(1, "bind");
 
@@ -170,7 +170,7 @@ bool setup(struct thread_t *thread)
 
 	e->event = mk_event();
 	e->event->id = e->fd;
-	
+
 	if (epoll_ctler(EPOLL_CTL_ADD, e->fd, EPOLLIN|EPOLLET, e->event) < 0)
 		fail(1, "epoll_ctl");
 

### 2011-11-13 src/Lang.hs
Add more to the language.

@@ -26,13 +26,42 @@ child_code = Thread [(Int,fd)] $ do
 
 -}
 
-type Prgm = [Stmt]
+type Block = [Stmt]
 type Var = String
-data Type = Int | Bool | ThreadT | FD | Buffer | Event
-data Stmt = Decl Type Var Expr 
-          | While Expr [Stmt]
-          | If Expr [Stmt] [Stmt] [Stmt]
+type VDecl = (Var, Type)
+type ThreadCode = ([VDecl], Block)
+
+data Type = Int | Bool | String | FD | Buffer | Event -- | ThreadT
+          deriving (Eq, Ord, Show)
+data Stmt = Decl VDecl Expr
+          | While Expr Block
+          | If Expr Block Block
           | Assign Var Expr
-data Expr = Call Prim 
-          | Thread [Stmt]
-data Prim = Unit
+          | Wait Expr
+          | Exit
+          deriving (Eq, Ord, Show)
+data Expr = Call Prim [Expr]
+          | Thread ThreadCode [Expr]
+          | Arith ArithOp Expr Expr
+          | ArithUnop ArithUnop Expr
+          | NumLit Integer
+          | StringLit String
+          deriving (Eq, Ord, Show)
+data ArithOp = Plus | Times | Minus | Div | Mod
+             | And | Or | Xor
+             | Rsh | Lsh
+             deriving (Eq, Ord, Show)
+data ArithUnop = Negate | Not
+               deriving (Eq, Ord, Show)
+data Prim = CFn String
+          deriving (Eq, Ord, Show)
+
+
+instance Num Expr where
+  fromInteger = NumLit
+  (+) = Arith Plus
+  (-) = Arith Minus
+  (*) = Arith Times
+  negate = ArithUnop Negate
+  abs = error "full of lies"
+  signum = error "full of lies"

### 2011-11-13 src/Lang.hs
Added in Var.

@@ -46,6 +46,7 @@ data Expr = Call Prim [Expr]
           | ArithUnop ArithUnop Expr
           | NumLit Integer
           | StringLit String
+          | Var Var
           deriving (Eq, Ord, Show)
 data ArithOp = Plus | Times | Minus | Div | Mod
              | And | Or | Xor

### 2011-11-13 src/Lang.hs
Tweaks to Lang

@@ -36,7 +36,8 @@ data Type = Int | Bool | String | FD | Buffer | Event -- | ThreadT
 data Stmt = Decl VDecl Expr
           | While Expr Block
           | If Expr Block Block
-          | Assign Var Expr
+          | Assign Expr Expr
+          | Exp Expr
           | Wait Expr
           | Exit
           deriving (Eq, Ord, Show)

### 2011-11-13 src/Back.hs
Started backend work.

@@ -1,0 +1,22 @@
+module Back where
+
+import qualified Lang
+
+type Prgm' = [Block']
+
+type BlockName = String
+type Block = (BlockName, [Lang.VDecl], [Stmt], Tail)
+
+data Stmt = Decl VDecl Lang.Expr
+          | Assign Var Lang.Expr
+          | Spawn BlockName [Expr]
+          | Exp Lang.Expr
+          deriving (Eq, Ord, Show)
+
+type Tail = If Lang.Expr Tail Tail
+          | Goto BlockName
+          | GotoWait BlockName
+          | Exit
+          deriving (Eq, Ord, Show)
+
+

### 2011-11-13 src/Lang.hs

@@ -37,11 +37,11 @@ data Stmt = Decl VDecl Expr
           | While Expr Block
           | If Expr Block Block
           | Assign Var Expr
+          | Spawn ThreadCode [Expr]
           | Wait Expr
           | Exit
           deriving (Eq, Ord, Show)
 data Expr = Call Prim [Expr]
-          | Thread ThreadCode [Expr]
           | Arith ArithOp Expr Expr
           | ArithUnop ArithUnop Expr
           | NumLit Integer

### 2011-12-04 src/Echo.hs
Major restructuring to limit our exports, part 1.

@@ -1,11 +1,6 @@
 module Echo where
 
-import Lang
-import Sugar
-import Lib
-import Pretty (pretty)
-import Back (backend)
-import Codegen (codegen)
+import Language.Foo
 
 q_limit = 1024
 port = 2023
@@ -41,10 +36,4 @@ main_loop = do
   while 1 $ do
     fd' <- accept fd e
     spawn child_code [fd']
-
-compile file p = do
-  writeFile file (codegen $ backend $ desugar p)
-
-main_loop_all = codegen $ backend $ desugar main_loop
-main_loop_back = pretty $ fst . backend $ desugar main_loop
 

### 2011-12-04 src/Language/Foo.hs

@@ -1,0 +1,21 @@
+module Language.Foo
+  (module Language.Foo.Lib,
+   module Language.Foo.Sugar,
+   module Language.Foo.Lang,
+   compile,
+   testFront, testBack, testAll)
+  where
+
+import Language.Foo.Lang
+import Language.Foo.Lib
+import Language.Foo.Sugar -- (desugar)
+import Language.Foo.Back (backend)
+import Language.Foo.Codegen (codegen)
+import Language.Foo.Pretty
+
+compile file p = do writeFile file (codegen $ backend $ desugar p)
+
+testFront = pretty . desugar
+testBack = pretty . fst . backend . desugar
+testAll = codegen . backend . desugar
+

### 2011-12-04 src/Language/Foo/Back.hs

@@ -1,6 +1,6 @@
-module Back where
+module Language.Foo.Back where
 
-import qualified Lang
+import qualified Language.Foo.Lang as Lang
 import Control.Monad.RWS
 import Control.Monad.Writer
 import qualified Data.Set as Set

### 2011-12-04 src/Language/Foo/Codegen.hs

@@ -1,8 +1,8 @@
-module Codegen where
+module Language.Foo.Codegen where
 
-import Lang hiding(Stmt,Decl,While,If,Spawn,Assign,Exp,Wait,Exit,Block)
-import Back
-import CLib
+import Language.Foo.Lang hiding(Stmt,Decl,While,If,Spawn,Assign,Exp,Wait,Exit,Block)
+import Language.Foo.Back
+import Language.Foo.CLib
 import Data.List
 import Data.Maybe
 

### 2011-12-04 src/Language/Foo/Lang.hs

@@ -1,4 +1,4 @@
-module Lang where
+module Language.Foo.Lang where
 
 type Block = [Stmt]
 type Var = String

### 2011-12-04 src/Language/Foo/Lib.hs

@@ -1,7 +1,7 @@
-module Lib where
+module Language.Foo.Lib where
 
-import Lang
-import Sugar
+import Language.Foo.Lang
+import Language.Foo.Sugar
 
 -- Coreish runtime functions
 mk_nb_event fd modes =

### 2011-12-04 src/Language/Foo/Pretty.hs

@@ -1,8 +1,8 @@
 {-# LANGUAGE FlexibleInstances, TypeSynonymInstances #-}
-module Pretty where
+module Language.Foo.Pretty (pretty) where
 
-import qualified Lang
-import Back
+import qualified Language.Foo.Lang as Lang
+import Language.Foo.Back
 import Text.PrettyPrint
 import Data.List
 import qualified Language.C

### 2011-12-04 src/Language/Foo/Sugar.hs

@@ -1,6 +1,6 @@
-module Sugar where
+module Language.Foo.Sugar where
 
-import Lang
+import Language.Foo.Lang
 import Control.Monad.RWS
 
 type Prog = RWS () [Stmt] Int

### 2011-12-04 src/Test.hs

@@ -1,28 +1,0 @@
-module Test where
-
-import Lang
-import Back
-import Sugar
-import Lib
-import Codegen
-import Pretty
-
-testFront = pretty . desugar
-testBack = pretty . fst . backend . desugar
-testAll = pretty . codegen . backend . desugar
-
-main_loop0 = do
-  fd <- var "fd" Int 12
-  print_int fd
-  exit
-
-main_loop = do
-  fd <- var "fd" Int 12
-  ifE fd (
-    do print_int 7) (
-    do print_int fd)
-  exit
-
-main_loop_code = testFront main_loop
-main_loop_back = testBack main_loop
-main_loop_all = testAll main_loop

### 2011-12-04 src/Echo.hs
Package event and fd together for our things.

@@ -11,11 +11,11 @@ q_limit = 1024
 port = 2023
 bufsize = 4096
 
-setup_connection :: Expr -> Prog Expr
+setup_connection :: Expr -> Prog (Expr, Expr)
 setup_connection fd = do
   make_nb fd
   e <- mk_nb_event fd kEVENT_RDWR
-  return e
+  return (fd, e)
 
 setup_listener :: Expr -> Prog (Expr, Expr)
 setup_listener port = do
@@ -28,18 +28,18 @@ setup_listener port = do
 
 child_code = declare_thread [("child_fd", Int)] $
   \[child_fd] -> do
-  e <- (setup_connection child_fd)
+  ev <- setup_connection child_fd
   buf <- new_buf bufsize
   while 1 $ do
-    amount_read <- do_read child_fd e buf bufsize
+    amount_read <- do_read ev buf bufsize
     ifE' (amount_read .== 0) exit
-    amount_written <- full_write child_fd e buf amount_read
+    amount_written <- full_write ev buf amount_read
     ifE' (amount_written .< amount_read) exit
 
 main_loop = do
-  (fd, e) <- setup_listener port
+  ev <- setup_listener port
   while 1 $ do
-    fd' <- accept fd e
+    fd' <- accept ev
     spawn child_code [fd']
 
 compile file p = do

### 2011-12-04 src/Lib.hs

@@ -26,11 +26,20 @@ sock_write fd buf len =
   callName "amt_written" (CFn "write") Int [fd, buf, len]
 close fd =
   call (CFn "close") Int [fd]
+open path flags =
+  call (CFn "open") Int [path, flags]
+
+file_read fd buf len =
+  callName "amt_read" (CFn "read") Int [fd, buf, len]
+file_write fd buf len =
+  callName "amt_written" (CFn "write") Int [fd, buf, len]
+
 
 -- For debugging
 print_int n  =
   call (CFn "print_int") Int [n]
 
+errno = Var "errno" -- weeeee
 
 -- TODO: a bunch more
 
@@ -42,6 +51,10 @@ kEVENT_RD = Constant "EVENT_RD"
 kEVENT_WR = Constant "EVENT_WR"
 kEVENT_RDWR = Constant "EVENT_RDWR"
 
+kO_RDONLY = Constant "O_RDONLY"
+
+kEAGAIN = Constant "EAGAIN"
+
 
 do_nb_action :: Expr -> Prog Expr -> Prog Expr
 do_nb_action e action = do
@@ -52,15 +65,15 @@ do_nb_action e action = do
       res .=. action
   return res
 
-accept fd e = do_nb_action e (sock_accept fd)
-do_read fd e buf size = do_nb_action e (sock_read fd buf size)
-do_write fd e buf size = do_nb_action e (sock_write fd buf size)
+accept (fd, e) = do_nb_action e (sock_accept fd)
+do_read (fd, e) buf size = do_nb_action e (sock_read fd buf size)
+do_write (fd, e) buf size = do_nb_action e (sock_write fd buf size)
 
-full_write fd e buf size = do
+full_write ev buf size = do
   amt_written <- var "total_written" Int 0
   failed <- var "write_failed" Int 0
   while (amt_written .< size .&& failed .== 0) $ do
-    amt <- do_write fd e (buf + amt_written) (size - amt_written)
+    amt <- do_write ev (buf + amt_written) (size - amt_written)
     amt_written .= amt_written + amt
     ifE' (amt .== 0) $ do failed .= 1
   return amt_written

### 2011-12-04 src/Language/Foo.hs
Major restructuring, part 2.

@@ -1,16 +1,15 @@
 module Language.Foo
-  (module Language.Foo.Lib,
-   module Language.Foo.Sugar,
-   module Language.Foo.Lang,
+  (module Language.Foo.Syntax,
+   module Language.Foo.Lib,
    compile,
    testFront, testBack, testAll)
   where
 
-import Language.Foo.Lang
+import Language.Foo.Syntax
 import Language.Foo.Lib
-import Language.Foo.Sugar -- (desugar)
-import Language.Foo.Back (backend)
-import Language.Foo.Codegen (codegen)
+
+import Language.Foo.Backend
+import Language.Foo.Codegen
 import Language.Foo.Pretty
 
 compile file p = do writeFile file (codegen $ backend $ desugar p)

### 2011-12-04 src/Language/Foo/Backend.hs

@@ -1,6 +1,10 @@
-module Language.Foo.Back where
+module Language.Foo.Backend
+  (Stmt(..), Tail(..), Block, Thread,
+   backend)
+  where
 
-import qualified Language.Foo.Lang as Lang
+import qualified Language.Foo.Syntax as Front
+  (Stmt(..), Expr(..), VDecl, Prim(..), Block, Var)
 import Control.Monad.RWS
 import Control.Monad.Writer
 import qualified Data.Set as Set
@@ -11,21 +15,21 @@ type Prgm = [Block]
 type Label = Int
 type ThreadName = Label
 type Block = (Label, ThreadName, [Stmt], Tail)
-type Thread = (ThreadName, [Lang.VDecl])
+type Thread = (ThreadName, [Front.VDecl])
 
-data Stmt = Decl Lang.VDecl Lang.Expr
-          | Assign Lang.Expr Lang.Expr
-          | Exp Lang.Expr
-          | Spawn Label [(Lang.VDecl, Lang.Expr)]
+data Stmt = Decl Front.VDecl Front.Expr
+          | Assign Front.Expr Front.Expr
+          | Exp Front.Expr
+          | Spawn Label [(Front.VDecl, Front.Expr)]
           deriving (Eq, Ord, Show)
 
-data Tail = If Lang.Expr Tail Tail
+data Tail = If Front.Expr Tail Tail
           | Goto Label
           | GotoWait Label
           | Exit
           deriving (Eq, Ord, Show)
 
-backend :: Lang.Block -> ([Block],[Thread])
+backend :: Front.Block -> ([Block],[Thread])
 backend = collectFrees . (mapFst optimizeJumps) . flattenPrgm
 
 mapFst f (x,y) = (f x,y)
@@ -37,9 +41,9 @@ type Flattener = RWS ThreadName ([Block],[Thread]) Label
 flattenPrgm stmts = ((0,0,s,t):blocks,(0,[]):thds) where
   ((s,t),_,(blocks,thds)) = runRWS (flattenStmts stmts [] Exit) 0 1
   
-registerEvent event = [Lang.Exp $
-                       Lang.Call (Lang.CFn "register_event")
-                       [Lang.CurThread, event]]
+registerEvent event = [Front.Exp $
+                       Front.Call (Front.CFn "register_event")
+                       [Front.CurThread, event]]
 
 fresh = do
   x <- get
[…]

### 2011-12-04 src/Language/Foo/Codegen.hs

@@ -1,7 +1,9 @@
-module Language.Foo.Codegen where
+module Language.Foo.Codegen (codegen) where
 
-import Language.Foo.Lang hiding(Stmt,Decl,While,If,Spawn,Assign,Exp,Wait,Exit,Block)
-import Language.Foo.Back
+import Language.Foo.Syntax
+  (Prim(..), ArithOp(..), ArithUnop(..), RelnOp(..), Type(..), Expr(..))
+import Language.Foo.Backend
+  (Block, Thread, Stmt(..), Tail(..))
 import Language.Foo.CLib
 import Data.List
 import Data.Maybe

### 2011-12-04 src/Language/Foo/Lang.hs

@@ -1,49 +1,0 @@
-module Language.Foo.Lang where
-
-type Block = [Stmt]
-type Var = String
-type VDecl = (Var, Type)
-type ThreadCode = ([VDecl], Block)
-
-data Stmt = Decl VDecl Expr
-          | While Expr Block
-          | If Expr Block Block
-          | Spawn ThreadCode [Expr]
-          | Assign Expr Expr
-          | Exp Expr
-          | Wait Expr
-          | Exit
-          deriving (Eq, Ord, Show)
-
-data Type = Int | Bool | String | FD | Buffer | Event -- | ThreadT
-          deriving (Eq, Ord, Show)
-data Expr = Call Prim [Expr]
-          | Arith ArithOp Expr Expr
-          | ArithUnop ArithUnop Expr
-          | RelnOp RelnOp Expr Expr
-          | Constant String
-          | NumLit Integer
-          | StringLit String
-          | Var Var
-          | CurThread
-          deriving (Eq, Ord, Show)
-data ArithOp = Plus | Times | Minus | Div | Mod
-             | And | Or | Xor
-             | Rsh | Lsh
-             deriving (Eq, Ord, Show)
-data ArithUnop = Negate | Not
-               deriving (Eq, Ord, Show)
-data RelnOp = Eq | Less | Greater -- more
-             deriving (Eq, Ord, Show)
-data Prim = CFn String
-          deriving (Eq, Ord, Show)
-
-
-instance Num Expr where
-  fromInteger = NumLit
-  (+) = Arith Plus
-  (-) = Arith Minus
-  (*) = Arith Times
-  negate = ArithUnop Negate
-  abs = error "full of lies"
-  signum = error "full of lies"

### 2011-12-04 src/Language/Foo/Lib.hs

@@ -1,7 +1,6 @@
 module Language.Foo.Lib where
 
-import Language.Foo.Lang
-import Language.Foo.Sugar
+import Language.Foo.Syntax
 
 -- Coreish runtime functions
 mk_nb_event fd modes =

### 2011-12-04 src/Language/Foo/Pretty.hs

@@ -1,8 +1,8 @@
 {-# LANGUAGE FlexibleInstances, TypeSynonymInstances #-}
 module Language.Foo.Pretty (pretty) where
 
-import qualified Language.Foo.Lang as Lang
-import Language.Foo.Back
+import qualified Language.Foo.Syntax as Lang
+import Language.Foo.Backend
 import Text.PrettyPrint
 import Data.List
 import qualified Language.C

### 2011-12-04 src/Language/Foo/Syntax.hs

@@ -1,7 +1,58 @@
-module Language.Foo.Sugar where
+module Language.Foo.Syntax where
 
-import Language.Foo.Lang
 import Control.Monad.RWS
+
+--{{{ Front-end language
+
+type Block = [Stmt]
+type Var = String
+type VDecl = (Var, Type)
+type ThreadCode = ([VDecl], Block)
+
+data Stmt = Decl VDecl Expr
+          | While Expr Block
+          | If Expr Block Block
+          | Spawn ThreadCode [Expr]
+          | Assign Expr Expr
+          | Exp Expr
+          | Wait Expr
+          | Exit
+          deriving (Eq, Ord, Show)
+
+data Type = Int | Bool | String | FD | Buffer | Event -- | ThreadT
+          deriving (Eq, Ord, Show)
+data Expr = Call Prim [Expr]
+          | Arith ArithOp Expr Expr
+          | ArithUnop ArithUnop Expr
+          | RelnOp RelnOp Expr Expr
+          | Constant String
+          | NumLit Integer
+          | StringLit String
+          | Var Var
+          | CurThread
+          deriving (Eq, Ord, Show)
+data ArithOp = Plus | Times | Minus | Div | Mod
+             | And | Or | Xor
+             | Rsh | Lsh
+             deriving (Eq, Ord, Show)
+data ArithUnop = Negate | Not
+               deriving (Eq, Ord, Show)
+data RelnOp = Eq | Less | Greater -- more
+             deriving (Eq, Ord, Show)
+data Prim = CFn String
+          deriving (Eq, Ord, Show)
+
+instance Num Expr where
+  fromInteger = NumLit
+  (+) = Arith Plus
+  (-) = Arith Minus
+  (*) = Arith Times
+  negate = ArithUnop Negate
+  abs = error "full of lies"
+  signum = error "full of lies"
+
+--}}}
+--{{{ Monadic sugar
 
 type Prog = RWS () [Stmt] Int
 
@@ -51,7 +102,6 @@ callName name fn t args = var name t (Call fn args)
 call :: Prim -> Type -> [Expr] -> Prog Expr
 call fn t args = callName "tmp" fn t args
 
-
 spawn :: ThreadCode -> [Expr] -> Prog ()
 spawn thread args = add $ Spawn thread args
 
@@ -86,3 +136,5 @@ infix  4 .==, .<, .> --, ./=, .<=, .>=
[…]

### 2011-12-05 src/Echo.hs
Set SO_REUSEADDR.

@@ -15,6 +15,7 @@ setup_connection fd = do
 setup_listener :: Expr -> Prog (Expr, Expr)
 setup_listener port = do
   fd <- socket kAF_INET kSOCK_STREAM 0
+  set_sock_reuse fd
   make_nb fd
   sock_bind_v4 fd kINADDR_ANY port
   sock_listen fd q_limit

### 2011-12-05 src/Language/Foo/Lib.hs

@@ -12,6 +12,7 @@ new_buf size =
 -- Sugar for individual functions and whatnot
 socket domain typ protcol =
   callName "sock_fd" (CFn "socket") Int [domain, typ, protcol]
+set_sock_reuse fd = call (CFn "set_sock_reuse") Int [fd]
 make_nb fd = call (CFn "make_socket_non_blocking") Int [fd]
 sock_bind_v4 fd addr port =
   call (CFn "sock_bind_v4") Int [fd, addr, port]

### 2011-12-05 src/rt/lib.c

@@ -1,6 +1,13 @@
 #include "lib.h"
 #include "mainloop.h"
 
+int set_sock_reuse(int fd)
+{
+	int value = 1;
+	return setsockopt(fd, SOL_SOCKET, SO_REUSEADDR,
+	                  &value, sizeof(value));
+}
+
 int make_socket_non_blocking(int sfd)
 {
 	int flags, s;

### 2011-12-05 src/rt/lib.h

@@ -1,3 +1,4 @@
+#include <string.h>
 #include <unistd.h>
 #include <stdio.h>
 #include <sys/types.h>
@@ -18,6 +19,7 @@
 #define EVENT_WR (EPOLLOUT)
 #define EVENT_RDWR (EPOLLIN|EPOLLOUT)
 
+int set_sock_reuse(int fd);
 int make_socket_non_blocking(int sfd);
 int sock_bind_v4(int fd, int addr, int port);
 int print_int(int n);

### 2011-12-05 src/rt/mainloop.c
Actually free threads.

@@ -79,7 +79,6 @@ static void free_buf(buf_t *buf)
 
 void free_thread(thread_t *thread)
 {
-	return;
 	event_t *e;
 	while ((e = Q_GET_HEAD(&thread->nb_events))) {
 		Q_REMOVE(&thread->nb_events, e, gc_link);

### 2011-12-05 src/Http.hs
Write an HTTP server. Still needs aio for reads.

@@ -1,0 +1,68 @@
+module Echo where
+
+import Language.Foo
+
+q_limit = 1024
+port = 8080
+bufsize = 4096-1
+
+http_parse buf len =
+  callName "parse_result" (CFn "http_parse") Int [buf, len]
+
+response_header = "HTTP/1.0 200 OK\r\n\r\n"
+
+setup_connection :: Expr -> Prog (Expr, Expr)
+setup_connection fd = do
+  make_nb fd
+  e <- mk_nb_event fd kEVENT_RDWR
+  return (fd, e)
+
+setup_listener :: Expr -> Prog (Expr, Expr)
+setup_listener port = do
+  fd <- socket kAF_INET kSOCK_STREAM 0
+  set_sock_reuse fd  
+  make_nb fd
+  sock_bind_v4 fd kINADDR_ANY port
+  sock_listen fd q_limit
+  e <- mk_nb_event fd kEVENT_RD
+  return (fd, e)
+
+parse_request buf ev = do
+  parse_result <- var "parse_result" Int 0
+  request_size <- var "request_size" Int 0
+  while (parse_result .== 0) $ do
+    amt_read <- do_read ev buf (bufsize - request_size)
+    ifE' (amt_read .== 0) exit
+    request_size .= request_size + amt_read
+    parse_result .=. http_parse buf request_size
+  return parse_result
+
+child_code = declare_thread [("child_fd", Int)] $
+  \[child_fd] -> do
+  ev <- setup_connection child_fd
+  buf <- new_buf (bufsize+1) -- wee.
+  parse_result <- parse_request buf ev
+  -- XXX: error messages or something
+  ifE' (parse_result .> 1 .|| parse_result .< 0) exit
+
+  file_fd <- open buf kO_RDONLY
+  ifE' (file_fd .< 0) exit
+  let cleanup = close file_fd >> exit
+
+  let hdr_length = NumLit $ toInteger $ length response_header
+  header <- var "output_header" String (StringLit response_header)
+  amount_written <- full_write ev header hdr_length
+  ifE' (amount_written .< hdr_length) cleanup
+  
+  while 1 $ do
+    amount_read <- file_read file_fd buf bufsize
+    ifE' (amount_read .== 0) cleanup
+    amount_written <- full_write ev buf amount_read
+    ifE' (amount_written .< amount_read) cleanup
+
+
+main_loop = do
+  ev <- setup_listener port
+  while 1 $ do
+    fd' <- accept ev
+    spawn child_code [fd']

### 2011-12-05 src/Language/Foo/Codegen.hs

@@ -116,7 +116,7 @@ translateType t =
     Bool -> (cBool,[])
     FD -> (cInt,[])
     Event -> (cType "event_t",[cPtr])
-    String -> (cType "string",[]) --TODO
+    String -> (cChar,[cPtr])
     Buffer -> (cChar,[cPtr])
 
 --}}}

### 2011-12-05 src/rt/http_lib.c

@@ -1,0 +1,45 @@
+#include "lib.h"
+
+// buffer needs
+#define MAXLINE 4096
+
+#define ROOT_DIR "."
+
+int str_prefix(char *haystack, char *needle)
+{
+	while (*needle && *haystack == *needle)
+		haystack++, needle++;
+	return !*needle;
+}
+
+int http_parse(char *buf, int len)
+{
+	/* allocating this much space on the stack really bothers me... */
+	char method[MAXLINE], path[MAXLINE], version[MAXLINE];
+	
+	if (len >= MAXLINE) return -1;
+	buf[len] = '\0';
+	if (strlen(buf) != len) return -2;
+
+	char *line_end = strstr(buf, "\r\n");
+	// also allow '\n'
+	if (!line_end) line_end = strchr(buf, '\n');
+	if (!line_end) return 0;
+
+	sscanf(buf, "%s %s %s", method, path, version);
+
+	/* we only handle GET */
+	if (strcmp(method, "GET"))
+		return 501; // Not implemented
+
+	if (!str_prefix(version, "HTTP/1."))
+		return 400; // Bad prefix
+
+    // BUG: need to sanity check path
+	
+	if (snprintf(buf, MAXLINE, "%s%s", ROOT_DIR, path) >= MAXLINE)
+		return 500; // Internal server error; something better?
+
+	printf("opening file: %s\n", buf);
+	return 1;
+}

### 2011-12-05 src/rt/lib.h

@@ -23,3 +23,6 @@ int set_sock_reuse(int fd);
 int make_socket_non_blocking(int sfd);
 int sock_bind_v4(int fd, int addr, int port);
 int print_int(int n);
+
+// This shouldn't be here. Programs should be able to include other headers
+int http_parse(char *buf, int len);

### 2011-12-08 src/Memory.hs
Added (working) memory usage tester.

@@ -1,0 +1,15 @@
+module Memory where
+
+-- Tests our overhead by spawning as many threads as possible.
+
+import Language.Foo
+
+child_code = declare_thread [] $
+  \[] -> do x <- var "x" Int 1
+            while 1 $ do
+              y <- var "y" Int 1
+              return ()
+
+main_loop = do
+  while 1 $ do
+    spawn child_code []

### 2011-12-09 presentation/slides.tex
Added presentation.

\documentclass{beamer}
\usepackage{graphics}
\renewcommand{\t}{\texttt}

% Define Beamer appearance
\usetheme{Montpellier}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{headline}{} % comment out if you want navtree on slides
\setbeamertemplate{footline}[frame number]

\title{He-Man}
\subtitle{Haskell Event Manager, Apropos Networking}
\author{Carlo Angiuli \and Michael Sullivan}
\date{December 12, 2011}

\begin{document}

\begin{frame}
\maketitle
\end{frame}

\begin{frame}
\frametitle{The C10K problem}
How do you write network applications that scale to 10k clients?
\begin{itemize}
\item Concurrency is essential.
\item Must maintain per-client state.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Solutions}
\begin{itemize}
\item Threads: Write code in the obvious, straight line manner, using blocking
IO. Memory-intensive, scheduled by OS.
\item Event loops: Use non-blocking IO and multiplexing like \t{epoll()}.
Requires refactoring as a state machine.
\item Threads are easier to program, event loops perform better.
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Threads}
\small
\begin{verbatim}
main() {
  ...
  while (1) {
    fd = accept(...);
    spawn(thread, fd);
  }
}

thread(fd) {
  while (1) {
    len = read(fd, buf, SIZE);
    written = 0;
    while (written < len)
      written += write(fd, buf + written, len - written);
  }
}
\end{verbatim}
\end{frame}

\begin{frame}
\frametitle{Event loops}
\includegraphics[scale=.5]{graph.pdf}
\end{frame}

\begin{frame}
\frametitle{Best of both worlds}
Write straight-line code and generate state machine automatically.
\begin{itemize}
\item Embedded domain-specific language in Haskell.
\item Compiles to a C event loop, links to a small runtime \\
(and the rest of your code).
\item Generates basic blocks not containing blocking IO.
\item All blocking occurs in the runtime's main loop.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Status}
\begin{itemize}
\item We have a working echo server and rudimentary HTTP server.
\item Supports asynchronous disk IO and non-blocking network IO.
\item Doing performance testing on HTTP server.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Future work}
\begin{itemize}
\item Multithreaded event loop.
\item Typed source language (via phantom types or GADTs).
\item Optimized runtime.
\end{itemize}
\end{frame}

\end{document}

\end{document}

### 2011-12-09 src/Echo.hs
Renamed to HeMan.

@@ -1,6 +1,6 @@
 module Echo where
 
-import Language.Foo
+import Language.HeMan
 
 q_limit = 1024
 port = 2023

### 2011-12-09 src/Http.hs

@@ -1,6 +1,6 @@
-module Echo where
+module Http where
 
-import Language.Foo
+import Language.HeMan
 
 q_limit = 1024
 port = 8080

### 2011-12-09 src/Language/HeMan.hs

@@ -1,16 +1,16 @@
-module Language.Foo
-  (module Language.Foo.Syntax,
-   module Language.Foo.Lib,
+module Language.HeMan
+  (module Language.HeMan.Syntax,
+   module Language.HeMan.Lib,
    compile,
    testFront, testBack, testAll)
   where
 
-import Language.Foo.Syntax
-import Language.Foo.Lib
+import Language.HeMan.Syntax
+import Language.HeMan.Lib
 
-import Language.Foo.Backend
-import Language.Foo.Codegen
-import Language.Foo.Pretty
+import Language.HeMan.Backend
+import Language.HeMan.Codegen
+import Language.HeMan.Pretty
 
 compile file p = do writeFile file (codegen $ backend $ desugar p)
 

### 2011-12-09 src/Language/HeMan/Backend.hs

@@ -1,9 +1,9 @@
-module Language.Foo.Backend
+module Language.HeMan.Backend
   (Stmt(..), Tail(..), Block, Thread,
    backend)
   where
 
-import qualified Language.Foo.Syntax as Front
+import qualified Language.HeMan.Syntax as Front
   (Stmt(..), Expr(..), VDecl, Prim(..), Block, Var)
 import Control.Monad.RWS
 import Control.Monad.Writer

### 2011-12-09 src/Language/HeMan/Codegen.hs

@@ -1,10 +1,10 @@
-module Language.Foo.Codegen (codegen) where
+module Language.HeMan.Codegen (codegen) where
 
-import Language.Foo.Syntax
+import Language.HeMan.Syntax
   (Prim(..), ArithOp(..), ArithUnop(..), RelnOp(..), Type(..), Expr(..))
-import Language.Foo.Backend
+import Language.HeMan.Backend
   (Block, Thread, Stmt(..), Tail(..))
-import Language.Foo.CLib
+import Language.HeMan.CLib
 import Data.List
 import Data.Maybe
 

### 2011-12-09 src/Language/HeMan/Lib.hs

@@ -1,6 +1,6 @@
-module Language.Foo.Lib where
+module Language.HeMan.Lib where
 
-import Language.Foo.Syntax
+import Language.HeMan.Syntax
 
 -- Coreish runtime functions
 mk_nb_event fd modes =

### 2011-12-09 src/Language/HeMan/Pretty.hs

@@ -1,8 +1,8 @@
 {-# LANGUAGE FlexibleInstances, TypeSynonymInstances #-}
-module Language.Foo.Pretty (pretty) where
+module Language.HeMan.Pretty (pretty) where
 
-import qualified Language.Foo.Syntax as Lang
-import Language.Foo.Backend
+import qualified Language.HeMan.Syntax as Lang
+import Language.HeMan.Backend
 import Text.PrettyPrint
 import Data.List
 import qualified Language.C

### 2011-12-09 src/Language/HeMan/Syntax.hs

@@ -1,4 +1,4 @@
-module Language.Foo.Syntax where
+module Language.HeMan.Syntax where
 
 import Control.Monad.RWS
 

### 2011-12-09 src/Memory.hs

@@ -2,7 +2,7 @@ module Memory where
 
 -- Tests our overhead by spawning as many threads as possible.
 
-import Language.Foo
+import Language.HeMan
 
 child_code = declare_thread [] $
   \[] -> do x <- var "x" Int 1

### 2011-12-10 report/report.tex
Start report.
Started on report.

--- 2011-09-28 proposal/proposal.tex
@@
 \usepackage{cite}
+\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
+{hyperref}
 \renewcommand{\t}{\texttt}
 
-\title{15-744 Project Proposal}
+\title{He-Man: A domain-specific language for lightweight concurrency}
+%\subtitle{I said hey, what's goin' on}
 \author{Carlo Angiuli \and Michael Sullivan}
-\date{September 28, 2011}
+\date{December 12, 2011}
 
@@
 
-% Needs to include:
-%  - problem statement
-%  - state-of-the-art
-%  - work plan / milestones
-
-\section{Motivation}
+\section{Introduction}
 
 Modern network applications need to support many simultaneous clients---this is
-known as the C10K (``10,000 client'') problem\cite{Kegel}. But to maintain these
-simultaneous connections, an application must track each client's state
-independently. A straightforward way to do this is to spawn a thread for each
-client, but this approach does not scale as the number of clients grows.
+known as the C10K (``10,000 client'') problem\cite{Kegel}. Modern servers have
+enough memory and network resources to support this many
+
+To maintain these simultaneous connections, an application must maintain each
+client's state independently;
+
+while 
+block each one
+
+The simplest way to do this is to spawn a thread for each client, but this
+approach does not scale as the number of clients grows.
 
@@
 
-\section{Proposal}
+Because
 
@@
 
-The back-end of this system can be implemented via several different options
-which we are still evaluating. One option is to directly interface with the
-\t{epoll} hooks in the Haskell runtime system, bypassing the native lightweight
-threads. Another option is to generate C source code which itself implements the
-network application, optionally linkable with native C code.
+Generate C source code which itself implements the network application, linkable
+with native C code.
 
-\section{Milestones}
+simple runtime
 
-We will first collect a representative sample of Linux network applications and
-examine the range of high-level primitives required to support them. We will
-then decide on a specific back-end for our language, and design the front- and
-back-end in tandem. To evaluate the utility of this language, we will build a
-number of representative applications in our language (e.g., an echo server,
-simple HTTP server) and compare them to traditional implementations, with
-respect to both code simplicity and performance.
+\section{Related work}
+
+The Nginx\cite{Nginx} and Lighttpd\cite{Lighttpd} HTTP servers use an
+event-based architectures, unlike Apache's threads\cite{Apache}, with the
+specific goal of addressing the C10k problem,
+
+Haskell event loop stuff: \cite{LiZdancewic}
+
+Haskell has epoll support and lightweight threads
+
+Erlang
+
+\section{Front-end}
+
+\section{Back-end}
+
+\section{Performance}
+
+Each thread takes up XXX memory (see Memory.hs)
+
+HTTP server performance is XXX
+
+\section{Future work}
+
+\section{Conclusions}
 

### 2011-12-11 report/report.tex
Hacked out report introduction.

@@
-\documentclass{article}
+\documentclass[twocolumn]{article}
 \usepackage{cite}
@@
 \renewcommand{\t}{\texttt}
+\renewcommand{\b}{\textbf}
+\renewcommand{\i}{\textit}
+
+\newcommand{\prettybox}[1]{\fbox{\parbox{\linewidth}{#1}}}
+
+% A bunch of crazy BNF TeX I wrote once
+\def\bnf#1{\prettybox{\halign{##\hfil& \hfil ##\hfil& \ ##\hfil\cr #1}}}
+\def\p#1{& = & #1 \cr}
+\def\pp#1{& & #1 \cr}
+\def\a#1{& $\mid$ & #1 \cr}
 
@@
 
-Modern network applications need to support many simultaneous clients---this is
-known as the C10K (``10,000 client'') problem\cite{Kegel}. Modern servers have
-enough memory and network resources to support this many
+Network applications need to support many simultaneous clients---this is known
+as the C10K (``10,000 client'') problem\cite{Kegel}. Modern servers have enough
+memory and bandwidth for massive concurrency, but it is difficult to design
+applications to take full advantage of these resources.
 
-To maintain these simultaneous connections, an application must maintain each
-client's state independently;
+A server must maintain separate state for each connection, and often waits for
+network or disk I/O to complete before continuing to serve a particular client.
+The simplest solution is to maintain one thread per client, using blocking I/O
+within each thread. This incurs frequent and expensive context switches at the
+operating system level; because the threads have identical code and little
+private data, this approach is unnecessary and does not scale.
 
-while 
-block each one
+An improvement is to write multithreaded code, in which each thread serves
+multiple clients. This is the most popular approach, and is utilized in the
+Apache HTTP server \cite{Apache}.
 
-The simplest way to do this is to spawn a thread for each client, but this
-approach does not scale as the number of clients grows.
+Languages like Erlang and Haskell provide lightweight threads via user-space
+schedulers. This affords even better performance, but remains unnecessarily
+powerful for this problem space, imposing runtime overhead on applications which
+don't require preemptable, independently scheduled threads \cite{Vinoski}.
 
-A lightweight approach is to use kernel mechanisms for multiplexed, non-blocking
-I/O, such as \t{select}, \t{poll}, \t{epoll}, or \t{kquery}; unfortunately, to
-maintain separate state for each client, application programmers must manually
-structure their code as state machines, which is tedious and error-prone.
-Unsurprisingly, there exist many thin C/C++ wrappers for these facilities, but
-none avoid major code refactoring.
+By far the leanest approach is to write an event loop around kernel facilities
+for multiplexed, non-blocking I/O, such as \t{select}, \t{poll}, \t{epoll}, or
+\t{kquery}. In this model, all I/O occurs at a single call site in the code.
+The programmer is responsible for explicitly maintaining per-client state and
+invoking the appropriate client on each I/O event.
 
-To maintain the illusion of threads (in particular, client-specific state),
-languages like Erlang and Haskell provide lightweight threading via user-space
-schedulers and non-blocking IO. While this affords better performance, the full
-generality of this approach still imposes unnecessary overhead on networking
-applications which don't require preemptable, independently scheduled threads
-\cite{Vinoski}.
+This requires the entire application to be restructured as a state machine with
+explicit per-client control blocks, a model which is very difficult for
+programmers to reason about. There exist numerous C/C++ libraries for event
+loops, but none of these avoid the state machine paradigm.
 
-Because
+Our observation is that the event loop model, in programming languages parlance,
+is essentially trampolining continuation-passing-style, which can be obtained by
+mechanical translation from the straight-line threaded code. This suggests it is
+possible to obtain the best of both worlds---a threaded programming model with
+the performance of event loops.
 
-Hence, we intend to produce an embedded domain-specific language in Haskell to
-describe concurrent network applications. This language will provide composable,
-high-level network operations, in which programmers may express their
-application logic naturally, as if using threads.
+To that end, we present He-Man: Haskell Event Manager, Apropos Networking, an
+embedded domain-specific language (EDSL) hosted in Haskell, which provides a
+threaded abstraction for network application programming and compiles to a C
+event loop based around Linux's \t{epoll}.
 
-However, this high-level description will be automatically synthesized into a
-state machine representation usable in a C loop invoking \t{epoll}, using no
-threads whatsoever. We will additionally output a graphical representation of
-this underlying state machine.
-
-Generate C source code which itself implements the network application, linkable
-with native C code.
-
-simple runtime
+He-Man is a straightforward monadic language with built-in network primitives,
+offering a simple thread-per-client programming model. It is easily extensible
+in Haskell, and also allows calls to foreign C functions; it outputs C code
+which links with an event loop runtime and any code a programmer might prefer to
+write directly in C. Our vision is that C programmers can use He-Man for their
+core application logic, and can write the rest of their application in C.
 
@@
 
-The Nginx\cite{Nginx} and Lighttpd\cite{Lighttpd} HTTP servers use an
-event-based architectures, unlike Apache's threads\cite{Apache}, with the
-specific goal of addressing the C10k problem,
+The Lighttpd\cite{Lighttpd} and Nginx\cite{Nginx} HTTP servers use event-based
+architectures, with the stated goal of addressing the C10k problem. Indeed, they
+are fast (XXX CITATION)
 
@@
 
+As mentioned in the intro,
 Haskell has epoll support and lightweight threads
-
 Erlang
@@
 \section{Back-end}
+
+The back-end converts our linear front-end code into a collection of states
+
+is to compile C,
+
+It's important that the only blocking operations happen in the main loop
+
+We separate
+
+Threads are separated and replaced by labels
+
+pass to make it
+
+\begin{figure}[ht]
+\centering
+\bnf{
+\i{Thread}
+\p{(\i{thread\_label}, [(\i{Var}, \i{Type})])}
+\i{Block}
+\p{(\i{label}, \i{thread\_label}, [\i{Stmt}], \i{Tail})}
+\i{Tail}
+\p{\b{Goto} \i{label}}
+\a{\b{GotoWait} \i{label}}
+\a{\b{If} \i{Expr} \i{Tail} \i{Tail}}
+\a{\b{Exit}}
+\i{Stmt}
+\p{\b{Decl} (\i{Var}, \i{Type}) \i{Expr}}
+\a{\b{Assign} \i{Expr} \i{Expr}}
+\a{\b{Exp} \i{Expr}}
+\a{\b{Spawn} \i{label} [((\i{Var}, \i{Type}), \i{Expr})]}
+}
+\caption{Back-end grammar}
+\end{figure}
 
@@
 
+write a better allocator
+
+We will additionally output a graphical representation of this underlying state
+machine.
+
 \section{Conclusions}

### 2011-12-11 src/rt/http_lib.c
Less spew

@@ -40,6 +40,6 @@ int http_parse(char *buf, int len)
 	if (snprintf(buf, MAXLINE, "%s%s", ROOT_DIR, path) >= MAXLINE)
 		return 500; // Internal server error; something better?
 
-	printf("opening file: %s\n", buf);
+	//printf("opening file: %s\n", buf);
 	return 1;
 }

### 2011-12-11 src/rt/mainloop.c

@@ -189,7 +189,7 @@ static void do_poll(bool can_sleep)
 
 static void run_thread(thread_t *thread)
 {
-	printf("running %d\n", thread->tid);
+	//printf("running %d\n", thread->tid);
 	bool runnable = true;
 	// Run the thread until it has run for a while or has stopped being
 	// runnable.

### 2011-12-12 src/Http.hs
argh

@@ -4,7 +4,7 @@ import Language.HeMan
 
 q_limit = 1024
 port = 8080
-bufsize = 4096-1
+bufsize = 4096*4
 
 http_parse buf len =
   callName "parse_result" (CFn "http_parse") Int [buf, len]
@@ -40,7 +40,7 @@ parse_request buf ev = do
 child_code = declare_thread [("child_fd", Int)] $
   \[child_fd] -> do
   ev <- setup_connection child_fd
-  buf <- new_buf (bufsize+1) -- wee.
+  buf <- new_buf (bufsize+1) -- pfbbt.
   parse_result <- parse_request buf ev
   -- XXX: error messages or something
   ifE' (parse_result .> 1 .|| parse_result .< 0) exit

### 2011-12-12 src/rt/mainloop.c

@@ -10,6 +10,7 @@
 #include <time.h>
 #include <stdarg.h>
 #include <stdbool.h>
+#include <signal.h>
 
 #include <libaio.h>
 #include <sys/eventfd.h>
@@ -215,6 +216,9 @@ void main_loop(void)
 void setup_main_loop(void)
 {
 	int ret;
+
+	signal(SIGPIPE, SIG_IGN);
+	
 	static event_t aio_dummy_event;
 	state.aio_dummy_event = &aio_dummy_event;
 

### 2011-12-12 src/rt/variable_queue.h
Remove the counts from vq.

@@ -10,6 +10,14 @@
 #include <stdlib.h>
 #include <assert.h>
 
+#define VQ_DISABLE_COUNT
+
+#ifdef VQ_DISABLE_COUNT
+#define CNT(x)
+#else
+#define CNT(x) x
+#endif
+
 /**
  * @def Q_NEW_HEAD(Q_HEAD_TYPE, Q_ELEM_TYPE) 
  *
@@ -30,7 +38,7 @@
 	typedef struct Q_HEAD_TYPE {             \
 		struct Q_ELEM_TYPE *head;            \
 		struct Q_ELEM_TYPE *tail;            \
-		size_t count;                        \
+		CNT(size_t count;)					 \
 	} Q_HEAD_TYPE
 
 /**
@@ -72,7 +80,7 @@
  **/
 #define Q_INIT_HEAD(Q_HEAD) do {                \
 		(Q_HEAD)->head = (Q_HEAD)->tail = NULL; \
-		(Q_HEAD)->count = 0;                    \
+		CNT((Q_HEAD)->count = 0;)				\
 	} while (0)
 
 /**
@@ -136,7 +144,7 @@
 		(Q_HEAD)->head = (Q_ELEM);                      \
 		if (!(Q_HEAD)->tail)                            \
 			(Q_HEAD)->tail = (Q_ELEM);                  \
-		(Q_HEAD)->count++;                              \
+		CNT((Q_HEAD)->count++;)							\
 		                                                \
 		Q_ASSERT_CONSISTENT(Q_HEAD, Q_ELEM, LINK_NAME); \
 	} while (0)
@@ -165,7 +173,7 @@
 		(Q_HEAD)->tail = (Q_ELEM);					  	\
 		if (!(Q_HEAD)->head)							\
 			(Q_HEAD)->head = (Q_ELEM);					\
-		(Q_HEAD)->count++;                              \
+		CNT((Q_HEAD)->count++;)							\
 		                                                \
 		Q_ASSERT_CONSISTENT(Q_HEAD, Q_ELEM, LINK_NAME); \
 	} while (0)
@@ -262,7 +270,7 @@
 		(Q_INQ)->LINK_NAME.next = (Q_TOINSERT);                   \
 		if ((Q_INQ) == (Q_HEAD)->tail)                            \
 			(Q_HEAD)->tail = (Q_TOINSERT);                        \
-		(Q_HEAD)->count++;                                        \
+		CNT((Q_HEAD)->count++;)									  \
[…]

### 2011-12-12 report/report.tex
papering

@@
 \maketitle
+
+%XXX: abstract
 
@@
 
-\section{Performance}
+\section{Evaluation}
 
-Each thread takes up XXX memory (see Memory.hs)
+The per-thread memory overhead imposed by He-Man is very small: it
+consists of a thread id, list pointers for the scheduling queue, the
+next state to enter, a pointer to the most recently received event,
+and list heads and tails for lists of the events being waited on,
+events owned by the thread, and buffers owned by the thread (the
+latter two are so that we can automatically reclaim these resources
+when the thread exits). On a 64-bit system, the total overhead comes
+out to 88 bytes, and could easily be reduced further (e.g. by using
+singly linked lists).  For comparison, the default thread stack size
+under NPTL on Linux is 2 megabytes.
+
+say something about Memory.hs
+each thread takes up XXX memory (see Memory.hs)
+
+To evaluate He-Man,
 

### 2011-12-12 src/rt/echo_handwritten.c
delete some boilerplate

@@ -103,7 +103,6 @@ bool accept_state(struct thread_t *thread)
 	new_thread->thread.cont = setup_thread;
 	make_runnable(&new_thread->thread);
 
-	printf("got one!\n");
 	return true;
 }
 

### 2011-12-12 src/rt/echo_handwritten.c
delete some more boilerplate

@@ -28,7 +28,6 @@ bool read_state(struct thread_t *thread);
 bool write_state(struct thread_t *thread)
 {
 	echo_thread_t *e = (echo_thread_t *)thread;
-	thread->finished_event = NULL;
 	thread->cont = read_state;
 
 	int left = e->amt_read - e->amt_written;
@@ -54,7 +53,6 @@ bool write_state(struct thread_t *thread)
 bool read_state(struct thread_t *thread)
 {
 	echo_thread_t *e = (echo_thread_t *)thread;
-	thread->finished_event = NULL;
 	thread->cont = read_state;
 	
 	int count = read(e->fd, e->buf, sizeof(e->buf));
@@ -88,7 +86,6 @@ bool setup_thread(struct thread_t *thread)
 bool accept_state(struct thread_t *thread)
 {
 	echo_thread_t *e = (echo_thread_t *)thread;
-	thread->finished_event = NULL;
 	thread->cont = accept_state;
 	
 	int fd = accept(e->fd, NULL, NULL);

### 2011-12-12 report/report.tex
Migrate to SIGPLAN template.
write a bunch about evaluation

@@
-\documentclass[twocolumn]{article}
-\usepackage{cite}
+\documentclass[preprint,11pt]{sigplanconf}
 \usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
@@
 %\subtitle{I said hey, what's goin' on}
-\author{Carlo Angiuli \and Michael Sullivan}
+\authorinfo{Carlo Angiuli}{}{cangiuli@cs.cmu.edu}
+\authorinfo{Michael Sullivan}{}{mjsulliv@andrew.cmu.edu}
 \date{December 12, 2011}
@@
 
-\section{Front-end}
+\section{Implementation}
 
-\section{Back-end}
+\subsection{Front-end}
+
+\subsection{Back-end}
 
@@
 
+\subsection{Runtime}
+
 \section{Evaluation}
+
+To evaluate He-Man, we built two network applications with He-Man and
+compared them to similar network applications built in C.
+We also discuss He-Man's memory overhead.
+
+\subsection{Echo Server}
+
+To compare the ease of writing event based network applications, we
+built simple echo servers using both He-Man and C. The C version was
+written in as an event loop driven explicit state machine using the
+He-Man runtime. The C version, which is fairly tightly written, is 98
+lines of source, according to SLOCCount. It is fairly readable, but
+its structure is obscured by being split into states and the need to
+explicitly handle read and write returning EAGAIN and write returning
+short. The state machine has a total of five states. The He-Man
+version is 33 lines long and is a very clear representation of the
+control flow of each individual thread. (Additionally, the He-Man
+library has 17 lines of code implementing generic routines for
+simulating blocking IO using nonblocking IO and wait.) The total line
+counts, however, somewhat understate the difference between the
+versions, since the boilerplate for setting up connections is about
+the same for both versions. When just looking at the core read/write
+loop, the difference is much more apparent: the C version is 38 lines
+of code, and the He-Man version is 5.
+
+\subsection{HTTP Server}
+
+HTTP server performance is XXX
+
+\subsection{Memory Overhead}
 
@@
 
-To evaluate He-Man,
-
-HTTP server performance is XXX
-
 \section{Future work}

### 2011-12-12 report/report.tex
Added front-end grammar, other work

--- 2011-12-12 report/report.tex
@@
 
-%XXX: abstract
+\begin{abstract}
+TODO
+\end{abstract}
 
@@
 Network applications need to support many simultaneous clients---this is known
-as the C10K (``10,000 client'') problem\cite{Kegel}. Modern servers have enough
+as the C10K (``10,000 client'') problem \cite{Kegel}. Modern servers have enough
 memory and bandwidth for massive concurrency, but it is difficult to design
@@
 
-The Lighttpd\cite{Lighttpd} and Nginx\cite{Nginx} HTTP servers use event-based
+The Lighttpd \cite{Lighttpd} and Nginx \cite{Nginx} HTTP servers use event-based
 architectures, with the stated goal of addressing the C10k problem. Indeed, they
@@
 
-\section{Front-end}
+\section{The event loop model}
 
-\section{Back-end}
+Event loops are a good way to get the level of concurrency that you need
+
+we don't need arbitrary interleaving of instructions,
+
+there shouldn't be much
+
+and I/O time (both network and disk) probably dominates computation time
+
+thus it's important to run short computations until a thread blocks,
+
+The back-end converts our linear front-end code into a collection of states
+
+is to compile C,
+
+It's important that the only blocking operations happen in the main loop
+
+We separate
+
+Threads are separated and replaced by labels
+
+pass to make it
+
+\section{Implementation}
+
+\subsection{Front-end}
+
+\begin{figure}[ht]
+\centering
+\bnf{
+\i{Stmt}
+\p{\b{Decl} (\i{Var}, \i{Type}) \i{Expr}}
+\a{\b{While} \i{Expr} [\i{Stmt}]}
+\a{\b{If} \i{Expr} [\i{Stmt}] [\i{Stmt}]}
+\a{\b{Spawn} ([(\i{Var}, \i{Type})], [\i{Stmt}]) [\i{Expr}]}
+\a{\b{Assign} \i{Expr} \i{Expr}}
+\a{\b{Exp} \i{Expr}}
+\a{\b{Wait} \i{Expr}}
+\a{\b{Exit}}
+\i{Expr}
+\p{\b{Call} \i{String} [\i{Expr}]}
+\a{\b{Arith} \i{ArithOp} \i{Expr} \i{Expr}}
+\a{\b{ArithUnop} \i{ArithUnop} \i{Expr}}
+\a{\b{RelnOp} \i{RelnOp} \i{Expr} \i{Expr}}
+\a{\b{Constant} \i{String}}
+\a{\b{NumLit} \i{Integer}}
+\a{\b{StringLit} \i{String}}
+\a{\b{Var} \i{String}}
+\a{\b{CurThread}}
+\i{Type} 
+\p{\b{Int} $\mid$ \b{Bool} $\mid$ \b{String}
+   $\mid$ \b{FD} $\mid$ \b{Buffer} $\mid$ \b{Event}}
+}
+\caption{Front-end grammar.}
+\end{figure}
+
+\subsection{Back-end}
 
@@
 }
-\caption{Back-end grammar}
+\caption{Back-end grammar.}
 \end{figure}
+
+\subsection{Runtime}
 
@@
 
+less space per thread
+
+decide when to allow internal if statements
+
 We will additionally output a graphical representation of this underlying state
 machine.
+
+observable sharing (?)
 
@@
 \bibliography{citations}{}
-\bibliographystyle{plain}
+\bibliographystyle{abbrvnat}
 

### 2011-12-12 report/report.tex
minor fix to intro

@@
 
+To evaluate He-Man, we built two network applications with He-Man and
+compared them to similar network applications built in C.
+We also discuss He-Man's memory overhead.
+
+\subsection{Echo Server}
+
+To compare the ease of writing event based network applications, we
+built simple echo servers using both He-Man and C. The C version was
+written in as an event loop driven explicit state machine using the
+He-Man runtime. The C version, which is fairly tightly written, is 98
+lines of source, according to SLOCCount. It is fairly readable, but
+its structure is obscured by being split into states and the need to
+explicitly handle read and write returning EAGAIN and write returning
+short. The state machine has a total of five states. The He-Man
+version is 33 lines long and is a very clear representation of the
+control flow of each individual thread. (Additionally, the He-Man
+library has 17 lines of code implementing generic routines for
+simulating blocking IO using nonblocking IO and wait.) The total line
+counts, however, somewhat understate the difference between the
+versions, since the boilerplate for setting up connections is about
+the same for both versions. When just looking at the core read/write
+loop, the difference is much more apparent: the C version is 38 lines
+of code, and the He-Man version is 5.
+
+\subsection{HTTP Server}
+
+HTTP server performance is XXX
+
+\subsection{Memory Overhead}
+
 The per-thread memory overhead imposed by He-Man is very small: it
@@
 each thread takes up XXX memory (see Memory.hs)
-
-To evaluate He-Man,
-
-HTTP server performance is XXX
 

### 2011-12-12 report/report.tex
tweaks

@@
 for multiplexed, non-blocking I/O, such as \t{select}, \t{poll}, \t{epoll}, or
-\t{kquery}. In this model, all I/O occurs at a single call site in the code.
+\t{kquery}. In this model, all blocking occurs at a single call site in the code.
 The programmer is responsible for explicitly maintaining per-client state and
@@
 \p{\b{Int} $\mid$ \b{Bool} $\mid$ \b{String}
-   $\mid$ \b{FD} $\mid$ \b{Buffer} $\mid$ \b{Event}}
+   $\mid$ \b{Buffer} $\mid$ \b{Event}}
 }
@@
 
-say something about Memory.hs
-each thread takes up XXX memory (see Memory.hs)
+say something about Memory.hs??
 

### 2011-12-12 src/Language/HeMan/Lib.hs
some more bullshit

@@ -39,7 +39,7 @@ file_write fd buf len =
 print_int n  =
   call (CFn "print_int") Int [n]
 
-errno = Var "errno" -- weeeee
+errno = Constant "errno" -- weeeee
 
 -- TODO: a bunch more
 
@@ -76,4 +76,14 @@ full_write ev buf size = do
     amt <- do_write ev (buf + amt_written) (size - amt_written)
     amt_written .= amt_written + amt
     ifE' (amt .== 0) $ do failed .= 1
+  return amt_written
+
+full_write_good_but_broken (fd, e) buf size = do
+  amt_written <- var "total_written" Int 0
+  failed <- var "write_failed" Int 0
+  while (amt_written .< size .&& failed .== 0) $ do
+    amt <- sock_write fd (buf + amt_written) (size - amt_written)
+    ifE (amt .== -1) (wait e) $
+      (ifE (amt .== 0) (failed .= 1)
+       (amt_written .= amt_written + amt))
   return amt_written

### 2011-12-12 report/report.tex
add another note to future work
Evaluation basically done.

@@
 {hyperref}
+\usepackage{graphicx}
 \renewcommand{\t}{\texttt}
@@
 
+The generated code is substantially less tight than the C version,
+although it is unlikely to be any slower in practice. It has 15 states
+and is 319 lines of C. This could be reduced substantially by allowing
+conditionals outside of block tail position.
+
 \subsection{HTTP Server}
 
-HTTP server performance is XXX
+To test the suitability of our system for more complicated programs
+and to demonstrate seperating network logic and application logic, we
+built a simple (and bad) HTTP 1.0 server. The network logic is written
+in He-Man and request parsing is written in C. The C side is
+implemented as a function that takes a buffer and returns whether it
+forms a complete HTTP request and, if so, fills a buffer with the
+filename to serve. The entire server is 54 lines of He-Man haskell and
+27 lines of C. It suffers from a number of major deficiencies, but is
+complete enough to benchmark.
+
+Our benchmark measurements are taken using ab, the Apache HTTP server
+benchmarking tool. \cite{ApacheAB} Measurements were collected by
+having a variable number of client threads request a file 100 times;
+the experiment was repeated with a 16 kilobyte file and a 128 kilobyte
+file. We ran the tests against both our He-Man http server and Apache
+2.2.20, installed with the default Ubuntu 11.10 settings, except for
+client, thread, and process limits being adjusted upwards. Since the
+He-Man runtime is currently only single threaded, we ran the Apache
+web server pinned to a single processor. Since the test repeatedly
+requests the same file, it will always be in the kernel's disk cache,
+and thus this test is not disk bound.
+
+For our experiments, the server ran Linux kernel 3.0.0 on a Lenovo
+Thinkpad W520 with a quad core 2.3 GHz Intel Core i7-2820QM and 8
+gigabytes of RAM, and the client ran on a Lenovo Thinkpad T500 with a
+Intel Core 2 Duo and 8 gigabtyes of RAM.  % XXX: client specs
+The machines were connected by a gigabit Ethernet link.
+
+Unfortunately, ab is not particularly scalable, so we had trouble
+testing with more clients than 500.
+
+Figure \ref{fig:graph128} compares our server to Apache for a 128 KB
+file. Our server performs nearly identically to Apache on this load.
+Figure \ref{fig:graph16} compares our server to Apache for a 16 KB
+file. In this scenario, we substantially outperform Apache. We
+conjecture this is because with a smaller file size, overhead for
+connection setup begins to dominate, and connection setup in He-Man
+consists of a few memory allocations whereas connection setup under
+Apache may require spawning new OS level threads.
+
+\begin{figure*}[!htb]
+	\centering
+	\includegraphics[width=0.9\textwidth]{128kb_graph.png}
+	\caption{Web server benchmark, 128kb file}
+	\label{fig:graph128}
+\end{figure*}
+\begin{figure*}[!htb]
+	\centering
+	\includegraphics[width=0.9\textwidth]{16kb_graph.png}
+	\caption{Web server benchmark, 16kb file}
+	\label{fig:graph16}
+\end{figure*}
 
@@
 
+multithreaded runtime
+
 write a better allocator

### 2011-12-12 report/report.tex
Related work.

--- 2011-12-12 report/report.tex
@@
 
-The Lighttpd \cite{Lighttpd} and Nginx \cite{Nginx} HTTP servers use event-based
-architectures, with the stated goal of addressing the C10k problem. Indeed, they
-are fast (XXX CITATION)
+The C10K problem is particularly evident in the context of HTTP servers, which
+routinely maintain many concurrent connections. The popular Lighttpd and Nginx
+HTTP server projects were both started with the explicit goal of addressing the
+C10K problem \cite{Lighttpd,Nginx}, and both use event-based architectures
+similar to He-Man.
+
+A number of functional languages support lightweight threads with user-space
+scheduling, most notably Erlang, which was designed for highly concurrent
+telecommunications applications, and easily scales to 20,000 clients
+\cite{Hellstrom}. Haskell
 
 Haskell event loop stuff: \cite{LiZdancewic}
+
+FoxNet: \cite{BiagioniEtAl}
 
@@
 
-and I/O time (both network and disk) probably dominates computation time
+I/O time (both network and disk) dominates the computation time
 

### 2011-12-13 report/report.tex
Finished Related Work.

@@
 The C10K problem is particularly evident in the context of HTTP servers, which
-routinely maintain many concurrent connections. The popular Lighttpd and Nginx
-HTTP server projects were both started with the explicit goal of addressing the
-C10K problem \cite{Lighttpd,Nginx}, and both use event-based architectures
-similar to He-Man.
+routinely maintain many concurrent connections. As we discuss in Section
+\ref{sec:evaluation}, Apache does not scale particularly well. The popular
+Lighttpd and Nginx HTTP server projects were both started with the explicit goal
+of outperforming Apache with respect to the C10K problem \cite{Lighttpd,Nginx},
+and both use event-based architectures similar to He-Man.
 
@@
 telecommunications applications, and easily scales to 20,000 clients
-\cite{Hellstrom}. Haskell
+\cite{Hellstrom}. The Haskell runtime supports a variety of concurrency
+primitives which permit a lightweight threaded programming model \cite{LiEtAl}.
 
-Haskell event loop stuff: \cite{LiZdancewic}
+%A number of custom network stack implementations exist for functional
+%languages, such as FoxNet for Standard ML \cite{BiagioniEtAl}, and
 
-FoxNet: \cite{BiagioniEtAl}
-
-As mentioned in the intro,
-Haskell has epoll support and lightweight threads
-Erlang
+He-Man is most similar to Li and Zdancewic's Haskell implementation of a hybrid
+event/thread programming model for network applications \cite{LiZdancewic}.
+Their project has similar goals to ours, but is implemented entirely within
+Haskell. As a result, unlike He-Man, the entire application must be written in
+Haskell, and incurs runtime overhead from the Haskell system.
 
@@
 
-\section{Evaluation}
+\section{Evaluation}\label{sec:evaluation}
 

### 2011-12-13 report/report.tex
Write a bunch about the runtime.

--- 2011-12-12 report/report.tex
@@
 
-The C10K problem is particularly evident in the context of HTTP servers, which
-routinely maintain many concurrent connections. The popular Lighttpd and Nginx
-HTTP server projects were both started with the explicit goal of addressing the
-C10K problem \cite{Lighttpd,Nginx}, and both use event-based architectures
-similar to He-Man.
-
-A number of functional languages support lightweight threads with user-space
-scheduling, most notably Erlang, which was designed for highly concurrent
-telecommunications applications, and easily scales to 20,000 clients
-\cite{Hellstrom}. Haskell
+The Lighttpd \cite{Lighttpd} and Nginx \cite{Nginx} HTTP servers use event-based
+architectures, with the stated goal of addressing the C10k problem. Indeed, they
+are fast (XXX CITATION)
 
 Haskell event loop stuff: \cite{LiZdancewic}
-
-FoxNet: \cite{BiagioniEtAl}
 
@@
 
-I/O time (both network and disk) dominates the computation time
+and I/O time (both network and disk) probably dominates computation time
 
@@
 \subsection{Runtime}
+
+The He-Man runtime is written in C and provides a state machine based
+thread scheduling model. In the view of the runtime, a thread consists
+of some bookkeeping state, some private state, and the next
+continuation function the thread is to run. The main part of the
+runtime's state consists of the runqueue, which contains a list of all
+currently runnable threads.
+
+Continuation functions take a pointer to the thread information,
+update the continuation field in the thread data, and return whether
+the thread is still runnable; to block, a continuation function
+registers itself as blocking on an event and returns false; to exit,
+it frees the thread structure and returns false.
+Running a thread, then, consists of repeatedly invoking the thread's
+continuation function until it returns false. To prevent a
+thread that does not need to block from starving the rest of the
+system, after some number of invocations of the continuation
+(currently 50), the thread is paused and put on the back of the
+runqueue.
+
+The He-Man runtime main loop is very simple, then: it loops
+indefinitely, and on each iteration it removes the thread at the head
+of the runqueue (if it exists) and runs it, and polls for IO events.
+
+The runtime provides a blocking mechanism based on an event
+abstraction. To the runtime, an event consists of a pointer to a
+thread that is blocked on it and some arbitrary identifier (in
+practice the file descriptor). To create an event for use with
+nonblocking IO, an event is allocated, marked with the associated file
+descriptor, and a new edge-triggered epoll event is created for the
+file descriptor using epoll\_ctl, with the event structure as the
+associated pointer data.
+We use edge-triggered instead of level-triggered so that we can wait
+on both read and write events for a given file descriptor without
+having to adjust which we are waiting for at runtime, and because
+He-Man's programming model makes it easy to meet the requirement that
+blocking on a file descriptor is only done after an operation on the
+descriptor has returned EAGAIN.
+The actual polling for IO events is done by calling epoll. For each
+event that has been returned by epoll, if a thread is waiting on the
+event, the thread is added to the runqueue. If the the runqueue is
+empty, the call to epoll is blocking, otherwise it is called with a
+timeout of zero seconds, to simply poll for events.
+
+The runtime also has support for Linux Kernel Asynchronous IO for file
+IO, using a Linux eventfd to allow epoll to wait on AIO events. This
+support is currently unused, since it was not actually necessary to
+acheive acceptable performance for the web server.
 

### 2011-12-13 report/report.tex
cleanup

--- 2011-12-13 report/report.tex
@@
 
+The He-Man runtime is written in C and provides a state machine based
+thread scheduling model. In the view of the runtime, a thread consists
+of some bookkeeping state, some private state, and the next
+continuation function the thread is to run. The main part of the
+runtime's state consists of the runqueue, which contains a list of all
+currently runnable threads.
+
+Continuation functions take a pointer to the thread information,
+update the continuation field in the thread data, and return whether
+the thread is still runnable; to block, a continuation function
+registers itself as blocking on an event and returns false; to exit,
+it frees the thread structure and returns false.
+Running a thread, then, consists of repeatedly invoking the thread's
+continuation function until it returns false. To prevent a
+thread that does not need to block from starving the rest of the
+system, after some number of invocations of the continuation
+(currently 50), the thread is paused and put on the back of the
+runqueue.
+
+The He-Man runtime main loop is very simple, then: it loops
+indefinitely, and on each iteration it removes the thread at the head
+of the runqueue (if it exists) and runs it, and polls for IO events.
+
+The runtime provides a blocking mechanism based on an event
+abstraction. To the runtime, an event consists of a pointer to a
+thread that is blocked on it and some arbitrary identifier (in
+practice the file descriptor). To create an event for use with
+nonblocking IO, an event is allocated, marked with the associated file
+descriptor, and a new edge-triggered \t{epoll} event is created for the
+file descriptor using \t{epoll\_ctl}, with the event structure as the
+associated pointer data.
+We use edge-triggered instead of level-triggered so that we can wait
+on both read and write events for a given file descriptor without
+having to adjust which we are waiting for at runtime, and because
+He-Man's programming model makes it easy to meet the requirement that
+blocking on a file descriptor is only done after an operation on the
+descriptor has returned \t{EAGAIN}.
+The actual polling for IO events is done by calling \t{epoll}. For each
+event that has been returned by epoll, if a thread is waiting on the
+event, the thread is added to the runqueue. If the the runqueue is
+empty, the call to epoll is blocking, otherwise it is called with a
+timeout of zero seconds, to simply poll for events.
+
+The runtime also has support for Linux Kernel Asynchronous IO for file
+IO, using a Linux \t{eventfd} to allow \t{epoll} to wait on AIO
+events. This support is currently unused, since it was not actually
+necessary to acheive acceptable performance for the web server.
+
 \section{Evaluation}\label{sec:evaluation}

### 2011-12-13 report/report.tex
More

--- 2011-12-13 report/report.tex
@@
 for multiplexed, non-blocking I/O, such as \t{select}, \t{poll}, \t{epoll}, or
-\t{kquery}. In this model, all blocking occurs at a single call site in the code.
-The programmer is responsible for explicitly maintaining per-client state and
-invoking the appropriate client on each I/O event.
+\t{kquery}. In this model, all blocking occurs at a single call site in the
+code. The programmer is responsible for explicitly maintaining per-client state
+and invoking the appropriate client on each I/O event.
 
@@
 Their project has similar goals to ours, but is implemented entirely within
-Haskell. As a result, unlike He-Man, the entire application must be written in
-Haskell, and incurs runtime overhead from the Haskell system.
+Haskell. As a result, unlike He-Man, their framework requires entire
+applications to be written in Haskell, and incurs runtime overhead from the
+Haskell system.
 
@@
 
-Event loops are a good way to get the level of concurrency that you need
+Networking applications require a very constrained form of concurrency. Because
+I/O (both network and disk) latency dominates the computation time, each
+client's thread of execution spends most of its time waiting on I/O. In
+particular, we would like to always yield to a client no longer blocked on I/O;
+we need not yield to blocked clients, nor do we need arbitrary interleaving of
+execution.
 
-we don't need arbitrary interleaving of instructions,
+As of version 2.5.44, Linux has supported \t{epoll} for non-blocking
+multiplexing I/O. A server can register any number of file descriptors with
+\t{epoll\_ctl}, and a call to \t{epoll\_wait} returns the ready descriptors. In
+the event loop model, all blocking I/O occurs in a single loop which
+
+It's important that the only blocking operations happen in the main loop
+
+Event loops are a good way to obtain this 
+to get the level of concurrency that you need
 
 there shouldn't be much
-
-I/O time (both network and disk) dominates the computation time
 
@@
 is to compile C,
-
-It's important that the only blocking operations happen in the main loop
 

### 2011-12-13 report/report.tex
Finished event loop model section.

@@
 multiplexing I/O. A server can register any number of file descriptors with
-\t{epoll\_ctl}, and a call to \t{epoll\_wait} returns the ready descriptors. In
-the event loop model, all blocking I/O occurs in a single loop which
+\t{epoll\_ctl}, and a call to \t{epoll\_wait} returns the ready descriptors.
 
-It's important that the only blocking operations happen in the main loop
+In the event loop model, all blocking I/O must occur in a single loop around
+\t{epoll\_wait}. Each thread is stored as a \t{struct} containing thread-local
+state, including the next instruction pointer, and each registered descriptor
+has a pointer to the thread waiting on it. Instead of performing an I/O call, a
+thread simply registers a new descriptor and returns back to the main loop;
+whenever a descriptor is ready, the main loop jumps to the instruction pointer
+of the thread waiting on it. Spawning a new thread requires only allocating a
+new thread \t{struct}, and registering it in the main loop.
 
-Event loops are a good way to obtain this 
-to get the level of concurrency that you need
-
-there shouldn't be much
-
-thus it's important to run short computations until a thread blocks,
-
-The back-end converts our linear front-end code into a collection of states
-
-is to compile C,
-
-We separate
-
-Threads are separated and replaced by labels
-
-pass to make it
+Each thread must be split into one function for each jump target, and
+thread-local state must be accessed via the thread \t{struct} shared between
+these functions. Furthermore, because I/O only occurs in the main loop, these
+functions must return immediately after registering new descriptors. As a
+result, each thread is split up into quite a few functions in this paradigm.
+Because this translation from threaded to event-based is quite mechanical, we
+are able to perform it automatically in the He-Man back-end.
 

### 2011-12-13 report/report.tex
Future work.

@@
 gigabytes of RAM, and the client ran on a Lenovo Thinkpad T500 with a
-Intel Core 2 Duo and 8 gigabtyes of RAM.  % XXX: client specs
+Intel Core 2 Duo and 8 gigabytes of RAM.  % XXX: client specs
 The machines were connected by a gigabit Ethernet link.
@@
 
-say something about Memory.hs??
+% see how many threads we can spawn with Memory.hs
 
@@
 
-multithreaded runtime
+We will continue our work on He-Man, and hope to submit it to a programming
+languages conference such as ICFP. We plan to expand the language with the
+ability to declare functions and make non-external function calls within a
+thread. This will make code simpler and shorter by allowing the programmer to
+abstract away common functionality like read and write loops.
 
-write a better allocator
+In the runtime, our current thread \t{struct}s are larger than necessary.
+Because the runtime allocates a \t{struct} on every thread spawn, we should
+replace our \t{malloc} calls with calls to a custom slab allocator. We also plan
+on writing a multithreaded runtime, to allow servicing multiple threads' ready
+events simultaneously. Because our threads don't interact, this does not require
+exposing any locking to the programmer.
 
-less space per thread
+The backend can perform more aggressive optimization on the code blocks. In
+particular, we currently do not allow any control flow except at tail position;
+if both branches of an \t{if} are non-blocking, we should avoid translating it
+into multiple blocks, requiring a yield to the main loop. It should also be
+possible to gather interesting output from the backend, like a graphical
+representation of the generated state machine, for documentation purposes.
 
-decide when to allow internal if statements
-
-We will additionally output a graphical representation of this underlying state
-machine.
-
-observable sharing (?)
+We also hope to implement observable sharing, as in \cite{Gill}, to allow the
+same thread to have distinct spawn sites without code duplication. This does not
+add inefficiency, but does increase the size of our output code.
 

### 2011-12-13 report/report.tex
Wirte a lot for the frontend.

--- 2011-12-13 report/report.tex
@@
 \usepackage{graphicx}
+\usepackage{verbatim}
+\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
 \renewcommand{\t}{\texttt}
@@
 
+The underlying abstract syntax for the front-end He-Man language is a
+fairly conventional imperative, statement based, first order language,
+shown in Figure \ref{fig:front-syntax}. The notable constructs are
+\b{Spawn}, which creates a new thread that takes a list of arguments
+and launches the associated code with those arguments available to it,
+and \b{Wait}, which takes an event expression as an argument and
+blocks until the event has ``completed''. Note that \b{Call} is for
+calling foreign functions; He-Man does not have its own notion of
+function.
+
+Since writing programs using the abstract syntax would be highly
+inconvenient, we provide a monadic library for writing He-Man
+programs, which allows us to leverage Haskell's do notation to write
+He-Man programs using convenient imperative syntax. In this way, we
+are taking advantage of monads being ``programmable semi-colons''.
+
+We define the \t{Prog} monad, which contains a \t{State} monad used to
+maintain a counter for generating unique variable names and a
+\t{Writer} monad that is used to build up a list of statements. We
+then provide \t{Prog} wrappers for statements that add the statement
+to the list of statements being built up by the \t{Writer}. The
+wrappers for \b{Spawn}, \b{Assign}, \b{Wait}, and \b{Exit} simply take
+the obvious arguments, construct a \t{Stmt} and add it to the writer
+list. The \b{Assign} wrapper is an infix operator, and is named
+``.=''. The wrappers for \b{While} and \b{If} take values in \t{Prog}
+instead of statement lists, and run the \t{Writer} to extract the
+lists of statements. The monadic wrapper for \b{Decl} declares a
+variable (with a name made unique by appending a unique number) and
+then returns the \t{Expr} that references the declared variables. This
+makes declaring and using variables very natural. We also provide a
+wrapper for \b{Call} that declares a fresh variable, assigns the
+result of a call to it, and then returns the variable. Then,
+``declaring'' a foreign function consists of writing a convenient
+wrapper function that invokes the \b{Call} wrapper.
+
+At the expression level, \t{Expr} is an instance of the \t{Num} type
+class so that numeric literals and operations can be used in He-Man
+syntax. We also provide infix operators for other useful operations on
+expressions, some of them with mildly awkward names, since the good
+names were already taken by a function with the wrong type.
+
+The result of all of this is that He-Man code can be written in a
+fairly natural style, as shown in Figure \ref{fig:code-ex}.
+
+%We also provide an \b{Assign} wrapper that takes a \t{Prog Expr} as its argument
+
 \begin{figure}[ht]
@@
 \caption{Front-end grammar.}
+\label{fig:front-syntax}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\begin{smallverbatim}
+child_code = declare_thread [("child_fd", Int)] $
+  \[child_fd] -> do
+  ev <- setup_connection child_fd
+  buf <- new_buf bufsize
+  while 1 $ do
+    amt_read <- do_read ev buf bufsize
+    ifthen (amt_read .== 0) exit
+    amt_written <- full_write ev buf amt_read
+    ifthen (amt_written .< amt_read) exit
+\end{smallverbatim}
+\caption{He-Man example from echo server.}
+\label{fig:code-ex}
 \end{figure}

### 2011-12-13 report/report.tex
Finish up front-end

@@
 
-%We also provide an \b{Assign} wrapper that takes a \t{Prog Expr} as its argument
+Although the He-Man target language does not support defining
+functions, code modularity and reuse is easily supported while writing
+He-Man code by defining Haskell functions that produce values in
+\t{Prog}. When invoked, the code of these meta-level ``functions'' is
+included directly at the call-site. This is a very powerful tool, but
+can potentially result in code size explosion if used carelessly.
+
+As an example of the power that Haskell and the monadic style gives
+us, Figure \ref{fig:generic} shows our library routines for simulating
+blocking network operations. \t{do\_nb\_action} takes an event
+\t{Expr} and a \t{Prog Expr} action representing a network call that
+can return \t{EAGAIN} and produces an action that generates code to
+simulate a blocking network operation. This allows us to describe how
+to simulate blocking I/O in a generic way, and then invoke it to
+create our concrete wrappers.
+(\t{.=.} is a wrapper for
+\b{Assign} that takes a \t{Prog Expr} as its second argument and the
+\t{sock\_*} functions invoke \b{Call} wrappers on the appropriate
+network calls).
 
@@
 \label{fig:code-ex}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\begin{smallverbatim}
+do_nb_action :: Expr -> Prog Expr -> Prog Expr
+do_nb_action e action = do
+  res <- action
+  ifthen (res .< 0) $ do
+    while (res .< 0) $ do
+      wait e
+      res .=. action
+  return res
+
+accept (fd, e) =
+  do_nb_action e (sock_accept fd)
+do_read (fd, e) buf size =
+  do_nb_action e (sock_read fd buf size)
+do_write (fd, e) buf size =
+  do_nb_action e (sock_write fd buf size)
+\end{smallverbatim}
+\caption{Implementation of ``blocking'' network calls}
+\label{fig:generic}
 \end{figure}

### 2011-12-13 report/report.tex
Write a bad abstract

@@
 \begin{abstract}
-TODO
+We present He-Man, a Haskell compiled, embedded, domain specific
+language for building scalable network applications in a natural,
+straight-line manner while still emitting efficient event-loop based
+code. He-Man programs are compact and measurements indicate that it
+performs acceptably. 
 \end{abstract}

### 2011-12-13 report/report.tex
conclusions done

@@
 \section{Conclusions}
+Expressing network application logic as a compiled, embedded, domain
+specific language is an elegant and practical way to build scalable
+network applications without having to manually write state-machine
+based event-loop code. Embedding the language in Haskell allows
+powerful metaprogramming and allowed the language to be developed with
+a fraction of the effort of building a full language. Compiling to C
+allows C programmers to write parts of their system (for example, the
+caching layer) in C while writing the network logic in
+straightforward, straight-line He-Man code.
 

### 2011-12-13 report/report.tex
Almost done with back-end.

--- 2011-12-13 report/report.tex
@@
 \usepackage{graphicx}
-\usepackage{verbatim}
-\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
 \renewcommand{\t}{\texttt}
@@
 
-The underlying abstract syntax for the front-end He-Man language is a
-fairly conventional imperative, statement based, first order language,
-shown in Figure \ref{fig:front-syntax}. The notable constructs are
-\b{Spawn}, which creates a new thread that takes a list of arguments
-and launches the associated code with those arguments available to it,
-and \b{Wait}, which takes an event expression as an argument and
-blocks until the event has ``completed''. Note that \b{Call} is for
-calling foreign functions; He-Man does not have its own notion of
-function.
-
-Since writing programs using the abstract syntax would be highly
-inconvenient, we provide a monadic library for writing He-Man
-programs, which allows us to leverage Haskell's do notation to write
-He-Man programs using convenient imperative syntax. In this way, we
-are taking advantage of monads being ``programmable semi-colons''.
-
-We define the \t{Prog} monad, which contains a \t{State} monad used to
-maintain a counter for generating unique variable names and a
-\t{Writer} monad that is used to build up a list of statements. We
-then provide \t{Prog} wrappers for statements that add the statement
-to the list of statements being built up by the \t{Writer}. The
-wrappers for \b{Spawn}, \b{Assign}, \b{Wait}, and \b{Exit} simply take
-the obvious arguments, construct a \t{Stmt} and add it to the writer
-list. The \b{Assign} wrapper is an infix operator, and is named
-``.=''. The wrappers for \b{While} and \b{If} take values in \t{Prog}
-instead of statement lists, and run the \t{Writer} to extract the
-lists of statements. The monadic wrapper for \b{Decl} declares a
-variable (with a name made unique by appending a unique number) and
-then returns the \t{Expr} that references the declared variables. This
-makes declaring and using variables very natural. We also provide a
-wrapper for \b{Call} that declares a fresh variable, assigns the
-result of a call to it, and then returns the variable. Then,
-``declaring'' a foreign function consists of writing a convenient
-wrapper function that invokes the \b{Call} wrapper.
-
-At the expression level, \t{Expr} is an instance of the \t{Num} type
-class so that numeric literals and operations can be used in He-Man
-syntax. We also provide infix operators for other useful operations on
-expressions, some of them with mildly awkward names, since the good
-names were already taken by a function with the wrong type.
-
-The result of all of this is that He-Man code can be written in a
-fairly natural style, as shown in Figure \ref{fig:code-ex}.
-
-Although the He-Man target language does not support defining
-functions, code modularity and reuse is easily supported while writing
-He-Man code by defining Haskell functions that produce values in
-\t{Prog}. When invoked, the code of these meta-level ``functions'' is
-included directly at the call-site. This is a very powerful tool, but
-can potentially result in code size explosion if used carelessly.
-
-As an example of the power that Haskell and the monadic style gives
-us, Figure \ref{fig:generic} shows our library routines for simulating
-blocking network operations. \t{do\_nb\_action} takes an event
-\t{Expr} and a \t{Prog Expr} action representing a network call that
-can return \t{EAGAIN} and produces an action that generates code to
-simulate a blocking network operation. This allows us to describe how
-to simulate blocking I/O in a generic way, and then invoke it to
-create our concrete wrappers.
-(\t{.=.} is a wrapper for
-\b{Assign} that takes a \t{Prog Expr} as its second argument and the
-\t{sock\_*} functions invoke \b{Call} wrappers on the appropriate
-network calls).
-
 \begin{figure}[ht]
@@
 \caption{Front-end grammar.}
-\label{fig:front-syntax}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\begin{smallverbatim}
-child_code = declare_thread [("child_fd", Int)] $
-  \[child_fd] -> do
-  ev <- setup_connection child_fd
-  buf <- new_buf bufsize
-  while 1 $ do
-    amt_read <- do_read ev buf bufsize
-    ifthen (amt_read .== 0) exit
-    amt_written <- full_write ev buf amt_read
-    ifthen (amt_written .< amt_read) exit
-\end{smallverbatim}
-\caption{He-Man example from echo server.}
-\label{fig:code-ex}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\begin{smallverbatim}
-do_nb_action :: Expr -> Prog Expr -> Prog Expr
-do_nb_action e action = do
-  res <- action
-  ifthen (res .< 0) $ do
-    while (res .< 0) $ do
-      wait e
-      res .=. action
-  return res
-
-accept (fd, e) =
-  do_nb_action e (sock_accept fd)
-do_read (fd, e) buf size =
-  do_nb_action e (sock_read fd buf size)
-do_write (fd, e) buf size =
-  do_nb_action e (sock_write fd buf size)
-\end{smallverbatim}
-\caption{Implementation of ``blocking'' network calls}
-\label{fig:generic}
 \end{figure}
@@
 
-The back-end converts our linear front-end code into a collection of states
-
-is to compile C,
-
-It's important that the only blocking operations happen in the main loop
-
-We separate
-
-Threads are separated and replaced by labels
-
-pass to make it
-
-\begin{figure}[ht]
+\begin{figure}[ht]\label{fig:backend}
 \centering
@@
 \end{figure}
+
+The most important task for our back-end is to convert our linear front-end code
+into a collection of states appropriate for an event loop. This occurs in our
+\t{flattenPrgm} pass, which passes through the program in reverse order,
+
+is to compile C,
+
+It's important that the only blocking operations happen in the main loop
+
+We separate
+
+Threads are separated and replaced by labels
+
+This pass generates a correct state machine, albeit one which contains many
+blocks with only tail-position jumps. We follow it up with the \t{optimizeJumps}
+pass, which removes these empty blocks, redirecting all \t{Goto}s to their
+ultimate destinations, and allowing nested \t{If}s in tail position. Nested
+\t{If}s occur frequently in generated code, for example, immediately before an
+\t{if} statement at the end of a \t{while} loop, when the code might next jump
+inside the \t{if} statement, to the top of the \t{while} loop, or after the
+\t{while} loop, depending on multiple tests. Although this pass is, strictly
+speaking, an optimization, it is critical for obtaining high-quality output.
+
+At this point, the program's control flow is correct, but many blocks will
+contain uses of variables which have been declared in previous blocks. The
+\t{collectFrees} pass traverses each block, collecting the variables used but
+not declared in each block. (Recall that our front-end ensures that a variable
+is only ever declared in one location, so there is no ambiguity.) The local
+variables of each thread are the union of the used-not-declared variables in
+each of its blocks. We save this list of local variables in the appropriate
+\i{Thread} container.
+
+Note that we could simply mark all declared variables as thread-local, but
+variables which are only used in same block as their declaration need not be
+stored thread-locally. We want to minimize the number of thread-local variables,
+because this directly impacts the memory overhead of each thread.
+
+From this point, the path to C code is fairly straightforward. For each thread,
+we generate a \t{struct} \t{thread$n$\_t} declaration containing a \t{thread\_t}
+struct with generic thread control information, followed by all thread-local
+variables. Each block is translated to a function taking as argument a
+\t{thread\_t} passed in from the main loop. It casts this to the \t{struct} type
+for its thread in order to access thread-local variables.
+
+\t{Stmt}s are translated straightforwardly; threads are spawned by allocating a
+new \t{struct} for the corresponding thread, setting its continuation to the
+first block in the thread, and adding it to the list of runnable threads.
+\t{Expr}s and \t{Tail}s are similarly straightforward; variables in the thread
+\t{struct} must be replaced by references to their thread-local value, while
+other variables are guaranteed to be block-local. Finally, we must also generate
+a \t{main} function which spawns the first thread and jumps to the main loop.
 
@@
 We will continue our work on He-Man, and hope to submit it to a programming
-languages conference such as ICFP. We plan to expand the language with the
-ability to declare functions and make non-external function calls within a
-thread. This will make code simpler and shorter by allowing the programmer to
-abstract away common functionality like read and write loops.
+languages conference such as ICFP. Most importantly, we plan to rework our
+front-end so as to reject badly-typed programs; currently, we do not ensure
+that, e.g., we do not take the sum of two file handles.
+
+We also plan to expand the language with the ability to declare functions and
+make non-external function calls within a thread. This will make code simpler
+and shorter by allowing the programmer to abstract away common functionality
+like read and write loops.
 

### 2011-12-13 report/report.tex
Cleaned up miscellaneous things in report.

@@
 \begin{abstract}
-We present He-Man, a Haskell compiled, embedded, domain specific
-language for building scalable network applications in a natural,
-straight-line manner while still emitting efficient event-loop based
-code. He-Man programs are compact and measurements indicate that it
-performs acceptably. 
+We present He-Man, a compiled, embedded, domain specific language in Haskell for
+building scalable network applications in a natural, straight-line manner while
+still producing efficient event loop-based code. He-Man programs are compact and
+measurements indicate that it performs acceptably. Our implementation is
+available at
+\url{https://github.com/msullivan/networks-project}.
 \end{abstract}
@@
 
-An improvement is to write multithreaded code, in which each thread serves
+An improvement is to write multithreaded code in which each thread serves
 multiple clients. This is the most popular approach, and is utilized in the
@@
 
+The latest version of He-Man can be found at
+\url{https://github.com/msullivan/networks-project}.
+
 \subsection{Front-end}
@@
 The underlying abstract syntax for the front-end He-Man language is a
-fairly conventional imperative, statement based, first order language,
+fairly conventional imperative, statement-based, first-order language,
 shown in Figure \ref{fig:front-syntax}. The notable constructs are
-\b{Spawn}, which creates a new thread that takes a list of arguments
+\b{Spawn}, creating a new thread, which takes a list of arguments
 and launches the associated code with those arguments available to it,
@@
 He-Man programs using convenient imperative syntax. In this way, we
-are taking advantage of monads being ``programmable semi-colons''.
+are taking advantage of monads as ``programmable semicolons''.
 
@@
 
-At the expression level, \t{Expr} is an instance of the \t{Num} type
-class so that numeric literals and operations can be used in He-Man
+At the expression level, \t{Expr} is an instance of the \t{Num} typeclass 
+so that numeric literals and operations can be used in He-Man
 syntax. We also provide infix operators for other useful operations on
-expressions, some of them with mildly awkward names, since the good
-names were already taken by a function with the wrong type.
+expressions, some of them with mildly awkward names to avoid name conflicts.
 
-The result of all of this is that He-Man code can be written in a
-fairly natural style, as shown in Figure \ref{fig:code-ex}.
+As a result, He-Man code can be written in a fairly natural style with little
+need for normal Haskell code, as shown in Figure \ref{fig:code-ex}.
 
@@
 to simulate blocking I/O in a generic way, and then invoke it to
-create our concrete wrappers.
-(\t{.=.} is a wrapper for
-\b{Assign} that takes a \t{Prog Expr} as its second argument and the
-\t{sock\_*} functions invoke \b{Call} wrappers on the appropriate
-network calls).
+create our concrete wrappers. (\t{.=.} is a wrapper for \b{Assign} 
+that takes a \t{Prog Expr} as its second argument and the \t{sock\_*} 
+functions invoke \b{Call} wrappers on the appropriate network calls.)
 
@@
 \end{smallverbatim}
-\caption{Implementation of ``blocking'' network calls}
+\caption{Implementation of ``blocking'' network calls.}
 \label{fig:generic}
@@
 blocks with only tail-position jumps. We follow it up with the \t{optimizeJumps}
-pass, which removes these empty blocks, redirecting all \t{Goto}s to their
-ultimate destinations, and allowing nested \t{If}s in tail position. Nested
-\t{If}s occur frequently in generated code, for example, immediately before an
+pass, which removes these empty blocks, redirecting all \b{Goto}s to their
+ultimate destinations, and allowing nested \b{If}s in tail position. Nested
+\b{If}s occur frequently in generated code, for example, immediately before an
 \t{if} statement at the end of a \t{while} loop, when the code might next jump
@@
 \t{collectFrees} pass traverses each block, collecting the variables used but
-not declared in each block. (Recall that our front-end ensures that a variable
-is only ever declared in one location, so there is no ambiguity.) The local
-variables of each thread are the union of the used-not-declared variables in
-each of its blocks. We save this list of local variables in the appropriate
-\i{Thread} container.
+not declared in each block. (Our front-end ensures that a variable is only ever
+declared in one location, so there is no ambiguity.) The local variables of each
+thread are the union of the used-not-declared variables in each of its blocks.
+We save this list of local variables in the appropriate \i{Thread} container.
 
@@
 variables which are only used in same block as their declaration need not be
-stored thread-locally. We want to minimize the number of thread-local variables,
-because this directly impacts the memory overhead of each thread.
+retained outside that block. We want to minimize the number of thread-local
+variables, because this directly impacts the memory overhead of each thread.
 
@@
 
-\t{Stmt}s are translated straightforwardly; threads are spawned by allocating a
+\i{Stmt}s are translated straightforwardly; threads are spawned by allocating a
 new \t{struct} for the corresponding thread, setting its continuation to the
 first block in the thread, and adding it to the list of runnable threads.
-\t{Expr}s and \t{Tail}s are similarly straightforward; variables in the thread
+\i{Expr}s and \i{Tail}s are similarly straightforward; variables in the thread
 \t{struct} must be replaced by references to their thread-local value, while
@@
 indefinitely, and on each iteration it removes the thread at the head
-of the runqueue (if it exists) and runs it, and polls for IO events.
+of the runqueue (if it exists) and runs it, and polls for I/O events.
 
@@
 thread that is blocked on it and some arbitrary identifier (in
-practice the file descriptor). To create an event for use with
-nonblocking IO, an event is allocated, marked with the associated file
+practice, the file descriptor). To create an event for use with
+nonblocking I/O, an event is allocated, marked with the associated file
 descriptor, and a new edge-triggered \t{epoll} event is created for the
@@
 descriptor has returned \t{EAGAIN}.
-The actual polling for IO events is done by calling \t{epoll}. For each
-event that has been returned by epoll, if a thread is waiting on the
+The actual polling for I/O events is done by calling \t{epoll}. For each
+event that has been returned by \t{epoll}, if a thread is waiting on the
 event, the thread is added to the runqueue. If the the runqueue is
-empty, the call to epoll is blocking, otherwise it is called with a
+empty, the call to \t{epoll} is blocking, otherwise it is called with a
 timeout of zero seconds, to simply poll for events.
@@
 its structure is obscured by being split into states and the need to
-explicitly handle read and write returning EAGAIN and write returning
+explicitly handle read and write returning \t{EAGAIN} and write returning
 short. The state machine has a total of five states. The He-Man
@@
 Our benchmark measurements are taken using ab, the Apache HTTP server
-benchmarking tool. \cite{ApacheAB} Measurements were collected by
+benchmarking tool \cite{ApacheAB}. Measurements were collected by
 having a variable number of client threads request a file 100 times;
 the experiment was repeated with a 16 kilobyte file and a 128 kilobyte
-file. We ran the tests against both our He-Man http server and Apache
+file. We ran the tests against both our He-Man HTTP server and Apache
 2.2.20, installed with the default Ubuntu 11.10 settings, except for
 client, thread, and process limits being adjusted upwards. Since the
-He-Man runtime is currently only single threaded, we ran the Apache
+He-Man runtime is currently only single-threaded, we ran the Apache
 web server pinned to a single processor. Since the test repeatedly
@@
 The machines were connected by a gigabit Ethernet link.
-
 Unfortunately, ab is not particularly scalable, so we had trouble
-testing with more clients than 500.
+testing with more than 500 clients.
 
@@
 
-\begin{figure*}[!htb]
+\begin{figure*}[!ht]
 \centering
 \includegraphics[width=0.9\textwidth]{128kb_graph.png}
-	\caption{Web server benchmark, 128kb file}
+\caption{Web server benchmark, 128 KB file.}
 \label{fig:graph128}
 \end{figure*}
-\begin{figure*}[!htb]
+\begin{figure*}[!ht]
 \centering
 \includegraphics[width=0.9\textwidth]{16kb_graph.png}
-	\caption{Web server benchmark, 16kb file}
+\caption{Web server benchmark, 16 KB file.}
 \label{fig:graph16}
@@
 when the thread exits). On a 64-bit system, the total overhead comes
-out to 88 bytes, and could easily be reduced further (e.g. by using
-singly linked lists).  For comparison, the default thread stack size
+out to 88 bytes, and could easily be reduced further (e.g., by using
+singly-linked lists).  For comparison, the default thread stack size
 under NPTL on Linux is 2 megabytes.

### 2011-12-13 report/report.tex
Finished backend section. Report finished?

@@
 
-\begin{figure}[ht]\label{fig:backend}
+\begin{figure}[ht]
 \centering
@@
 \caption{Back-end grammar.}
+\label{fig:back-syntax}
 \end{figure}
 
-The most important task for our back-end is to convert our linear front-end code
-into a collection of states appropriate for an event loop. This occurs in our
-\t{flattenPrgm} pass, which passes through the program in reverse order,
+The next several passes convert our front-end grammar to a back-end grammar,
+shown in Figure \ref{fig:back-syntax}, which closely resembles the features of
+the C code that we subsequently generate. Most significantly, a program now
+consists of a list of \i{Block}s of code, each of which belongs to a \i{Thread},
+and a list of \i{Thread}s. \i{Expr}s remain unchanged from the front-end, while
+all control flow is delegated to the \i{Tail} position of a \i{Block} of code.
 
-is to compile C,
+As the grammar makes clear, the most important task for our back-end is to
+convert our linear front-end code into a collection of states appropriate for an
+event loop. This occurs in our \t{flattenPrgm} pass, which passes through the
+program in reverse order with a \t{RWS} monad, converting it to uniquely-named
+blocks whose only control-flow consists of tail-position jumps to other block
+labels. Here, the \t{State} maintains the next block number, the \t{Writer}
+collects lists of \i{Block}s and \i{Thread}s, and the \t{Reader} maintains the
+current thread number (defined as the first block in the current thread).
 
-It's important that the only blocking operations happen in the main loop
+As an illustrative example, consider encountering an \b{If} statement in our
+backwards pass, where we have access to the translation of the later statements,
+as well as the earlier statements. Morally, we would like to translate this to a
+conditional jump to the translation of either the first or second arm, each of
+which is terminated by a jump to the later statements. (Note that
+\t{flattenPrgm} leaves all optimization to the next pass.)
 
-We separate
+We package up the later statements in a new block, and then translate the
+preceding statements, terminated by a tail-position conditional \b{If}. We also
+translate each arm of the original \b{If}, terminated by a tail-position
+\b{Goto} to the block containing the later statements. Importantly, all three of
+these recursive translations may generate new blocks, which will be collected
+automatically by the \t{Writer} monad.
 
-Threads are separated and replaced by labels
+Because blocking I/O may only happen in the main loop, we also terminate a block
+after any blocking I/O call, by inserting a \b{GotoWait} to the block containing
+the later statements. \b{GotoWait} is identical to \b{Goto} except to signal to
+later passes that the thread \emph{must} yield to the main loop at this
+location, and thus the goto must not be optimized away.
 
@@
 
-From this point, the path to C code is fairly straightforward. For each thread,
-we generate a \t{struct} \t{thread$n$\_t} declaration containing a \t{thread\_t}
-struct with generic thread control information, followed by all thread-local
-variables. Each block is translated to a function taking as argument a
-\t{thread\_t} passed in from the main loop. It casts this to the \t{struct} type
-for its thread in order to access thread-local variables.
+From this point, the path to C code is fairly straightforward. Concretely, we
+transform our back-end grammar to the AST provided by the Language.C package,
+which we then output (as C source) to a file. For each thread, we generate a
+\t{struct} \t{thread$n$\_t} declaration containing a \t{thread\_t} struct with
+generic thread control information, followed by all thread-local variables. Each
+block is translated to a function taking as argument a \t{thread\_t} passed in
+from the main loop. It casts this to the \t{struct} type for its thread in order
+to access thread-local variables.
 

### 2012-02-15 src/Language/HeMan/Codegen.hs
Have the generated code refer to event_handle instead of event_t *. I may change how events are represented.

@@ -117,7 +117,7 @@ translateType t =
     IInt -> (cInt,[])
     IBool -> (cBool,[])
     IFD -> (cInt,[])
-    IEvent -> (cType "event_t",[cPtr])
+    IEvent -> (cType "event_handle",[])
     IBuffer -> (cChar,[cPtr])
 
 --}}}

### 2012-02-15 src/rt/mainloop.h

@@ -50,6 +50,7 @@ typedef struct event_t {
 	bool complete;
 	thread_t *thread;
 } event_t;
+typedef event_t * event_handle;
 
 char *new_buf(thread_t *t, int size);
 event_t *mk_nb_event(thread_t *t, int fd, int mode);

### 2012-02-15 src/Language/HeMan/Codegen.hs
Initialize the main loop before spawning the first thread.

@@ -34,9 +34,9 @@ translatePrgm (blocks, threads) =
          [main] ++
          map (translateBlock threads) blocks)
   where main = cFunction "main" [] [cInt] $ cCompound
-                 ((translateStmt [] $ Spawn 0 []) ++
-                  [Left $ cExpr $ cCall (cVar "setup_main_loop") [],
-                   Left $ cExpr $ cCall (cVar "main_loop") [],
+                 ([Left $ cExpr $ cCall (cVar "setup_main_loop") []] ++
+                  (translateStmt [] $ Spawn 0 []) ++
+                  [Left $ cExpr $ cCall (cVar "main_loop") [],
                    Left $ cReturn $ cIntConst 0])
 
 translateThread (name, vars) =

### 2012-02-16 src/rt/lib.h
Cleanups and fixes.

@@ -1,3 +1,6 @@
+#ifndef LIB_H
+#define LIB_H
+
 #include <string.h>
 #include <unistd.h>
 #include <stdio.h>
@@ -26,3 +29,5 @@ int print_int(int n);
 
 // This shouldn't be here. Programs should be able to include other headers
 int http_parse(char *buf, int len);
+
+#endif

### 2012-02-16 src/rt/lib.h
Add a notion of "preparing" an event.

@@ -27,6 +27,8 @@ int make_socket_non_blocking(int sfd);
 int sock_bind_v4(int fd, int addr, int port);
 int print_int(int n);
 
+#define prepare_event(event, new_mode) 0
+
 // This shouldn't be here. Programs should be able to include other headers
 int http_parse(char *buf, int len);
 

### 2012-02-16 src/rt/lib.h
Implement a multithreaded runtime and switch to oneshot level triggered epolls.

@@ -27,7 +27,7 @@ int make_socket_non_blocking(int sfd);
 int sock_bind_v4(int fd, int addr, int port);
 int print_int(int n);
 
-#define prepare_event(event, new_mode) 0
+#define prepare_event(event, new_mode) (event->mode = new_mode)
 
 // This shouldn't be here. Programs should be able to include other headers
 int http_parse(char *buf, int len);

### 2012-02-16 src/rt/mainloop.c

@@ -15,12 +15,20 @@
 #include <libaio.h>
 #include <sys/eventfd.h>
 
+#include <pthread.h>
+
 #include "variable_queue.h"
 #include "mainloop.h"
+#include "thrpool.h"
+
+#define MT_RUNTIME 1
 
 #define MAX_EVENTS 10
 #define MAX_AIO_EVENTS 100 // dunno what this should be...
 #define MAX_ITERS 50
+#define MAX_THREADS 4
+
+int thread_count = MAX_THREADS;
 
 int next_tid = 0;
 
@@ -29,8 +37,11 @@ static struct {
 	io_context_t aio_ctx;
 	int aio_eventfd;
 	struct event_t *aio_dummy_event;
-	//thread_pool_t sched_queue;
+#if MT_RUNTIME
+	thread_pool_t sched_queue;
+#else
 	thread_queue_t sched_queue;
+#endif
 } state;
 
 // XXX: we want something faster than this
@@ -55,9 +66,10 @@ event_t *mk_nb_event(thread_t *thread, int fd, int mode)
 {
 	event_t *e = mk_event();
 	e->id = fd;
+	e->mode = mode;
 
-	if (epoll_ctler(EPOLL_CTL_ADD, fd, mode|EPOLLET, e) < 0)
-		fail(1, "epoll_ctl");
+	if (epoll_ctler(EPOLL_CTL_ADD, fd, 0, e) < 0)
+		fail(1, "epoll_ctl: %d", fd);
 
 	Q_INSERT_TAIL(&thread->nb_events, e, gc_link);
 
@@ -93,10 +105,15 @@ void free_thread(thread_t *thread)
 }
 
 
-void register_event(thread_t *thread, event_t *event)
+void register_event(thread_t *thread, event_t *e)
 {
-	Q_INSERT_TAIL(&thread->pending_events, event, q_link);
-	event->thread = thread;
+	// helgrind will report a data race having to do with this, but
+	// I'm pretty sure it's actually fine. The event pointer is handed
+	// off through epoll.
+	e->thread = thread;
+	// XXX: this only makes sense for epoll events, not AIO ones
+	if (epoll_ctler(EPOLL_CTL_MOD, e->id, e->mode|EPOLLONESHOT, e) < 0)
+		fail(1, "epoll_ctl: %d", e->id);
 }
 
 int epoll_ctler(int op, int fd, uint32_t events, void *ptr)
[…]

### 2012-02-16 src/rt/mainloop.h

@@ -5,6 +5,7 @@
 #include <stdint.h>
 #include <libaio.h>
 #include "variable_queue.h"
+#include "thrpool.h"
 
 extern int next_tid;
 
@@ -32,20 +33,18 @@ typedef struct buf_t {
 } buf_t;
 
 typedef struct thread_t {
+	work_item_t work_item;
 	Q_NEW_LINK(thread_t) q_link;
 	int tid;
-	event_queue_t pending_events;
 	event_queue_t nb_events;
 	buf_queue_t bufs;
-	struct event_t *finished_event;
 	thread_cont *cont;
 } thread_t;
 
-// XXX: track read/write event interest
 typedef struct event_t {
-	Q_NEW_LINK(event_t) q_link;
 	Q_NEW_LINK(event_t) gc_link;
 	int id;
+	int mode;
 	bool complete;
 	thread_t *thread;
 } event_t;

### 2012-02-16 src/rt/mainloop.c
Switch to having every thread call epoll instead of being thrpool based.

@@ -19,7 +19,6 @@
 
 #include "variable_queue.h"
 #include "mainloop.h"
-#include "thrpool.h"
 
 #define MT_RUNTIME 1
 
@@ -38,11 +37,21 @@ static struct {
 	int aio_eventfd;
 	struct event_t *aio_dummy_event;
 #if MT_RUNTIME
-	thread_pool_t sched_queue;
-#else
-	thread_queue_t sched_queue;
+	pthread_mutex_t sched_lock;
 #endif
+	thread_queue_t sched_queue;
 } state;
+
+static inline void sched_lock(void) {
+#if MT_RUNTIME
+	pthread_mutex_lock(&state.sched_lock);
+#endif
+}
+static inline void sched_unlock(void) {
+#if MT_RUNTIME
+	pthread_mutex_unlock(&state.sched_lock);
+#endif
+}
 
 // XXX: we want something faster than this
 // and maybe want to recover from failure
@@ -126,11 +135,9 @@ int epoll_ctler(int op, int fd, uint32_t events, void *ptr)
 
 void make_runnable(thread_t *t)
 {
-#if MT_RUNTIME
-	thread_pool_push(&state.sched_queue, t);
-#else
+	sched_lock();
 	Q_INSERT_TAIL(&state.sched_queue, t, q_link);
-#endif
+	sched_unlock();
 }
 
 
@@ -208,29 +215,41 @@ static void run_thread(void *threadp)
 }
 
 // A single threaded main loop
-void main_loop(void)
+void *work_loop(void *p)
 {
 	for (;;) {
-#if MT_RUNTIME
-		do_poll(false);
-#else
 		thread_t *thread;
+		sched_lock();
 		if ((thread = Q_GET_HEAD(&state.sched_queue))) {
 			Q_REMOVE(&state.sched_queue, thread, q_link);
+			sched_unlock();
 			run_thread(thread);
+			sched_lock();
 		}
-		do_poll(!Q_GET_HEAD(&state.sched_queue));
-#endif
+		bool can_sleep = !Q_GET_HEAD(&state.sched_queue);
+		sched_unlock();
+		do_poll(can_sleep);
 	}
+	return NULL;
 }
+
+void main_loop(void)
+{
+#if MT_RUNTIME
+	pthread_t thread;
+	for (int i = 0; i < thread_count - 1; i++) {
+		int ret = pthread_create(&thread, NULL, work_loop, NULL);
[…]

### 2012-02-16 src/rt/mainloop.h

@@ -5,7 +5,6 @@
 #include <stdint.h>
 #include <libaio.h>
 #include "variable_queue.h"
-#include "thrpool.h"
 
 extern int next_tid;
 
@@ -33,7 +32,6 @@ typedef struct buf_t {
 } buf_t;
 
 typedef struct thread_t {
-	work_item_t work_item;
 	Q_NEW_LINK(thread_t) q_link;
 	int tid;
 	event_queue_t nb_events;

### 2012-02-17 src/Language/HeMan/Syntax.hs
Add a logical negation function.

@@ -179,6 +179,8 @@ infixl 6 +*
 -- Pointer arithmetic
 (+*) :: BufferE -> IntE -> BufferE
 (+*) = typ2 $ Arith Plus
+notE :: BoolE -> BoolE
+notE = typ1 $ ArithUnop Not
 
 -- Class for types we can do a failure test on
 class ExprFailable a where

### 2012-02-17 src/Language/HeMan/Lib.hs
Implement a version of do_nb_action that relies on level triggering.

@@ -81,22 +81,23 @@ kO_RDONLY = constant "O_RDONLY"
 
 kEAGAIN = constant "EAGAIN"
 
-
-do_nb_action :: ExprFailable a => Expr Event -> Prog (Expr a) -> Prog (Expr a)
-do_nb_action e action = do
-  res <- action
-  ifE' (isFailure res) $ do
-    while (isFailure res) $ do
-      wait e
-      res .=. action
+-- This implementation depends on level triggered semantics
+do_nb_action :: ExprFailable a =>
+                Type a -> IntE -> Expr Event -> Prog (Expr a) ->
+                Prog (Expr a)
+do_nb_action t mode e action = do
+  res <- var "result" t (E (NumLit (-1))) -- HACK! bad!
+  while (isFailure res) $ do
+    prepare_event e mode
+    wait e
+    res .=. action
   return res
 
-accept (fd, e) = do_nb_action e (prepare_event e kEVENT_RD >> sock_accept fd)
-do_read (fd, e) buf size = do_nb_action e
-                           (prepare_event e kEVENT_RD >> sock_read fd buf size)
-do_write (fd, e) buf size = do_nb_action e
-                            (prepare_event e kEVENT_WR >>
-                             sock_write fd buf size)
+accept (fd, e) = do_nb_action FD kEVENT_RD e (sock_accept fd)
+do_read (fd, e) buf size = do_nb_action Int kEVENT_RD e
+                           (sock_read fd buf size)
+do_write (fd, e) buf size = do_nb_action Int kEVENT_WR e
+                            (sock_write fd buf size)
 
 -- TODO remove this
 full_write_naive ev buf size = do

### 2012-02-17 src/Language/HeMan/Codegen.hs
Have GotoWait know the event being slept on. Setting the continuation before registering the event is critical for the MT runtime.

@@ -100,7 +100,7 @@ translateTail vars tail =
   case tail of
     Exit -> [cleanup, return 0]
     Goto target -> [jump target, return 1]
-    GotoWait target -> [jump target, return 0]
+    GotoWait e target -> [jump target, register e, return 0]
     If e (ss1,t1) (ss2,t2) -> [Left $
       cIfThen (translateExpr vars e)
         (cCompound $ transSs ss1 ++ translateTail vars t1)
@@ -109,6 +109,8 @@ translateTail vars tail =
           jump target = Left $ cExpr (cAssign
                                       (cArrow (cVar "thread") "cont")
                                       (cVar (blockName target)))
+          register e = Left $ cExpr $ cCall (cVar "register_event") $
+                       [translateExpr vars CurThread, translateExpr vars e]
           cleanup = Left $ cExpr (cCall (cVar "free_thread") [cVar "thread"])
           transSs = concatMap (translateStmt vars)
 

### 2012-02-17 src/Language/HeMan/Pretty.hs

@@ -89,7 +89,7 @@ instance PP Tail where
     If expr (ss,tail) (ss',tail') ->
       text "if" <+> parens (pretty expr) $$ nest 2 (pretty ss $$ pretty tail) 
       $$ text "else" $$ nest 2 (pretty ss' $$ pretty tail')
-    GotoWait label -> text "gotoWait" <+> int label
+    GotoWait e label -> text "gotoWait" <+> pretty e <+> int label
     Goto label -> text "goto" <+> int label
     Exit -> text "exit"
 

### 2012-02-21 src/rt/mainloop.c
Some cleanup of element names, dead code.

@@ -17,7 +17,6 @@
 
 #include "variable_queue.h"
 #include "mainloop.h"
-//#include "thrpool.h"
 
 #define MAX_EVENTS 10
 #define MAX_AIO_EVENTS 100 // dunno what this should be...
@@ -40,8 +39,8 @@ static struct {
 char *new_buf(thread_t *thread, int size) {
 	buf_t *buf = malloc(sizeof(buf_t) + size);
 	if (!buf) fail(1, "allocating buffer");
-	Q_INIT_ELEM(buf, q_link);
-	Q_INSERT_TAIL(&thread->bufs, buf, q_link);
+	Q_INIT_ELEM(buf, gc_link);
+	Q_INSERT_TAIL(&thread->bufs, buf, gc_link);
 	return buf->buffer;
 }
 
@@ -61,7 +60,7 @@ event_t *mk_nb_event(thread_t *thread, int fd, int mode)
 	if (epoll_ctler(EPOLL_CTL_ADD, fd, mode|EPOLLET, e) < 0)
 		fail(1, "epoll_ctl");
 
-	Q_INSERT_TAIL(&thread->nb_events, e, q_link);
+	Q_INSERT_TAIL(&thread->nb_events, e, gc_link);
 
 	return e;
 }
@@ -87,7 +86,7 @@ void free_thread(thread_t *thread)
 	}
 	buf_t *buf;
 	while ((buf = Q_GET_HEAD(&thread->bufs))) {
-		Q_REMOVE(&thread->bufs, buf, q_link);
+		Q_REMOVE(&thread->bufs, buf, gc_link);
 		free_buf(buf);
 	}
 
@@ -112,7 +111,6 @@ int epoll_ctler(int op, int fd, uint32_t events, void *ptr)
 void make_runnable(thread_t *t)
 {
 	Q_INSERT_TAIL(&state.sched_queue, t, q_link);
-	//thread_pool_push(&state.sched_queue, t);
 }
 
 

### 2012-02-21 src/rt/mainloop.h

@@ -27,12 +27,11 @@ Q_NEW_HEAD(buf_queue_t, buf_t);
 typedef bool thread_cont(struct thread_t *thread);
 
 typedef struct buf_t {
-	Q_NEW_LINK(buf_t) q_link;
+	Q_NEW_LINK(buf_t) gc_link;
 	char buffer[];
 } buf_t;
 
 typedef struct thread_t {
-	//work_item_t work_item;
 	Q_NEW_LINK(thread_t) q_link;
 	int tid;
 	event_queue_t pending_events;

### 2012-02-21 src/rt/mainloop.c
Get rid of the useless global expect_aio.

@@ -25,7 +25,6 @@
 int next_tid = 0;
 
 static struct {
-	bool expect_aio;
 	int epoll_fd;
 	io_context_t aio_ctx;
 	int aio_eventfd;
@@ -163,20 +162,19 @@ static void do_poll(bool can_sleep)
 	int aio_cnt = io_getevents(state.aio_ctx, 0, MAX_EVENTS,
 	                           aio_events, NULL);
 	if (aio_cnt < 0) { fail2(1, -aio_cnt, "io_getevents"); }
-	state.expect_aio = aio_cnt == MAX_EVENTS;
+	bool expect_aio = aio_cnt == MAX_EVENTS;
 	for (int i = 0; i < aio_cnt; i++) {
 		handle_aio_event(&aio_events[i]);
 	}
 
 	// If we aren't expecting aio, block indefinitely, otherwise
 	// just poll.
-	int epoll_timeout = state.expect_aio || !can_sleep ? 0 : -1;
+	int epoll_timeout = expect_aio || !can_sleep ? 0 : -1;
 	int epoll_cnt = epoll_wait(state.epoll_fd, epoll_events,
 	                           MAX_EVENTS, epoll_timeout);
 	for (int i = 0; i < epoll_cnt; i++) {
 		if (epoll_events[i].data.ptr == state.aio_dummy_event) {
 			uint64_t eventfd_val;
-			state.expect_aio = true;
 			if (read(state.aio_eventfd, &eventfd_val,
 			         sizeof(eventfd_val)) < 0)
 				fail(1, "eventfd read");

### 2012-02-22 src/Language/HeMan/Codegen.hs
Put thread spawning code in braces, to allow multiple spawns in a block. Closes #11.

@@ -82,7 +82,8 @@ translateStmt vars stmt =
         [Right $ cDecl [typ] indirs x (Just $ transE e)]
     Assign e1 e2 -> [Left $ cExpr (cAssign (transE e1) (transE e2))]
     Exp e -> [Left $ cExpr $ transE e]
-    Spawn thd args -> makeT:setupTVars ++ setupTRun where
+    Spawn thd args -> [Left $ cCompound stmts] where
+      stmts = makeT:setupTVars ++ setupTRun
       thread = threadName thd
       makeT = Right $ cDecl [cType thread] [cPtr] "new_thread" $ 
         Just $ cCall (cVar $ "mk_" ++ thread) []

### 2012-03-02 paper/paper.tex
add some notes to the paper tex

--- 2011-12-13 report/report.tex
@@
-\documentclass[preprint,11pt]{sigplanconf}
+\documentclass[preprint,10pt]{sigplanconf}
 \usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
@@
 
+% TODO: talk about type system, runtime improvements, optimizations
+
 \begin{abstract}
+% XXX: better abstract;
+% and where should we put the code link?
+% we should also move it to some better URL
+
 We present He-Man, a compiled, embedded, domain specific language in Haskell for
@@
 
+%% XXX: Should we put this at the end?
 \section{Related work}
@@
 
+%% XXX: update
 The He-Man runtime is written in C and provides a state machine based
@@
 
+% XXX: this has changed
 The He-Man runtime main loop is very simple, then: it loops
@@
 
+% XXX: this is not quite true; also who cares?
 The runtime also has support for Linux Kernel Asynchronous IO for file
@@
 
-To evaluate He-Man, we built two network applications with He-Man and
-compared them to similar network applications built in C.
-We also discuss He-Man's memory overhead.
+To evaluate He-Man, we built several network applications with He-Man
+and compared them to similar network applications built in C.  We also
+discuss He-Man's memory overhead.
 
@@
 
+%% XXX: needs serious work
 \section{Future work}

### 2012-03-02 src/rt/mainloop.c
Seperate out MT and non MT runtime into a link time decision.

@@ -20,8 +20,6 @@
 #include "variable_queue.h"
 #include "mainloop.h"
 
-#define MT_RUNTIME 0
-
 #define MAX_EVENTS 10
 #define MAX_AIO_EVENTS 100 // dunno what this should be...
 #define MAX_ITERS 50

### 2012-03-02 src/testing/tester.c
Add a sutpid tester program that I used to test the echo server.

@@ -1,0 +1,37 @@
+// Really stupid program for doing some testing of an echo server
+// I run it like:
+// nc.traditional server port -e ./tester
+
+#include <stdlib.h>
+#include <unistd.h>
+#include <fcntl.h>
+#include <errno.h>
+
+int make_socket_non_blocking(int sfd)
+{
+	int flags, s;
+
+	flags = fcntl(sfd, F_GETFL, 0);
+	if (flags == -1) exit(1);
+
+	flags |= O_NONBLOCK;
+	s = fcntl (sfd, F_SETFL, flags);
+	if (s == -1) exit(1);
+
+	return 0;
+}
+
+
+int main() {
+	char buf[1024];
+	srand(getpid());
+	make_socket_non_blocking(0);
+	while (1) {
+		if (write(1, "hello\n", 6) < 0 && errno != EAGAIN) {
+			//perror("write");
+			return 1;
+		}
+		int err = read(0, buf, sizeof(buf));
+		sleep(rand() % 10);
+	}
+}

### 2012-03-03 paper/paper.tex
some notes and shit.

@@
 
+% Contributions; probably should be elsewhere
+We identify our contributions as:
+\begin{enumerate}
+
+\item We present a case study of compiled embedded domain specific
+  language design and implementation.
+\end{enumerate}
+
 %% XXX: Should we put this at the end?
@@
 \section{The event loop model}
+%% XXX: it isn't really clear what's going on in this section; it
+%% doesn't /really/ talk about event loop stuff in general. It is
+%% talking about one possible way to write event based things (the way
+%% He-Man generates).
 
-Networking applications require a very constrained form of concurrency. Because
-I/O (both network and disk) latency dominates the computation time, each
-client's thread of execution spends most of its time waiting on I/O. In
-particular, we would like to always yield to a client no longer blocked on I/O;
-we need not yield to blocked clients, nor do we need arbitrary interleaving of
-execution.
+Networking applications require a very constrained form of
+concurrency. Because I/O (both network and disk) latency dominates the
+computation time, each client's thread of execution spends most of its
+time waiting on I/O. In particular, we would like to always yield to a
+client no longer blocked on I/O; we need not yield to blocked clients,
+nor do we (usually) need arbitrary interleaving of execution.
 
+%% XXX: do wereally want to talk about this?
 As of version 2.5.44, Linux has supported \t{epoll} for non-blocking

### 2012-03-05 paper/paper.tex
tweak

@@
 %\subtitle{I said hey, what's goin' on}
-\authorinfo{Carlo Angiuli}{}{cangiuli@cs.cmu.edu}
-\authorinfo{Michael Sullivan}{}{mjsulliv@andrew.cmu.edu}
-\date{December 12, 2011}
+\authorinfo{Carlo Angiuli}{Carnegie Mellon University}{cangiuli@cs.cmu.edu}
+\authorinfo{Michael Sullivan}{Carnegie Mellon University}{mjsulliv@cs.cmu.edu}
 
@@
 available at
-\url{https://github.com/msullivan/networks-project}.
+\url{https://github.com/msullivan/he-man}.
 \end{abstract}
@@
 
-%% XXX: do wereally want to talk about this?
+%% XXX: do we really want to talk about this?
 As of version 2.5.44, Linux has supported \t{epoll} for non-blocking
@@
 The latest version of He-Man can be found at
-\url{https://github.com/msullivan/networks-project}.
+\url{https://github.com/msullivan/he-man}.
 

### 2012-03-06 paper/paper.tex
some minor updates

@@
 Continuation functions take a pointer to the thread information,
+perform some computation (such as making non-blocking I/O calls),
 update the continuation field in the thread data, and return whether
@@
 registers itself as blocking on an event and returns false; to exit,
-it frees the thread structure and returns false.
-Running a thread, then, consists of repeatedly invoking the thread's
-continuation function until it returns false. To prevent a
-thread that does not need to block from starving the rest of the
-system, after some number of invocations of the continuation
-(currently 50), the thread is paused and put on the back of the
-runqueue.
+it frees the thread structure and returns false.  Running a thread,
+then, consists of repeatedly invoking the thread's continuation
+function until it returns false. To prevent a thread that does not
+need to block from starving the rest of the system, after some number
+of invocations of the continuation (currently 50), the thread is
+paused and put on the back of the runqueue.
 
 % XXX: this has changed
-The He-Man runtime main loop is very simple, then: it loops
-indefinitely, and on each iteration it removes the thread at the head
-of the runqueue (if it exists) and runs it, and polls for I/O events.
+The He-Man runtime main loop is very simple, then: each worker thread
+loops indefinitely, and on each iteration it removes the thread at the
+head of the runqueue (if it exists) and runs it, and polls for I/O
+events.
 
@@
 associated pointer data.
+
+% XXX: not true anymore
 We use edge-triggered instead of level-triggered so that we can wait

### 2012-03-06 paper/paper.tex
Rewrite lots of the runtime section.

@@
 
-%% XXX: update
 The He-Man runtime is written in C and provides a state machine based
@@
 
-% XXX: this has changed
 The He-Man runtime main loop is very simple, then: each worker thread
@@
 head of the runqueue (if it exists) and runs it, and polls for I/O
-events.
+events. A global runqueue is shared between threads; a more clever
+runtime could have per-worker-thread runqueues and do some sort of
+work stealing.
 
 The runtime provides a blocking mechanism based on an event
-abstraction. To the runtime, an event consists of a pointer to a
-thread that is blocked on it and some arbitrary identifier (in
-practice, the file descriptor). To create an event for use with
-nonblocking I/O, an event is allocated, marked with the associated file
-descriptor, and a new edge-triggered \t{epoll} event is created for the
-file descriptor using \t{epoll\_ctl}, with the event structure as the
-associated pointer data.
+abstraction. To the runtime, an event consists of a type tag, a
+pointer to a thread that is blocked on it and some event type specific
+data (such as a file descriptor or a pointer to an associated
+channel). Events are level-triggered: blocking on an event will return
+whenever the event ``is ready'', as opposed to when the event
+``becomes ready'' as in edge-triggered (using edge-triggered semantics
+causes extra difficulties in an environment with multiple threads
+processing events).
 
-% XXX: not true anymore
-We use edge-triggered instead of level-triggered so that we can wait
-on both read and write events for a given file descriptor without
-having to adjust which we are waiting for at runtime, and because
-He-Man's programming model makes it easy to meet the requirement that
-blocking on a file descriptor is only done after an operation on the
-descriptor has returned \t{EAGAIN}.
-The actual polling for I/O events is done by calling \t{epoll}. For each
-event that has been returned by \t{epoll}, if a thread is waiting on the
-event, the thread is added to the runqueue. If the the runqueue is
-empty, the call to \t{epoll} is blocking, otherwise it is called with a
+Currently, the He-Man runtime provides support for two types of
+events: socket/pipe file I/O and asynchronous He-Man channels.  The
+runtime has partial support for Linux Kernel Asynchronous IO for file
+IO, using a Linux \t{eventfd} to allow \t{epoll} to wait on AIO
+events. Adding new types of events that an application requires would
+be simple, but require some runtime hacking.
+
+The actual polling for I/O events is done by calling \t{epoll}
+(although it would not be difficult to adapt it to use BSD's
+\t{kqueue} or the more standard \t{poll} or \t{select}).  For each
+event that has been returned by \t{epoll}, if a thread is waiting on
+the event, the thread is added to the runqueue. If the the runqueue is
+empty, the call to \t{epoll} blocks, otherwise it is called with a
 timeout of zero seconds, to simply poll for events.
-
-% XXX: this is not quite true; also who cares?
-The runtime also has support for Linux Kernel Asynchronous IO for file
-IO, using a Linux \t{eventfd} to allow \t{epoll} to wait on AIO
-events. This support is currently unused, since it was not actually
-necessary to acheive acceptable performance for the web server.
 
@@
 written in as an event loop driven explicit state machine using the
-He-Man runtime. The C version, which is fairly tightly written, is 98
-lines of source, according to SLOCCount. It is fairly readable, but
-its structure is obscured by being split into states and the need to
-explicitly handle read and write returning \t{EAGAIN} and write returning
-short. The state machine has a total of five states. The He-Man
-version is 33 lines long and is a very clear representation of the
-control flow of each individual thread. (Additionally, the He-Man
-library has 17 lines of code implementing generic routines for
-simulating blocking IO using nonblocking IO and wait.) The total line
-counts, however, somewhat understate the difference between the
-versions, since the boilerplate for setting up connections is about
-the same for both versions. When just looking at the core read/write
-loop, the difference is much more apparent: the C version is 38 lines
-of code, and the He-Man version is 5.
+He-Man runtime. The C version, which is fairly tightly written and
+which uses the He-Man runtime in its capacity as a somewhat generic
+event based IO framework, is 98 lines of source, according to
+SLOCCount. It is fairly readable, but its structure is obscured by
+being split into states and the need to explicitly handle read and
+write returning \t{EAGAIN} and write returning short. The state
+machine has a total of five states. The He-Man version is 33 lines
+long and is a very clear representation of the control flow of each
+individual thread. (Additionally, the He-Man library has 17 lines of
+code implementing generic routines for simulating blocking IO using
+nonblocking IO and wait.) The total line counts, however, somewhat
+understate the difference between the versions, since the boilerplate
+for setting up connections is about the same for both versions. When
+just looking at the core read/write loop, the difference is much more
+apparent: the C version is 38 lines of code, and the He-Man version is
+5.
 
+% XXX: update
 The generated code is substantially less tight than the C version,

### 2012-03-06 src/Language/HeMan/Lib.hs
Make do_nb_action a bit less generic to make it easier to include in the paper. Also shifts the FD hack for accept to a slightly less bad one.

@@ -1,6 +1,7 @@
 module Language.HeMan.Lib where
 
 import Language.HeMan.Syntax
+import Control.Applicative
 
 -- Coreish runtime functions
 mk_nb_event :: FdE -> IntE -> Prog EventE
@@ -37,9 +38,9 @@ sock_listen :: FdE -> IntE -> Prog IntE
 sock_listen fd q_limit =
   call (CFn "listen") Int (fd, q_limit)
 
-sock_accept :: FdE -> Prog FdE
+sock_accept :: FdE -> Prog IntE
 sock_accept fd =
-  callName "accepted_fd" (CFn "accept") FD (fd, num 0, num 0)
+  callName "accepted_fd" (CFn "accept") Int (fd, num 0, num 0)
 
 sock_read :: FdE -> BufferE -> IntE -> Prog IntE
 sock_read fd buf len =
@@ -120,7 +121,10 @@ data_to_int :: DataE -> IntE
 data_to_int d = callE (CFn "data_to_int") Int (d)
 
 
--- For debugging
+mk_fd :: IntE -> FdE
+mk_fd = unsafeExprCoerce
+
+-- for debugging
 print_int n  =
   call' (CFn "print_int") (n)
 
@@ -141,23 +145,23 @@ kO_RDONLY = constant "O_RDONLY"
 kEAGAIN = constant "EAGAIN"
 
 -- This implementation depends on level triggered semantics
-do_nb_action :: ExprFailable a =>
-                Type a -> IntE -> Expr Event -> Prog (Expr a) ->
-                Prog (Expr a)
-do_nb_action t mode e action = do
-  res <- var "result" t (E (NumLit (-1))) -- HACK! bad!
+do_nb_action :: IntE -> Expr Event -> Prog IntE ->
+                Prog IntE
+do_nb_action mode e action = do
+  res <- var "result" Int (-1)
   err <- var "err" Int kEAGAIN
-  while (isFailure res .&& err .== kEAGAIN) $ do
+  while (res .< 0 .&& err .== kEAGAIN) $ do
     prepare_event e mode
     wait e
     res .=. action
     err .= errno
   return res
 
-accept (fd, e) = do_nb_action FD kEVENT_RD e (sock_accept fd)
-do_read (fd, e) buf size = do_nb_action Int kEVENT_RD e
+accept (fd, e) = mk_fd <$> do_nb_action kEVENT_RD e
+                           (sock_accept fd)
+do_read (fd, e) buf size = do_nb_action kEVENT_RD e
[…]

### 2012-03-06 paper/paper.tex
Update the blocking network call example

@@
 \begin{smallverbatim}
-do_nb_action :: Expr -> Prog Expr -> Prog Expr
-do_nb_action e action = do
-  res <- action
-  ifthen (res .< 0) $ do
+do_nb_action :: IntE -> EventE -> Prog IntE -> Prog IntE
+do_nb_action mode e action = do
+  res <- var "result" Int (-1)
   while (res .< 0) $ do
+    set_nb_event_mode e mode
     wait e
@@
 accept (fd, e) =
-  do_nb_action e (sock_accept fd)
+  mk_fd <$> do_nb_action kEVENT_RD e (sock_accept fd)
 do_read (fd, e) buf size =
-  do_nb_action e (sock_read fd buf size)
+  do_nb_action kEVENT_RD e (sock_read fd buf size)
 do_write (fd, e) buf size =
-  do_nb_action e (sock_write fd buf size)
+  do_nb_action kEVENT_WR e (sock_write fd buf size)
 \end{smallverbatim}

### 2012-03-06 src/Language/HeMan/Lib.hs

@@ -145,8 +145,7 @@ kO_RDONLY = constant "O_RDONLY"
 kEAGAIN = constant "EAGAIN"
 
 -- This implementation depends on level triggered semantics
-do_nb_action :: IntE -> Expr Event -> Prog IntE ->
-                Prog IntE
+do_nb_action :: IntE -> EventE -> Prog IntE -> Prog IntE
 do_nb_action mode e action = do
   res <- var "result" Int (-1)
   err <- var "err" Int kEAGAIN
@@ -157,12 +156,12 @@ do_nb_action mode e action = do
     err .= errno
   return res
 
-accept (fd, e) = mk_fd <$> do_nb_action kEVENT_RD e
-                           (sock_accept fd)
-do_read (fd, e) buf size = do_nb_action kEVENT_RD e
-                           (sock_read fd buf size)
-do_write (fd, e) buf size = do_nb_action kEVENT_WR e
-                            (sock_write fd buf size)
+accept (fd, e) =
+  mk_fd <$> do_nb_action kEVENT_RD e (sock_accept fd)
+do_read (fd, e) buf size =
+  do_nb_action kEVENT_RD e (sock_read fd buf size)
+do_write (fd, e) buf size =
+  do_nb_action kEVENT_WR e (sock_write fd buf size)
 
 -- TODO remove this
 full_write_naive ev buf size = do

### 2012-03-06 paper/paper.tex
Excising some old text; added citation.

--- 2012-03-06 paper/paper.tex
@@
 
-Our observation is that the event loop model, in programming languages parlance,
-is essentially trampolining continuation-passing-style, which can be obtained by
-mechanical translation from the straight-line threaded code. This suggests it is
-possible to obtain the best of both worlds---a threaded programming model with
-the performance of event loops.
+Our observation is that the event loop model is simply trampolining
+continuation-passing-style, % TODO citation
+which can be obtained by mechanical translation from the straight-line threaded
+code. This suggests it is possible to obtain the best of both worlds---a
+threaded programming model with the performance of event loops.
 
@@
 
-We will continue our work on He-Man, and hope to submit it to a programming
-languages conference such as ICFP. Most importantly, we plan to rework our
-front-end so as to reject badly-typed programs; currently, we do not ensure
-that, e.g., we do not take the sum of two file handles.
-
 We also plan to expand the language with the ability to declare functions and
@@
 exposing any locking to the programmer.
-
-The backend can perform more aggressive optimization on the code blocks. In
-particular, we currently do not allow any control flow except at tail position;
-if both branches of an \t{if} are non-blocking, we should avoid translating it
-into multiple blocks, requiring a yield to the main loop. It should also be
-possible to gather interesting output from the backend, like a graphical
-representation of the generated state machine, for documentation purposes.
 

### 2012-03-06 paper/paper.tex
Fix the grammars a bit.
Some tweaks and whatnot.

@@
 The underlying abstract syntax for the front-end He-Man language is a
-fairly conventional imperative, statement-based, first-order language,
-shown in Figure \ref{fig:front-syntax}. The notable constructs are
-\b{Spawn}, creating a new thread, which takes a list of arguments
-and launches the associated code with those arguments available to it,
-and \b{Wait}, which takes an event expression as an argument and
-blocks until the event has ``completed''. Note that \b{Call} is for
-calling foreign functions; He-Man does not have its own notion of
-function.
+fairly conventional imperative, block-structured, statement-based,
+first-order language, shown in Figure \ref{fig:front-syntax}. The
+notable constructs are \b{Spawn}, creating a new thread, which takes a
+list of arguments and launches the associated code with those
+arguments available to it, and \b{Wait}, which takes an event
+expression as an argument and blocks until the event is
+``ready''. Note that \b{Call} is for calling foreign functions; He-Man
+does not have its own notion of function.
 
@@
 \a{\b{StringLit} \i{String}}
-\a{\b{Var} \i{String}}
-\a{\b{CurThread}}
+\a{\b{Var} \i{Var}}
 \i{Type} 
-\p{\b{Int} $\mid$ \b{Bool} $\mid$ \b{String}
-   $\mid$ \b{Buffer} $\mid$ \b{Event}}
+\p{\b{Int} $\mid$ \b{Bool} $\mid$ \b{FD}
+   $\mid$ \b{Buffer} $\mid$ \b{Event} $\mid$ $\cdots$}
 }
-\caption{Front-end grammar.}
+\caption{Front-end abstract syntax.}
 \label{fig:front-syntax}
@@
 \begin{smallverbatim}
-do_nb_action :: IntE -> EventE -> Prog IntE -> Prog IntE
+do_nb_action :: IntE -> EventE -> Prog IntE
+                -> Prog IntE
 do_nb_action mode e action = do
@@
 }
-\caption{Back-end grammar.}
+\caption{Back-end abstract syntax.}
 \label{fig:back-syntax}
@@
 Because the runtime allocates a \t{struct} on every thread spawn, we should
-replace our \t{malloc} calls with calls to a custom slab allocator. We also plan
-on writing a multithreaded runtime, to allow servicing multiple threads' ready
-events simultaneously. Because our threads don't interact, this does not require
-exposing any locking to the programmer.
+replace our \t{malloc} calls with calls to a custom slab allocator.
 

### 2012-03-06 paper/paper.tex
spellcheck

@@
 for multiplexed, non-blocking I/O, such as \t{select}, \t{poll}, \t{epoll}, or
-\t{kquery}. In this model, all blocking occurs at a single call site in the
+\t{kqueue}. In this model, all blocking occurs at a single call site in the
 code. The programmer is responsible for explicitly maintaining per-client state
@@
 To test the suitability of our system for more complicated programs
-and to demonstrate seperating network logic and application logic, we
+and to demonstrate separating network logic and application logic, we
 built a simple (and bad) HTTP 1.0 server. The network logic is written
@@
 forms a complete HTTP request and, if so, fills a buffer with the
-filename to serve. The entire server is 54 lines of He-Man haskell and
+file name to serve. The entire server is 54 lines of He-Man Haskell and
 27 lines of C. It suffers from a number of major deficiencies, but is

### 2012-03-06 src/Language/HeMan/Codegen.hs
Rename DExpr to IExpr to keep Bob from yelling at us about the phrase dynamic typing

@@ -1,7 +1,7 @@
 module Language.HeMan.Codegen (codegen) where
 
 import Language.HeMan.Syntax
-  (Prim(..), ArithOp(..), ArithUnop(..), RelnOp(..), IType(..), DExpr(..))
+  (Prim(..), ArithOp(..), ArithUnop(..), RelnOp(..), IType(..), IExpr(..))
 import Language.HeMan.Backend
   (Block, Thread, Stmt(..), Tail(..))
 import Language.HeMan.CLib

### 2012-03-06 src/Language/HeMan/Pretty.hs

@@ -48,7 +48,7 @@ instance PP Lang.Stmt where
     Lang.Wait expr -> text "wait" <> parens (pretty expr)
     Lang.Exit -> text "exit"
 
-instance PP Lang.DExpr where
+instance PP Lang.IExpr where
   pretty e = case e of
     Lang.Call (Lang.CFn prim) exprs ->
       text prim <> (parens $ commaSeparated (map pretty exprs))

### 2012-03-06 src/Language/HeMan/Syntax.hs

@@ -23,23 +23,23 @@ data Data
 data Msg
 data Channel
 
-data Stmt = Decl VDecl DExpr
-          | While DExpr Block
-          | If DExpr Block Block
-          | Spawn IThreadCode [DExpr]
-          | Assign DExpr DExpr
-          | Exp DExpr
-          | Wait DExpr
+data Stmt = Decl VDecl IExpr
+          | While IExpr Block
+          | If IExpr Block Block
+          | Spawn IThreadCode [IExpr]
+          | Assign IExpr IExpr
+          | Exp IExpr
+          | Wait IExpr
           | Exit
           deriving (Eq, Ord, Show)
 
 data IType = IInt | IBool | IFD | IBuffer | IEvent | IData | IMsg | IChannel
           deriving (Eq, Ord, Show)
 
-data DExpr = Call Prim [DExpr]
-          | Arith ArithOp DExpr DExpr
-          | ArithUnop ArithUnop DExpr
-          | RelnOp RelnOp DExpr DExpr
+data IExpr = Call Prim [IExpr]
+          | Arith ArithOp IExpr IExpr
+          | ArithUnop ArithUnop IExpr
+          | RelnOp RelnOp IExpr IExpr
           | Constant String
           | NumLit Integer
           | StringLit String
@@ -82,7 +82,7 @@ mkIType Data = IData
 mkIType Msg = IMsg
 mkIType Channel = IChannel
 
-newtype Expr a = E DExpr
+newtype Expr a = E IExpr
                deriving (Eq, Ord, Show)
 
 type IntE = Expr Int
@@ -93,10 +93,10 @@ type EventE = Expr Event
 type ChannelE = Expr Channel
 type DataE = Expr Data
 
--- Lift functions of DExprs to ones over Exprs
-typ1 :: (DExpr -> DExpr) -> (Expr a -> Expr b)
+-- Lift functions of IExprs to ones over Exprs
+typ1 :: (IExpr -> IExpr) -> (Expr a -> Expr b)
 typ1 f (E e1) = E (f e1)
-typ2 :: (DExpr -> DExpr -> DExpr) -> (Expr a -> Expr b -> Expr c)
+typ2 :: (IExpr -> IExpr -> IExpr) -> (Expr a -> Expr b -> Expr c)
 typ2 f (E e1) (E e2) = E (f e1 e2)
 
[…]

### 2012-03-06 paper/paper.tex
Update back-end grammar.
Add in the listings stuff.

@@
 \usepackage{verbatim}
+
+\usepackage{listings}
+\lstset{
+  language=Haskell,
+  basicstyle=\small\ttfamily,
+  frame=trbl,
+  basewidth={0.5em,0.45em},
+%  commentstyle=\ttfamily\color{ForestGreen},
+  literate=*{+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
+            {>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
+            {\\\\}{{\char`\\\char`\\}}1
+            {->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
+            {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
+            %{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
+            {>>}{{>>}}2 {>>=}{{>>=}}2
+            {|}{{$\mid$}}1
+}
+
+\lstnewenvironment{egcode}
+  {\lstset{frame=TRBL}}
+  {\lstset{frame=trbl}}
+
+\lstnewenvironment{code}
+  {\lstset{frame=}}
+  {}
+
 \newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
@@
 \centering
-\begin{smallverbatim}
+\begin{code}
 child_code = declare_thread [("child_fd", Int)] $
@@
     ifthen (amt_written .< amt_read) exit
-\end{smallverbatim}
+\end{code}
 \caption{He-Man example from echo server.}
@@
 \centering
-\begin{smallverbatim}
+\begin{code}
 do_nb_action :: IntE -> EventE -> Prog IntE
@@
   do_nb_action kEVENT_WR e (sock_write fd buf size)
-\end{smallverbatim}
+\end{code}
 \caption{Implementation of ``blocking'' network calls.}

### 2012-03-06 src/Language/HeMan/Syntax.hs

@@ -93,7 +93,7 @@ type EventE = Expr Event
 type ChannelE = Expr Channel
 type DataE = Expr Data
 
--- Lift functions of IExprs to ones over Exprs
+-- Lift functions over IExprs to ones over Exprs
 typ1 :: (IExpr -> IExpr) -> (Expr a -> Expr b)
 typ1 f (E e1) = E (f e1)
 typ2 :: (IExpr -> IExpr -> IExpr) -> (Expr a -> Expr b -> Expr c)

### 2012-03-06 src/Language/HeMan/Syntax.hs
Rename mkIType to toIType.

@@ -72,15 +72,15 @@ data Type a where
   Msg :: Type Msg
   Channel :: Type Channel
 
-mkIType :: Type a -> IType
-mkIType Int = IInt
-mkIType Bool = IBool
-mkIType FD = IFD
-mkIType Buffer = IBuffer
-mkIType Event = IEvent
-mkIType Data = IData
-mkIType Msg = IMsg
-mkIType Channel = IChannel
+toIType :: Type a -> IType
+toIType Int = IInt
+toIType Bool = IBool
+toIType FD = IFD
+toIType Buffer = IBuffer
+toIType Event = IEvent
+toIType Data = IData
+toIType Msg = IMsg
+toIType Channel = IChannel
 
 newtype Expr a = E IExpr
                deriving (Eq, Ord, Show)
@@ -145,24 +145,24 @@ instance ArgPacket (Expr a, Expr b, Expr c, Expr d, Expr e)
 instance ArgDecls () () where
   toVDecl () = []
 instance ArgDecls (TVDecl a) (OneArg a) where
-  toVDecl (x1, t1) = [(x1, mkIType t1)]
+  toVDecl (x1, t1) = [(x1, toIType t1)]
 
 -- Generic binary arg decls; this makes us need UndecidableInstances
 instance (ArgDecls b c) => ArgDecls (TVDecl a, b) (a, c) where
-  toVDecl ((x, t), xts) = (x, mkIType t) : toVDecl xts
+  toVDecl ((x, t), xts) = (x, toIType t) : toVDecl xts
 
 -- Hacky 3,4,5-ary arg decls
 instance ArgDecls (TVDecl a, TVDecl b, TVDecl c) (a, b, c) where
   toVDecl ((x1, t1), (x2, t2), (x3, t3)) =
-    [(x1, mkIType t1), (x2, mkIType t2), (x3, mkIType t3)]
+    [(x1, toIType t1), (x2, toIType t2), (x3, toIType t3)]
 instance ArgDecls (TVDecl a, TVDecl b, TVDecl c, TVDecl d) (a, b, c, d) where
   toVDecl ((x1, t1), (x2, t2), (x3, t3), (x4, t4)) =
-    [(x1, mkIType t1), (x2, mkIType t2), (x3, mkIType t3), (x4, mkIType t4)]
+    [(x1, toIType t1), (x2, toIType t2), (x3, toIType t3), (x4, toIType t4)]
[…]

### 2012-03-06 paper/paper.tex
add a section about the type system

@@
 
+\subsection{Type system}
+
+It is important to provide some sort of type discipline for He-Man
+programs. Providing a type system in the front end allows us to
+guarentee that the C we output does not contain type errors, vastly
+simplifying development. We also take advantage of this to provide
+type abstractions above and beyond C's limited type system. % XXX: bad wording
+For example, He-Man's type system contains an abstract
+$\b{FD}$ type that is distinct from the type of integers.
+
+We offload the work of type checking He-Man programs to the Haskell
+type checker. Our method for doing this is fairly standard, following
+Leijen and Meijer \cite{Leijen:1999:DSE:1267936.1267945} in using
+phantom types to enforce typing.
+
+A fragment of the code implementing the He-Man type system is shown in
+Figure \ref{fig:types}. The internal type and expression
+representations are the datatypes \t{IExpr} and \t{IType}, and
+typed expressions are represented with the \t{Expr} data type
+constructor, which is a trival wrapper around \t{IExpr} parameterized
+over a phantom type. Operations on typed expressions have type
+signatures that restrict them to the proper types, and then peel away
+the wrapper to work on the \t{IExpr} underneath.
+
+% XXX: say something about GADTs??
+Types are represented by the \t{Type} GADT that is also parameterized
+over a phantom type representing the underlying type. Expression
+builder functions will convert \t{Type} values are to \t{Itype} using
+the \t{toItype} function, which does a simple pass-through conversion.
+The decision to make \t{Type} a GADT instead of a simple wrapper with
+constructor functions in the same manner as \t{Expr} is for purely
+aesthetic reasons: we felt that type annotations in He-Man code were
+more clearly readable if they were captialized, which required that
+they be datatype constructors.
+\footnote{The authors believe that this may be the worst reason to
+  choose to use GADTs to date.}
+
+\begin{figure}[ht]
+\centering
+\begin{code}
+-- Some dummy types for the phantom types
+data FD
+data Buffer
+data Event
+
+-- The Expr types
+newtype Expr a = E IExpr
+type IntE = Expr Int
+type FdE = Expr FD
+
+data Type a where
+  Int :: Type Int
+  FD :: Type FD
+  -- ...
+
+toIType :: Type a -> IType
+toIType Int = IInt
+toIType FD = IFD
+-- ...
+
+-- Lift functions over IExprs to ones over Exprs
+typ1 :: (IExpr -> IExpr) -> (Expr a -> Expr b)
+typ1 f (E e1) = E (f e1)
+typ2 :: (IExpr -> IExpr -> IExpr) ->
+        (Expr a -> Expr b -> Expr c)
+typ2 f (E e1) (E e2) = E (f e1 e2)
+
+-- Some typed He-Man functions
+(.<) :: IntE -> IntE -> BoolE
+(.<) = typ2 $ RelnOp Lt
+(.==) :: Expr a -> Expr a -> BoolE
+(.==) = typ2 $ RelnOp Eq
+var :: String -> Type a -> Expr a -> Prog (Expr a)
+var = ...
+\end{code}
+\caption{He-Man expression typing} % XXX
+\label{fig:types}
+\end{figure}
+
+TODO: Talk about argument packets.
+
 \subsection{Back-end}

### 2012-03-06 paper/paper.tex
Add important citation.

@@
 
-We present He-Man, a compiled, embedded, domain specific language in Haskell for
+We present He-Man \cite{HeMan}, a compiled, embedded, domain specific language in Haskell for
 building scalable network applications in a natural, straight-line manner while
@@
 \p{\b{Goto} \i{label}}
-\a{\b{GotoWait} \i{label}}
-\a{\b{If} \i{Expr} \i{Tail} \i{Tail}}
+\a{\b{GotoWait} \i{Expr} \i{label}}
+\a{\b{If} \i{Expr} ([\i{Stmt}], \i{Tail}) ([\i{Stmt}], \i{Tail})}
 \a{\b{Exit}}

### 2012-03-06 paper/paper.tex
Added citations; partly cleaned up backend section.

--- 2012-03-06 paper/paper.tex
@@
 \usepackage{verbatim}
-
-\usepackage{listings}
-\lstset{
-  language=Haskell,
-  basicstyle=\small\ttfamily,
-  frame=trbl,
-  basewidth={0.5em,0.45em},
-%  commentstyle=\ttfamily\color{ForestGreen},
-  literate=*{+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
-            {>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
-            {\\\\}{{\char`\\\char`\\}}1
-            {->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
-            {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
-            %{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
-            {>>}{{>>}}2 {>>=}{{>>=}}2
-            {|}{{$\mid$}}1
-}
-
-\lstnewenvironment{egcode}
-  {\lstset{frame=TRBL}}
-  {\lstset{frame=trbl}}
-
-\lstnewenvironment{code}
-  {\lstset{frame=}}
-  {}
-
 \newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
@@
 \centering
-\begin{code}
+\begin{smallverbatim}
 child_code = declare_thread [("child_fd", Int)] $
@@
     ifthen (amt_written .< amt_read) exit
-\end{code}
+\end{smallverbatim}
 \caption{He-Man example from echo server.}
@@
 \centering
-\begin{code}
+\begin{smallverbatim}
 do_nb_action :: IntE -> EventE -> Prog IntE
@@
   do_nb_action kEVENT_WR e (sock_write fd buf size)
-\end{code}
+\end{smallverbatim}
 \caption{Implementation of ``blocking'' network calls.}
@@
 The next several passes convert our front-end grammar to a back-end grammar,
-shown in Figure \ref{fig:back-syntax}, which closely resembles the features of
-the C code that we subsequently generate. Most significantly, a program now
-consists of a list of \i{Block}s of code, each of which belongs to a \i{Thread},
-and a list of \i{Thread}s. \i{Expr}s remain unchanged from the front-end, while
-all control flow is delegated to the \i{Tail} position of a \i{Block} of code.
+shown in Figure \ref{fig:back-syntax}, which closely resembles the C code we
+ultimately generate. In particular, a back-end program is a list of code
+\i{Block}s, each belonging to one of a list of \i{Thread}s. \i{Expr}s remain
+unchanged from the front-end, but critically, control flow is relegated to
+\i{Tail} position in each block.
 
-As the grammar makes clear, the most important task for our back-end is to
-convert our linear front-end code into a collection of states appropriate for an
-event loop. This occurs in our \t{flattenPrgm} pass, which passes through the
-program in reverse order with a \t{RWS} monad, converting it to uniquely-named
-blocks whose only control-flow consists of tail-position jumps to other block
-labels. Here, the \t{State} maintains the next block number, the \t{Writer}
-collects lists of \i{Block}s and \i{Thread}s, and the \t{Reader} maintains the
-current thread number (defined as the first block in the current thread).
+Our back-end's main task is to convert our linear front-end code into a state
+machine of code blocks. This occurs in our \t{flattenPrgm} pass, which passes
+through program \i{Stmt}s in reverse order, generating uniquely-named blocks
+whose only control-flow consists of (non-nested) tail-position jumps to other
+blocks. Threads are identified with their first block.
+% TODO discuss the transformations
 
-As an illustrative example, consider encountering an \b{If} statement in our
-backwards pass, where we have access to the translation of the later statements,
-as well as the earlier statements. Morally, we would like to translate this to a
-conditional jump to the translation of either the first or second arm, each of
-which is terminated by a jump to the later statements. (Note that
-\t{flattenPrgm} leaves all optimization to the next pass.)
+%As an illustrative example, consider encountering an \b{If} statement in our
+%backwards pass, where we have access to the translation of the later
+%statements, as well as the earlier statements. Morally, we would like to
+%translate this to a conditional jump to the translation of either the first or
+%second arm, each of which is terminated by a jump to the later statements.
+%(Note that \t{flattenPrgm} leaves all optimization to the next pass.)
 
-We package up the later statements in a new block, and then translate the
-preceding statements, terminated by a tail-position conditional \b{If}. We also
-translate each arm of the original \b{If}, terminated by a tail-position
-\b{Goto} to the block containing the later statements. Importantly, all three of
-these recursive translations may generate new blocks, which will be collected
-automatically by the \t{Writer} monad.
+%We package up the later statements in a new block, and then translate the
+%preceding statements, terminated by a tail-position conditional \b{If}. We also
+%translate each arm of the original \b{If}, terminated by a tail-position
+%\b{Goto} to the block containing the later statements. Importantly, all three
+%of these recursive translations may generate new blocks, which will be
+%collected automatically by the \t{Writer} monad.
 
-Because blocking I/O may only happen in the main loop, we also terminate a block
-after any blocking I/O call, by inserting a \b{GotoWait} to the block containing
-the later statements. \b{GotoWait} is identical to \b{Goto} except to signal to
-later passes that the thread \emph{must} yield to the main loop at this
-location, and thus the goto must not be optimized away.
+%Because blocking I/O may only happen in the main loop, we also terminate a
+%block after any blocking I/O call, by inserting a \b{GotoWait} to the block
+%containing the later statements. \b{GotoWait} is identical to \b{Goto} except
+%to signal to later passes that the thread \emph{must} yield to the main loop at
+%this location, and thus the goto must not be optimized away.
+% TODO for concurrency-related reasons
 
-This pass generates a correct state machine, albeit one which contains many
-blocks with only tail-position jumps. We follow it up with the \t{optimizeJumps}
-pass, which removes these empty blocks, redirecting all \b{Goto}s to their
-ultimate destinations, and allowing nested \b{If}s in tail position. Nested
-\b{If}s occur frequently in generated code, for example, immediately before an
-\t{if} statement at the end of a \t{while} loop, when the code might next jump
-inside the \t{if} statement, to the top of the \t{while} loop, or after the
-\t{while} loop, depending on multiple tests. Although this pass is, strictly
-speaking, an optimization, it is critical for obtaining high-quality output.
+This pass generates a correct, albeit unnecessarily large, state machine with
+many empty blocks. Over the next few passes, our goal is to shrink this state
+machine as much as possible, because yielding to the main loop is the only
+overhead in the He-Man system. Although these passes are, strictly speaking,
+optimizations, they are very important for performance.
 
-At this point, the program's control flow is correct, but many blocks will
-contain uses of variables which have been declared in previous blocks. The
-\t{collectFrees} pass traverses each block, collecting the variables used but
-not declared in each block. (Our front-end ensures that a variable is only ever
-declared in one location, so there is no ambiguity.) The local variables of each
-thread are the union of the used-not-declared variables in each of its blocks.
-We save this list of local variables in the appropriate \i{Thread} container.
+Our \t{simplifyJumps} pass traverses the program collecting empty blocks,
+redirecting control flow to its ultimate destination. In particular, we often
+introduce nested \b{If}s in tail-position, for example, before an \t{if}
+statement at the end of a \t{while} loop, where control resumes in one of three
+blocks.
 
-Note that we could simply mark all declared variables as thread-local, but
-variables which are only used in same block as their declaration need not be
-retained outside that block. We want to minimize the number of thread-local
-variables, because this directly impacts the memory overhead of each thread.
+At this point, we need to eliminate spurious control flow edges from our state
+machine in order to enable further block elimination. To do this, we perform
+block-local constant propagation and folding, and dead code elimination. At face
+value, it is not important for us to optimize straight-line code: our output is
+further optimized by a C compiler, and execution time is dominated by I/O, not
+computation. However, simple constant folding eliminates many \t{if} statements
+%TODO examples, justification
+
+Once our 
+% TODO fuseBlocks pass
+
+At this point, many blocks refer to variables declared in other blocks. The
+\t{collectFrees} pass traverses each block, collecting the variables each thread
+must close over. Because our front-end ensures each variable declaration is
+unique, we need not consider control flow; the set of thread-local variables is
+merely the union of the used-not-declared variables in each of its blocks.  We
+save this list of local variables in the appropriate \i{Thread} container. (Of
+course, it would be adequate to mark all variables thread-local, but this would
+negatively impact our per-thread memory overhead.)
 

### 2012-03-06 paper/paper.tex

@@
 \end{figure}
+
+\subsection{Type system}
+
+It is important to provide some sort of type discipline for He-Man
+programs. Providing a type system in the front end allows us to
+guarentee that the C we output does not contain type errors, vastly
+simplifying development. We also take advantage of this to provide
+type abstractions above and beyond C's limited type system. % XXX: bad wording
+For example, He-Man's type system contains an abstract
+$\b{FD}$ type that is distinct from the type of integers.
+
+We offload the work of type checking He-Man programs to the Haskell
+type checker. Our method for doing this is fairly standard, following
+Leijen and Meijer \cite{Leijen:1999:DSE:1267936.1267945} in using
+phantom types to enforce typing.
+
+A fragment of the code implementing the He-Man type system is shown in
+Figure \ref{fig:types}. The internal type and expression
+representations are the datatypes \t{IExpr} and \t{IType}, and
+typed expressions are represented with the \t{Expr} data type
+constructor, which is a trival wrapper around \t{IExpr} parameterized
+over a phantom type. Operations on typed expressions have type
+signatures that restrict them to the proper types, and then peel away
+the wrapper to work on the \t{IExpr} underneath.
+
+% XXX: say something about GADTs??
+Types are represented by the \t{Type} GADT that is also parameterized
+over a phantom type representing the underlying type. Expression
+builder functions will convert \t{Type} values are to \t{Itype} using
+the \t{toItype} function, which does a simple pass-through conversion.
+The decision to make \t{Type} a GADT instead of a simple wrapper with
+constructor functions in the same manner as \t{Expr} is for purely
+aesthetic reasons: we felt that type annotations in He-Man code were
+more clearly readable if they were captialized, which required that
+they be datatype constructors.
+\footnote{The authors believe that this may be the worst reason to
+  choose to use GADTs to date.}
+
+\begin{figure}[ht]
+\centering
+\begin{code}
+-- Some dummy types for the phantom types
+data FD
+data Buffer
+data Event
+
+-- The Expr types
+newtype Expr a = E IExpr
+type IntE = Expr Int
+type FdE = Expr FD
+
+data Type a where
+  Int :: Type Int
+  FD :: Type FD
+  -- ...
+
+toIType :: Type a -> IType
+toIType Int = IInt
+toIType FD = IFD
+-- ...
+
+-- Lift functions over IExprs to ones over Exprs
+typ1 :: (IExpr -> IExpr) -> (Expr a -> Expr b)
+typ1 f (E e1) = E (f e1)
+typ2 :: (IExpr -> IExpr -> IExpr) ->
+        (Expr a -> Expr b -> Expr c)
+typ2 f (E e1) (E e2) = E (f e1 e2)
+
+-- Some typed He-Man functions
+(.<) :: IntE -> IntE -> BoolE
+(.<) = typ2 $ RelnOp Lt
+(.==) :: Expr a -> Expr a -> BoolE
+(.==) = typ2 $ RelnOp Eq
+var :: String -> Type a -> Expr a -> Prog (Expr a)
+var = ...
+\end{code}
+\caption{He-Man expression typing} % XXX
+\label{fig:types}
+\end{figure}
+
+TODO: Talk about argument packets.
 

### 2012-03-06 src/Language/HeMan/Syntax.hs
Use [a] as the constructor for unary arg packets.

@@ -109,17 +109,12 @@ class ArgPacket a b | a -> b, b -> a where
 class ArgDecls a b | a -> b, b -> a where
   toVDecl :: a -> [VDecl]
 
--- A dummy one argument constructor for use in guarenteeing that
--- ArgPacket and ArgDecl type classes have a one to one mapping
--- between their parameters.
-data OneArg a
-
 -- This is some ugly shit.
 -- Nullary and unary arg packets
 instance ArgPacket () () where
   toIExprList () = []
   makeVars [] = ()
-instance ArgPacket (Expr a) (OneArg a) where
+instance ArgPacket (Expr a) [a] where
   toIExprList (E x) = [x]
   makeVars [x] = E $ Var x
 
@@ -144,7 +139,7 @@ instance ArgPacket (Expr a, Expr b, Expr c, Expr d, Expr e)
 -- Nullary and unary arg decls
 instance ArgDecls () () where
   toVDecl () = []
-instance ArgDecls (TVDecl a) (OneArg a) where
+instance ArgDecls (TVDecl a) [a] where
   toVDecl (x1, t1) = [(x1, toIType t1)]
 
 -- Generic binary arg decls; this makes us need UndecidableInstances

### 2012-03-06 paper/paper.tex
Changed font size for reality check.
Add arg packet figure.

@@
 
+%XXX: new title?
 \title{He-Man: A domain-specific language for lightweight concurrency}
@@
 % and where should we put the code link?
-% we should also move it to some better URL
 
@@
 
+\begin{figure}[ht]
+\centering
+\begin{code}
+class ArgPacket a b | a -> b, b -> a where
+  toIExprList :: a -> [IExpr]
+  makeVars :: [String] -> a
+class ArgDecls a b | a -> b, b -> a where
+  toVDecl :: a -> [VDecl]
+
+instance ArgPacket (Expr a) [a] where
+  toIExprList (E x) = [x]
+  makeVars [x] = E (Var x)
+instance ArgPacket (Expr a, Expr b) (a, b) where
+  toIExprList (E x1, E x2) = [x1, x2]
+  makeVars [x1, x2] = (E (Var x1), E (Var x2))
+-- ...
+
+instance ArgDecls (TVDecl a) [a] where
+  toVDecl (x1, t1) = [(x1, toIType t1)]
+instance ArgDecls (TVDecl a, TVDecl b) (a, b) where
+  toVDecl ((x1, t1), (x2, t2)) =
+    [(x1, toIType t1), (x2, toIType t2)]
+-- ...
+
+declare_thread :: (ArgPacket a c, ArgDecls b c) =>
+                  b -> (a -> Prog ()) ->
+                  ThreadCode c
+declare_thread = ...
+spawn :: ArgPacket a b => ThreadCode b -> a ->
+                          Prog ()
+spawn = ...
+call :: ArgPacket a b => Prim -> Type c -> a ->
+                         Prog (Expr c)
+call = ...
+
+\end{code}
+\caption{He-Man argument packets} % XXX
+\label{fig:arg-packets}
+\end{figure}
+
 \subsection{Back-end}

### 2012-03-06 paper/paper.tex
some stuff; going to change decls

@@
-\documentclass[preprint,10pt]{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 \usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
@@
 \begin{code}
-child_code = declare_thread [("child_fd", Int)] $
+child_code = declare_thread (Int) $
   \[child_fd] -> do
@@
 
-We offload the work of type checking He-Man programs to the Haskell
-type checker. Our method for doing this is fairly standard, following
-Leijen and Meijer \cite{Leijen:1999:DSE:1267936.1267945} in using
-phantom types to enforce typing.
+We offload the work of type checking He-Man programs to the Haskell type
+checker. Our method for doing this is fairly standard, following Leijen and
+Meijer \cite{LeijenMeijer} in using phantom types to enforce typing.
 
@@
 
-TODO: Talk about argument packets.
+Type checking foreign function calls and thread declaration and
+spawning is somewhat more interesting.
+% and maybe a bit novel? I dunno.
+For foreign function calls, our main concern was being able to write a
+single wrapper for call that handles a varying number of typed
+arguments. To this end, we created an \t{ArgPacket} type class that
+requires the ability to convert a packet into a list of \t{IExpr} and
+to construct an \t{ArgPacket} from a list of variable names. Instances
+of \t{ArgPacket} are provided for packets of $0$ to $5$ \t{Expr}
+arguments. Now we can parameterize a \t{Call} wrapper over an
+\t{ArgPacket} to provide a \t{Expr} based interface.
+
+For thread declarations, we can produce a similar \t{ArgDecl} class
+and associated instances that can convert a packet of \t{Type}s into a
+list of \t{IType}s.
+
+We want more than this, however: we want to ensure that both the code
+for a declared thread uses its arguments properly at the declared
+types code and that spawning threads provides arguments of the correct
+types. To do this, we need some way to extract all of the phantom
+types of the \t{Expr}s and \t{Type}s in \t{ArgPacket}s and
+\t{ArgDecl}s
 
@@
 class ArgDecls a b | a -> b, b -> a where
-  toVDecl :: a -> [VDecl]
+  toITypeList :: a -> [IType]
 
@@
 
-instance ArgDecls (TVDecl a) [a] where
-  toVDecl (x1, t1) = [(x1, toIType t1)]
-instance ArgDecls (TVDecl a, TVDecl b) (a, b) where
-  toVDecl ((x1, t1), (x2, t2)) =
-    [(x1, toIType t1), (x2, toIType t2)]
+instance ArgDecls (Type a) [a] where
+  toITypeList t1 = [toIType t1]
+instance ArgDecls (Type a, Type b) (a, b) where
+  toITypeList (t1, t2) = [toIType t1, toIType t2]
 -- ...
@@
 under NPTL on Linux is 2 megabytes.
+%% XXX update with new info?
 

### 2012-03-06 src/Echo.hs
Revamp how args work. Gets rid of ArgDecl.

@@ -22,7 +22,7 @@ setup_listener port = do
   e <- mk_nb_event fd kEVENT_RD
   return (fd, e)
 
-child_code = declare_thread ("child_fd", FD) $
+child_code = declare_thread FD $
   \child_fd -> do
   ev <- setup_connection child_fd
   buf <- new_buf bufsize

### 2012-03-06 src/Echo2.hs

@@ -22,7 +22,7 @@ setup_listener port = do
   e <- mk_nb_event fd kEVENT_RD
   return (fd, e)
 
-writer_code = declare_thread (("child_fd", FD), ("channel", Channel)) $
+writer_code = declare_thread (FD, Channel) $
   \(fd, ch) -> do
     ev <- setup_connection fd
     while 1 $ do
@@ -40,7 +40,7 @@ duplicate_buf buf sz =
      memcpy new_buf buf sz
      return new_buf
 
-child_code = declare_thread ("child_fd", FD) $
+child_code = declare_thread (FD) $
   \child_fd -> do
   ev <- setup_connection child_fd
   ch <- new_channel

### 2012-03-06 src/Http.hs

@@ -38,7 +38,7 @@ parse_request buf ev = do
     parse_result .=. http_parse buf request_size
   return parse_result
 
-child_code = declare_thread ("child_fd", FD) $
+child_code = declare_thread (FD) $
   \child_fd -> do
   ev <- setup_connection child_fd
   buf <- new_buf (bufsize+1) -- pfbbt.

### 2012-03-06 src/Language/HeMan/Syntax.hs

@@ -106,59 +106,52 @@ unsafeExprCoerce (E x) = E x
 class ArgPacket a b | a -> b, b -> a where
   toIExprList :: a -> [IExpr]
   makeVars :: [String] -> a
-class ArgDecls a b | a -> b, b -> a where
-  toVDecl :: a -> [VDecl]
+  toITypeList :: b -> [IType]
 
 -- This is some ugly shit.
 -- Nullary and unary arg packets
 instance ArgPacket () () where
   toIExprList () = []
   makeVars [] = ()
-instance ArgPacket (Expr a) [a] where
-  toIExprList (E x) = [x]
+  toITypeList () = []
+instance ArgPacket (Expr a) (Type a) where
+  toIExprList (E e) = [e]
   makeVars [x] = E $ Var x
+  toITypeList t = [toIType t]
 
 -- Generic binary packets; this makes us need UndecidableInstances
-instance (ArgPacket b c) => ArgPacket (Expr a, b) (a, c) where
-  toIExprList (E x, xs) = x : toIExprList xs
+instance (ArgPacket b c) => ArgPacket (Expr a, b) (Type a, c) where
+  toIExprList (E e, es) = e : toIExprList es
   makeVars (x : xs) = (E $ Var x, makeVars xs)
+  toITypeList (t, ts) = toIType t : toITypeList ts
+
+{-
+instance ArgPacket (Expr a, Expr b) (Type a, Type b) where
+  toIExprList (E e1, E e2) = [e1, e2]
+  makeVars [x1, x2] = (E $ Var x1, E $ Var x2)
+  toITypeList (t1, t2) = [toIType t1, toIType t2]
+-}
 
 -- Hacky 3,4,5-ary packets
-instance ArgPacket (Expr a, Expr b, Expr c) (a, b, c) where
-  toIExprList (E x1, E x2, E x3) = [x1, x2, x3]
+instance ArgPacket (Expr a, Expr b, Expr c) (Type a, Type b, Type c) where
+  toIExprList (E e1, E e2, E e3) = [e1, e2, e3]
   makeVars [x1, x2, x3] = (E $ Var x1, E $ Var x2, E $ Var x3)
-instance ArgPacket (Expr a, Expr b, Expr c, Expr d) (a, b, c, d) where
-  toIExprList (E x1, E x2, E x3, E x4) = [x1, x2, x3, x4]
[…]

### 2012-03-07 src/Language/HeMan/Syntax.hs
Use more descriptive type variable names.

@@ -103,10 +103,10 @@ unsafeExprCoerce :: Expr a -> Expr b
 unsafeExprCoerce (E x) = E x
 
 -- TODO: Document this trick.
-class ArgPacket a b | a -> b, b -> a where
-  toIExprList :: a -> [IExpr]
-  makeVars :: [String] -> a
-  toITypeList :: b -> [IType]
+class ArgPacket exps tys | exps -> tys, tys -> exps where
+  toIExprList :: exps -> [IExpr]
+  makeVars :: [String] -> exps
+  toITypeList :: tys -> [IType]
 
 -- This is some ugly shit.
 -- Nullary and unary arg packets
@@ -269,7 +269,9 @@ exit = add Exit
 wait :: EventE  -> Prog ()
 wait (E e) = add $ Wait e
 
-spawn :: ArgPacket a b => ThreadCode a -> a -> Prog ()
+spawn :: ArgPacket exps tys =>
+         ThreadCode exps -> exps ->
+         Prog ()
 spawn (Thr thread) args = add $ Spawn thread (toIExprList args)
 
 extract :: Prog a -> Prog (a, [Stmt])
@@ -295,7 +297,9 @@ call' fn args = add $ Exp (Call fn (toIExprList args))
 callName :: ArgPacket a b => String -> Prim -> Type c -> a -> Prog (Expr c)
 callName name fn t args = var name t (E $ Call fn (toIExprList args))
 
-call :: ArgPacket a b => Prim -> Type c -> a -> Prog (Expr c)
+call :: ArgPacket exps tys =>
+        Prim -> Type c -> exps ->
+        Prog (Expr c)
 call fn t args = callName "tmp" fn t args
 
 callE :: ArgPacket a b => Prim -> Type c -> a -> (Expr c)
@@ -303,8 +307,9 @@ callE fn t args = E $ Call fn (toIExprList args)
 
 
 -- Helper to construct a ThreadCode
-declare_thread :: (ArgPacket a b) =>
-                  b -> (a -> Prog ()) -> ThreadCode a
+declare_thread :: (ArgPacket exps tys) =>
+                  tys -> (exps -> Prog ()) ->
+                  ThreadCode exps
 declare_thread decls f = Thr (decls', desugar prog)
   where prog = f (makeVars (map fst decls'))
         arg_names = map (\n -> "arg_" ++ show n) [0..]

### 2012-03-07 paper/paper.tex
Update arg packet stuff in paper.

@@
 \begin{code}
-class ArgPacket a b | a -> b, b -> a where
-  toIExprList :: a -> [IExpr]
-  makeVars :: [String] -> a
-class ArgDecls a b | a -> b, b -> a where
-  toITypeList :: a -> [IType]
+class ArgPacket exps tys | exps -> tys,
+                           tys -> exps where
+  toIExprList :: exps -> [IExpr]
+  makeVars :: [String] -> exps
+  toITypeList :: tys -> [IType]
 
-instance ArgPacket (Expr a) [a] where
-  toIExprList (E x) = [x]
-  makeVars [x] = E (Var x)
-instance ArgPacket (Expr a, Expr b) (a, b) where
-  toIExprList (E x1, E x2) = [x1, x2]
-  makeVars [x1, x2] = (E (Var x1), E (Var x2))
--- ...
-
-instance ArgDecls (Type a) [a] where
-  toITypeList t1 = [toIType t1]
-instance ArgDecls (Type a, Type b) (a, b) where
+instance ArgPacket (Expr a) (Type a) where
+  toIExprList (E e) = [e]
+  makeVars [x] = E $ Var x
+  toITypeList t = [toIType t]
+instance ArgPacket (Expr a, Expr b) (Type a, Type b) where
+  toIExprList (E e1, E e2) = [e1, e2]
+  makeVars [x1, x2] = (E $ Var x1, E $ Var x2)
   toITypeList (t1, t2) = [toIType t1, toIType t2]
@@
 
-declare_thread :: (ArgPacket a c, ArgDecls b c) =>
-                  b -> (a -> Prog ()) ->
-                  ThreadCode c
+declare_thread :: (ArgPacket exps tys) =>
+                  tys -> (exps -> Prog ()) ->
+                  ThreadCode exps
 declare_thread = ...
-spawn :: ArgPacket a b => ThreadCode b -> a ->
+spawn :: ArgPacket exps tys =>
+         ThreadCode exps -> exps ->
          Prog ()
 spawn = ...
-call :: ArgPacket a b => Prim -> Type c -> a ->
+call :: ArgPacket exps tys =>
+        Prim -> Type c -> exps ->
         Prog (Expr c)

### 2012-03-07 paper/paper.tex
Finish writing up the types section?

@@
 programs. Providing a type system in the front end allows us to
-guarentee that the C we output does not contain type errors, vastly
+guarantee that the C we output does not contain type errors, vastly
 simplifying development. We also take advantage of this to provide
@@
 typed expressions are represented with the \t{Expr} data type
-constructor, which is a trival wrapper around \t{IExpr} parameterized
+constructor, which is a trivial wrapper around \t{IExpr} parameterized
 over a phantom type. Operations on typed expressions have type
@@
 aesthetic reasons: we felt that type annotations in He-Man code were
-more clearly readable if they were captialized, which required that
+more clearly readable if they were capitalized, which required that
 they be datatype constructors.
@@
 (.<) :: IntE -> IntE -> BoolE
-(.<) = typ2 $ RelnOp Lt
+(.<) = typ2 (RelnOp Lt)
 (.==) :: Expr a -> Expr a -> BoolE
-(.==) = typ2 $ RelnOp Eq
+(.==) = typ2 (RelnOp Eq)
 var :: String -> Type a -> Expr a -> Prog (Expr a)
@@
 Type checking foreign function calls and thread declaration and
-spawning is somewhat more interesting.
+spawning is somewhat more interesting.  For foreign function calls,
+our main concern was being able to write a single wrapper for call
+that handles a varying number of typed arguments. We want more than
+this, however: we want to ensure that both the code for a declared
+thread uses its arguments properly at the declared types code and that
+spawning threads provides arguments of the correct types.
 % and maybe a bit novel? I dunno.
-For foreign function calls, our main concern was being able to write a
-single wrapper for call that handles a varying number of typed
-arguments. To this end, we created an \t{ArgPacket} type class that
-requires the ability to convert a packet into a list of \t{IExpr} and
-to construct an \t{ArgPacket} from a list of variable names. Instances
-of \t{ArgPacket} are provided for packets of $0$ to $5$ \t{Expr}
-arguments. Now we can parameterize a \t{Call} wrapper over an
-\t{ArgPacket} to provide a \t{Expr} based interface.
 
-For thread declarations, we can produce a similar \t{ArgDecl} class
-and associated instances that can convert a packet of \t{Type}s into a
-list of \t{IType}s.
+A fragment of the code implementing typing for these features is shown
+in Figure \ref{fig:arg-packets}. The key is the multi-parameter type
+class \t{ArgPacket}. \t{ArgPacket} is parameterized over related
+tuples \footnote{Technically speaking, the unary \t{ArgPacket}s are
+  not actually tuples.} of expressions and types. The idea is that
+there exists an instance \t{ArgPacket exps tys} if and only if
+\t{exps} and \t{tys} are tuples of \t{Expr}s and \t{Type}s,
+respectively, the tuples have the same number of elements, and for
+each pair of corresponding elements in the two tuples, the \t{Type}
+represents the type of the \t{Expr} (that is, they have the same
+phantom type argument). To enforce this one-to-one correspondence and
+to allow the Haskell inferencer to use the information it provides, we
+use functional dependencies % XXX: cite, or something?
+to declare that both \t{exps} and \t{tys} are uniquely determined by
+the other.
 
-We want more than this, however: we want to ensure that both the code
-for a declared thread uses its arguments properly at the declared
-types code and that spawning threads provides arguments of the correct
-types. To do this, we need some way to extract all of the phantom
-types of the \t{Expr}s and \t{Type}s in \t{ArgPacket}s and
-\t{ArgDecl}s
+Then, by parameterizing the wrapper functions for foreign function
+calls and thread declaration and spawning over \t{ArgPacket exps tys},
+we get the desired typing guarantees.
+% XXX: is this explained fully?
 
@@
   toIExprList (E e) = [e]
-  makeVars [x] = E $ Var x
+  makeVars [x] = E (Var x)
   toITypeList t = [toIType t]
@@
   toIExprList (E e1, E e2) = [e1, e2]
-  makeVars [x1, x2] = (E $ Var x1, E $ Var x2)
+  makeVars [x1, x2] = (E (Var x1), E (Var x2))
   toITypeList (t1, t2) = [toIType t1, toIType t2]

### 2012-03-08 paper/paper.tex
Minor fixes

@@
 \def\a#1{& $\mid$ & #1 \cr}
+
+% XXX: We need to cite ElliottEtAl somewhere!
 
@@
 us, Figure \ref{fig:generic} shows our library routines for simulating
-blocking network operations. \t{do\_nb\_action} takes an event
-\t{Expr} and a \t{Prog Expr} action representing a network call that
+blocking network operations. \t{do\_nb\_action} takes an
+\t{Expr} and a \t{Prog IntE} action representing a network call that
 can return \t{EAGAIN} and produces an action that generates code to
@@
 child_code = declare_thread (Int) $
-  \[child_fd] -> do
+  \child_fd -> do
   ev <- setup_connection child_fd

### 2012-03-08 paper/paper.tex
Handle some XXXs

@@
 \end{enumerate}
-
-%% XXX: Should we put this at the end?
-\section{Related work}
-
-The C10K problem is particularly evident in the context of HTTP servers, which
-routinely maintain many concurrent connections. As we discuss in Section
-\ref{sec:evaluation}, Apache does not scale particularly well. The popular
-Lighttpd and Nginx HTTP server projects were both started with the explicit goal
-of outperforming Apache with respect to the C10K problem \cite{Lighttpd,Nginx},
-and both use event-based architectures similar to He-Man.
-
-A number of functional languages support lightweight threads with user-space
-scheduling, most notably Erlang, which was designed for highly concurrent
-telecommunications applications, and easily scales to 20,000 clients
-\cite{Hellstrom}. The Haskell runtime supports a variety of concurrency
-primitives which permit a lightweight threaded programming model \cite{LiEtAl}.
-
-%A number of custom network stack implementations exist for functional
-%languages, such as FoxNet for Standard ML \cite{BiagioniEtAl}, and
-
-He-Man is most similar to Li and Zdancewic's Haskell implementation of a hybrid
-event/thread programming model for network applications \cite{LiZdancewic}.
-Their project has similar goals to ours, but is implemented entirely within
-Haskell. As a result, unlike He-Man, their framework requires entire
-applications to be written in Haskell, and incurs runtime overhead from the
-Haskell system.
 
@@
 
-% XXX: update
 The generated code is substantially less tight than the C version,
-although it is unlikely to be any slower in practice. It has 15 states
-and is 319 lines of C. This could be reduced substantially by allowing
-conditionals outside of block tail position.
+although it is unlikely to be any slower in practice. It has 10 states
+and is 240 lines of C.
+%This could be reduced substantially by allowing
+%conditionals outside of block tail position.
 
@@
 next state to enter, a pointer to the most recently received event,
-and list heads and tails for lists of the events being waited on,
-events owned by the thread, and buffers owned by the thread (the
-latter two are so that we can automatically reclaim these resources
-when the thread exits). On a 64-bit system, the total overhead comes
-out to 88 bytes, and could easily be reduced further (e.g., by using
-singly-linked lists).  For comparison, the default thread stack size
-under NPTL on Linux is 2 megabytes.
-%% XXX update with new info?
+and list heads and tails for lists of events owned by the thread and
+buffers owned by the thread (the latter two are so that we can
+automatically reclaim these resources when the thread exits). On a
+64-bit system, the total overhead comes out to 64 bytes, and could
+easily be reduced further (e.g., by using singly-linked lists).  For
+comparison, the default thread stack size under NPTL on Linux is 2
+megabytes.
 
 % see how many threads we can spawn with Memory.hs
+
+\section{Related work}
+
+The C10K problem is particularly evident in the context of HTTP
+servers, which routinely maintain many concurrent connections. As we
+discussed in Section \ref{sec:evaluation}, Apache does not scale
+particularly well. The popular Lighttpd and Nginx HTTP server projects
+were both started with the explicit goal of outperforming Apache with
+respect to the C10K problem \cite{Lighttpd,Nginx}, and both use
+event-based architectures similar to He-Man.
+
+A number of functional languages support lightweight threads with
+user-space scheduling, most notably Erlang, which was designed for
+highly concurrent telecommunications applications, and easily scales
+to 20,000 clients \cite{Hellstrom}. The Haskell runtime supports a
+variety of concurrency primitives which permit a lightweight threaded
+programming model \cite{LiEtAl}.
+
+%A number of custom network stack implementations exist for functional
+%languages, such as FoxNet for Standard ML \cite{BiagioniEtAl}, and
+
+He-Man is most similar to Li and Zdancewic's Haskell implementation of
+a hybrid event/thread programming model for network applications
+\cite{LiZdancewic}.  Their project has similar goals to ours, but is
+implemented entirely within Haskell. As a result, unlike He-Man, their
+framework requires entire applications to be written in Haskell, and
+incurs runtime overhead from the Haskell system.
 

### 2012-03-08 paper/paper.tex
Rewrite the event loop model section.

@@
 
-Networking applications require a very constrained form of
-concurrency. Because I/O (both network and disk) latency dominates the
-computation time, each client's thread of execution spends most of its
-time waiting on I/O. In particular, we would like to always yield to a
-client no longer blocked on I/O; we need not yield to blocked clients,
-nor do we (usually) need arbitrary interleaving of execution.
+Network applications require a very constrained form of
+concurrency. Because in most network applications, I/O (both network
+and disk) latency dominates the computation time, each client's thread
+of execution spends most of its time waiting on I/O. In particular, we
+would like to always yield to a client no longer blocked on I/O; we
+need not yield to blocked clients, nor do we (usually) need arbitrary
+interleaving of execution.
 
-%% XXX: do we really want to talk about this?
-As of version 2.5.44, Linux has supported \t{epoll} for non-blocking
-multiplexing I/O. A server can register any number of file descriptors with
-\t{epoll\_ctl}, and a call to \t{epoll\_wait} returns the ready descriptors.
+In the event loop model, network I/O is done using non-blocking
+\t{read} and \t{write} calls and all of the blocking for I/O is
+centralized in a single call to a multiplexing system call such as
+\t{select} or \t{epoll}. The multiplexing call will block until there
+is some I/O ready to be performed and indicate to the application
+which file descriptors are ready. From this, the application then
+needs to decide how handle these events. This requires, for each
+connection, having some sort of state machine to track what state the
+connection is in and what needs to happen next, as well as other local
+state.
 
-In the event loop model, all blocking I/O must occur in a single loop around
-\t{epoll\_wait}. Each thread is stored as a \t{struct} containing thread-local
-state, including the next instruction pointer, and each registered descriptor
-has a pointer to the thread waiting on it. Instead of performing an I/O call, a
-thread simply registers a new descriptor and returns back to the main loop;
-whenever a descriptor is ready, the main loop jumps to the instruction pointer
-of the thread waiting on it. Spawning a new thread requires only allocating a
-new thread \t{struct}, and registering it in the main loop.
+One common scheme for doing this involves registering a function
+pointer with the event loop to be called on an event. Typically each
+connection is also associated with a structure of connection local
+data. Instead of making a blocking system call, then, an application
+registers an event with the event loop along with the function to call
+when it occurs, and then returns back to the loop.
 
-Each thread must be split into one function for each jump target, and
-thread-local state must be accessed via the thread \t{struct} shared between
-these functions. Furthermore, because I/O only occurs in the main loop, these
-functions must return immediately after registering new descriptors. As a
-result, each thread is split up into quite a few functions in this paradigm.
-Because this translation from threaded to event-based is quite mechanical, we
-are able to perform it automatically in the He-Man back-end.
+The advantage of this scheme is that it is highly scalable and can be
+efficient: handling a new connection typically just involves
+allocating a new structure and registering it with the event loop. The
+disadvantage is that now network facing code needs to be scattered
+through many small functions, with all access to connection state
+mediated through a shared structure.
+
+The event loop model is then, essentially, built around closure
+converted trampolining continuation passing style. These are well
+studied problems and can easily be done mechanically in the He-Man
+backend.
 

### 2012-03-09 src/rt/mainloop.c
Disable the async io stuff.

@@ -12,8 +12,10 @@
 #include <stdbool.h>
 #include <signal.h>
 
+#if ENABLE_AIO
 #include <libaio.h>
 #include <sys/eventfd.h>
+#endif
 
 #include <pthread.h>
 
@@ -31,9 +33,11 @@ int next_tid = 0;
 
 static struct {
 	int epoll_fd;
+#if ENABLE_AIO
 	io_context_t aio_ctx;
 	int aio_eventfd;
 	struct event_t *aio_dummy_event;
+#endif
 #if MT_RUNTIME
 	pthread_mutex_t sched_lock;
 #endif
@@ -238,7 +242,7 @@ msg_data_t channel_recv(channel_t *ch)
 	return ret;
 }
 
-
+#if ENABLE_AIO
 static void handle_aio_event(struct io_event *event)
 {
 	abort();
@@ -248,6 +252,7 @@ static void handle_aio_event(struct io_event *event)
 	handle_event(ev);
 	*/
 }
+#endif
 static void handle_epoll_event(struct epoll_event *event)
 {
 	event_t *ev = event->data.ptr;
@@ -258,8 +263,9 @@ static void handle_epoll_event(struct epoll_event *event)
 // and no aio events are expected, epoll will block.
 static void do_poll(bool can_sleep)
 {
-	struct io_event aio_events[MAX_EVENTS];
 	struct epoll_event epoll_events[MAX_EVENTS];
+#if ENABLE_AIO
+	struct io_event aio_events[MAX_EVENTS];
 
 	// We need to be careful to not let either aio or epoll events
 	// starve the other. We do this by always doing nonblocking polls
@@ -274,6 +280,9 @@ static void do_poll(bool can_sleep)
 	for (int i = 0; i < aio_cnt; i++) {
 		handle_aio_event(&aio_events[i]);
 	}
+#else
+#define expect_aio 1
+#endif
 
 	// If we aren't expecting aio, block indefinitely, otherwise
 	// just poll.
@@ -281,12 +290,15 @@ static void do_poll(bool can_sleep)
 	int epoll_cnt = epoll_wait(state.epoll_fd, epoll_events,
 	                           MAX_EVENTS, epoll_timeout);
 	for (int i = 0; i < epoll_cnt; i++) {
+#if ENABLE_AIO
 		if (epoll_events[i].data.ptr == state.aio_dummy_event) {
[…]

### 2012-03-09 src/rt/mainloop.c
Don't include libaio.h from mainloop.h.

@@ -1,5 +1,6 @@
 #include <unistd.h>
 #include <stdio.h>
+#include <string.h>
 #include <sys/types.h>
 #include <sys/stat.h>
 #include <sys/param.h>

### 2012-03-09 src/rt/mainloop.h

@@ -3,7 +3,6 @@
 
 #include <stdbool.h>
 #include <stdint.h>
-#include <libaio.h>
 #include "variable_queue.h"
 
 typedef enum event_type_t { EVENT_NB, EVENT_CHANNEL, /* EVENT_AIO */ }
@@ -77,7 +76,6 @@ typedef struct thread_t {
 	buf_queue_t bufs;
 	thread_cont *cont;
 } thread_t;
-
 
 
 void inc_refcount(void *p);

### 2012-03-09 src/rt/http_lib.c
Make the http server send a content length.

@@ -43,3 +43,13 @@ int http_parse(char *buf, int len)
 	//printf("opening file: %s\n", buf);
 	return 1;
 }
+
+int http_make_hdr(char *buf, int len, int fd)
+{
+	struct stat sbuf;
+	fstat(fd, &sbuf);
+
+	return snprintf(buf, len,
+	                "HTTP/1.1 200 OK\r\nConnection: close\r\nContent-Length: %d\r\n\r\n",
+	                sbuf.st_size);
+}

### 2012-03-09 src/rt/lib.h

@@ -38,5 +38,6 @@ int print_int(int n);
 
 // This shouldn't be here. Programs should be able to include other headers
 int http_parse(char *buf, int len);
+int http_make_hdr(char *buf, int len, int fd);
 
 #endif

### 2012-03-09 src/Http.hs
Moar sendfile.

@@ -9,13 +9,17 @@ bufsize = 4096*4
 http_parse :: BufferE -> IntE -> Prog IntE
 http_parse buf len =
   callName "parse_result" (CFn "http_parse") Int (buf, len)
-http_make_hdr :: BufferE -> IntE -> FdE -> Prog IntE
-http_make_hdr buf len fd =
-  callName "hdr_length" (CFn "http_make_hdr") Int (buf, len, fd)
+get_file_size :: FdE -> Prog IntE
+get_file_size fd =
+  callName "file_size" (CFn "get_file_size") Int fd
+http_make_hdr :: BufferE -> IntE -> IntE -> Prog IntE
+http_make_hdr buf len fsize =
+  callName "hdr_length" (CFn "http_make_hdr") Int (buf, len, fsize)
 
 setup_connection :: FdE -> Prog (FdE, EventE)
 setup_connection fd = do
   make_nb fd
+  set_sock_cork fd
   e <- mk_nb_event fd kEVENT_RDWR
   return (fd, e)
 
@@ -39,6 +43,21 @@ parse_request buf ev = do
     parse_result .=. http_parse buf request_size
   return parse_result
 
+sendfile :: FdE -> FdE -> IntE -> Prog IntE
+sendfile out_fd in_fd len =
+  callName "amt_written" (CFn "sendfile") Int (out_fd, in_fd, num 0, len)
+
+full_sendfile (out_fd, e) in_fd size = do
+  amt_written <- var "total_written" Int 0
+  failed <- var "write_failed" Int 0
+  while (amt_written .< size .&& failed .== 0) $ do
+    prepare_event e kEVENT_WR
+    amt <- sendfile out_fd in_fd (size - amt_written)
+    ifE (amt .== -1 .&& errno .== kEAGAIN) (wait e) $
+      (ifE (amt .<= 0) (failed .= 1)
+       (amt_written .= amt_written + amt))
+  return amt_written
+
 child_code = declare_thread (FD) $
   \child_fd -> do
   ev <- setup_connection child_fd
@@ -51,16 +70,13 @@ child_code = declare_thread (FD) $
   ifE' (isFailure file_fd) exit
   let cleanup = close file_fd >> exit
 
-  hdr_length <- http_make_hdr buf bufsize file_fd
+  file_size <- get_file_size file_fd
+  hdr_length <- http_make_hdr buf bufsize file_size
   amount_written <- full_write ev buf hdr_length
[…]

### 2012-03-09 src/rt/http_lib.c

@@ -44,12 +44,16 @@ int http_parse(char *buf, int len)
 	return 1;
 }
 
-int http_make_hdr(char *buf, int len, int fd)
+int get_file_size(int fd)
 {
 	struct stat sbuf;
 	fstat(fd, &sbuf);
+	return sbuf.st_size;
+}
 
+int http_make_hdr(char *buf, int len, int size)
+{
 	return snprintf(buf, len,
 	                "HTTP/1.1 200 OK\r\nConnection: close\r\nContent-Length: %d\r\n\r\n",
-	                sbuf.st_size);
+	                size);
 }

### 2012-03-09 src/rt/lib.c

@@ -8,6 +8,13 @@ int set_sock_reuse(int fd)
 	                  &value, sizeof(value));
 }
 
+int set_sock_cork(int fd)
+{
+	int value = 1;
+	return setsockopt(fd, IPPROTO_TCP, TCP_CORK,
+	                  &value, sizeof(value));
+}
+
 int make_socket_non_blocking(int sfd)
 {
 	int flags, s;

### 2012-03-09 src/rt/lib.h

@@ -17,12 +17,16 @@
 #include <sys/socket.h>
 #include <sys/types.h>
 #include <arpa/inet.h>
+#include <netinet/in.h>
+#include <netinet/tcp.h>
+#include <sys/sendfile.h>
 
 #define EVENT_RD (EPOLLIN)
 #define EVENT_WR (EPOLLOUT)
 #define EVENT_RDWR (EPOLLIN|EPOLLOUT)
 
 int set_sock_reuse(int fd);
+int set_sock_cork(int fd);
 int make_socket_non_blocking(int sfd);
 int sock_bind_v4(int fd, int addr, int port);
 int print_int(int n);
@@ -38,6 +42,7 @@ int print_int(int n);
 
 // This shouldn't be here. Programs should be able to include other headers
 int http_parse(char *buf, int len);
-int http_make_hdr(char *buf, int len, int fd);
+int http_make_hdr(char *buf, int len, int size);
+int get_file_size(int fd);
 
 #endif

### 2012-03-10 paper/paper.tex
Added citations; rewrote abstract.

@@
 \begin{abstract}
-% XXX: better abstract;
-% and where should we put the code link?
+The event-driven model is the gold standard for highly concurrent network
+application programming; while threaded code is easier to write and reason
+about, it is considerably less scalable.
 
-We present He-Man \cite{HeMan}, a compiled, embedded, domain specific language in Haskell for
-building scalable network applications in a natural, straight-line manner while
-still producing efficient event loop-based code. He-Man programs are compact and
-measurements indicate that it performs acceptably. Our implementation is
-available at
+We observe that these paradigms are not only interchangeable, but primarily
+confined to an application's core control flow; we propose a hybrid model in
+which programmers express this core logic at a high level of abstraction, yet
+are free to implement the remainder of the application at a low level.
+
+This paper describes the design and Haskell implementation of He-Man, a domain
+specific language for network application programming which presents a
+sequential thread-like abstraction, but compiles to performant event-driven C
+code. Our implementation is available at
 \url{https://github.com/msullivan/he-man}.
+%\cite{HeMan}
 \end{abstract}
@@
 \section{Introduction}
+building scalable concurrent network applications. The
+language presents a
+
+presents difficult concurrency challenges 
+implementation challenge to developers
+supporting many simultaneous clients
+although memory and bandwidth is plentiful
+difficult to design applications
+to take full advantage of these resources.
+threads inefficient (scheduling)
+event loops (hard to write)
+
+\cite{LauerNeedham},
+
+in a natural, straight-line manner while
+still producing efficient event loop-based code.
+
+He-Man programs are compact and
+measurements indicate that it performs acceptably. Our implementation is
+available at
+, embedded, domain specific language in Haskell for
+building
 
@@
 core application logic, and can write the rest of their application in C.
+
+\cite{*}
 
@@
 
+\pagebreak
 \bibliography{citations}{}

### 2012-03-10 benchmarking/graph.py
Add in partial script for making the graphs.

@@ -1,0 +1,78 @@
+#!/usr/bin/env python
+
+import sys
+from scipy.stats import stats
+import Gnuplot, Gnuplot.funcutils
+
+port_map = {"80": "Apache", "81": "lighttpd", "8080": "HeMan"}
+
+def process_file(file_name):
+    _, it, size, conns, port = file_name.split("_")
+    server = port_map[port]
+    lines = [line.strip() for line in open(file_name)]
+    finished = lines[-4]
+    parts = finished.split(" ")
+    time = "%d.%0.3d%0.3d" % (int(parts[2]), int(parts[4]), int(parts[7]))
+
+    requests = lines[-3]
+    parts = requests.split(" ")
+    successes = parts[7]
+    rate = float(successes)/float(time)
+
+    return (server, size, int(conns), it, time, rate)
+
+def map_add(m, key, x):
+    if key not in m:
+        m[key] = []
+    m[key].append(x)
+
+def organize_data(data, size):
+    data = [x[:1]+x[2:] for x in data if x[1] == size]
+    servers = {x[0]: [] for x in data}
+    for x in data:
+        servers[x[0]].append(x[1:])
+    for v in servers.values():
+        v.sort()
+    return servers
+
+def produce_stats(data):
+    m = {}
+    for (server, size, conns, it, time, rate) in data:
+        map_add(m, (server, size, conns), rate)
+    data = []
+    for k, v in m.items():
+        mean = stats.nanmean(v)
+        stddev = stats.nanstd(v)
+        data += [k + (mean, stddev)]
+    return data
+
+def graph(size, table):
+
+    g = Gnuplot.Gnuplot(debug=1)
+    g.title(size)
+    g('set style data errorlines')
+    g('set yrange [0:]')
+    g.xlabel('# of Concurrent requests')
+    g.ylabel('Requests/second')
+
+    lines = []
+    for server, data in table.items():
+        xs, ys, stds = zip(*data)
+        lines.append(Gnuplot.Data(xs, ys, stds, title=server))
+    g.replot(*lines)
+
+    raw_input('Please press return to continue...\n')
+
+
+def main(args):
+    data = [process_file(f) for f in args[1:]]
+    data = produce_stats(data)
+
+    sizes = set(x[1] for x in data)
+
+    tables = {size: organize_data(data, size) for size in sizes}
+    size = '16K'
+    graph(size, tables[size])
+
+if __name__ == '__main__':
+    sys.exit(main(sys.argv))

### 2012-03-10 benchmarking/graph.py
Graph production.

@@ -47,11 +47,11 @@ def produce_stats(data):
     return data
 
 def graph(size, table):
-
-    g = Gnuplot.Gnuplot(debug=1)
-    g.title(size)
+    g = Gnuplot.Gnuplot(debug=0)
+    #g.title(size)
     g('set style data errorlines')
     g('set yrange [0:]')
+    g('set xrange [0:525]') # XXX: change if we change max conns
     g.xlabel('# of Concurrent requests')
     g.ylabel('Requests/second')
 
@@ -59,9 +59,9 @@ def graph(size, table):
     for server, data in table.items():
         xs, ys, stds = zip(*data)
         lines.append(Gnuplot.Data(xs, ys, stds, title=server))
-    g.replot(*lines)
+    g.itemlist = lines
 
-    raw_input('Please press return to continue...\n')
+    g.hardcopy('graph_' + size + '.ps', enhanced=1, color=1)
 
 
 def main(args):
@@ -71,8 +71,8 @@ def main(args):
     sizes = set(x[1] for x in data)
 
     tables = {size: organize_data(data, size) for size in sizes}
-    size = '16K'
-    graph(size, tables[size])
+    for size, table in tables.items():
+        graph(size, table)
 
 if __name__ == '__main__':
     sys.exit(main(sys.argv))

### 2012-03-11 src/rt/http_lib.c
Send more header information.

@@ -54,6 +54,6 @@ int get_file_size(int fd)
 int http_make_hdr(char *buf, int len, int size)
 {
 	return snprintf(buf, len,
-	                "HTTP/1.1 200 OK\r\nConnection: close\r\nContent-Length: %d\r\n\r\n",
+	                "HTTP/1.1 200 OK\r\nConnection: close\r\nServer: Masters of the Universe\r\nContent-Length: %d\r\nX-Diqs: 8=============================================================================================================================================================================================D\r\n\r\n",
 	                size);
 }

### 2012-03-11 paper/paper.tex
Graph production.
Wrote abstract and most of introduction; added citations.

@@
 % XXX: We need to cite ElliottEtAl somewhere!
+% XXX: standardize on IO vs I/O
+% XXX: standardize on domain-specific vs domain specific
 
 %XXX: new title?
-\title{He-Man: A domain-specific language for lightweight concurrency}
+\title{He-Man: A domain specific language for lightweight concurrency}
 %\subtitle{I said hey, what's goin' on}
@@
 application programming; while threaded code is easier to write and reason
-about, it is considerably less scalable.
-
-We observe that these paradigms are not only interchangeable, but primarily
-confined to an application's core control flow; we propose a hybrid model in
-which programmers express this core logic at a high level of abstraction, yet
-are free to implement the remainder of the application at a low level.
+about, it is considerably less scalable. We observe that these paradigms are not
+only interchangeable, but primarily confined to an application's core control
+flow; we propose a hybrid model in which programmers express this core logic at
+a high level of abstraction, yet are free to implement the remainder of the
+application at a low level.
 
@@
 code. Our implementation is available at
-\url{https://github.com/msullivan/he-man}.
-%\cite{HeMan}
+\url{https://github.com/msullivan/he-man}. %\cite{HeMan}
 \end{abstract}
 
+% Related work:
+%\cite{VanderwaartCrary} re bounded time
+%\cite{vonBehrenEtAl} re Capriccio
+%\cite{vonBehrenConditBrewer} re cooperative multithreading
+%\cite{LiZdancewic}
+%\cite{LauerNeedham} also intro
+%\cite{MarlowEtAl} also intro
+
+% Functional programming:
+%\cite{Vinoski}
+%\cite{Hellstrom}
+%\cite{BiagioniEtAl}
+%\cite{LiEtAl}
+%\cite{GraunkeEtAl}
+%\cite{Queinnec}
+
+% Haskell EDSLs:
+%\cite{Gill}
+%\cite{ElliottEtAl}
+%\cite{LeijenMeijer}
+
 \section{Introduction}
-building scalable concurrent network applications. The
-language presents a
 
-presents difficult concurrency challenges 
-implementation challenge to developers
-supporting many simultaneous clients
-although memory and bandwidth is plentiful
-difficult to design applications
-to take full advantage of these resources.
-threads inefficient (scheduling)
-event loops (hard to write)
+Modern network servers, like HTTP, IRC, and multiplayer games, need to support
+many simultaneous, possibly idle, connections---this is known as the C10K
+(``10,000 client'') problem \cite{Kegel}. Because I/O latency is high,
+particularly over a network, our goal is to process ready clients while others
+are blocked on I/O. To achieve maximum throughput, such servers are usually
+written in low-level languages like C.
 
-\cite{LauerNeedham},
+The simplest way to achieve this concurrency is to use kernel threads, like the
+industry-standard Apache web server \cite{Apache}. These incur a fair amount of
+overhead, and do not allow control over scheduling; they waste time, for
+example, yielding to I/O blocked clients.
 
-in a natural, straight-line manner while
-still producing efficient event loop-based code.
+More performant web servers, such as Nginx \cite{Nginx} and Lighttpd
+\cite{Lighttpd}, use non-blocking I/O routines, multiplexed at a single call
+site (the ``event loop''). This allows one thread to serve many clients,
+handling each only when its I/O has completed. This event-driven model requires
+the routines to explicitly yield to the event loop, forcing programmers to
+restructure their code as a state machine, save state across function
+boundaries, and implement a scheduler.
 
-He-Man programs are compact and
-measurements indicate that it performs acceptably. Our implementation is
-available at
-, embedded, domain specific language in Haskell for
-building
+Although the event-driven model is typically chosen by developers who wish to
+maximize performance (as with Nginx and Lighttpd), von Behren, Condit, and
+Brewer \cite{vonBehrenConditBrewer} argue that events are an unnatural
+abstraction for this problem domain, and that an appropriate threading library
+(in particular, without preemption) can achieve equivalent performance.
 
-Network applications need to support many simultaneous clients---this is known
-as the C10K (``10,000 client'') problem \cite{Kegel}. Modern servers have enough
-memory and bandwidth for massive concurrency, but it is difficult to design
-applications to take full advantage of these resources.
+In an oft-cited 1979 paper, Lauer and Needham \cite{LauerNeedham} argue that the
+threaded and event-driven models are dual, where the thread scheduler plays the
+role of an event loop. Then the translation from threads is simply trampolining
+continuation-passing-style, where, in this case, I/O operations must yield.
 
-A server must maintain separate state for each connection, and often waits for
-network or disk I/O to complete before continuing to serve a particular client.
-The simplest solution is to maintain one thread per client, using blocking I/O
-within each thread. This incurs frequent and expensive context switches at the
-operating system level; because the threads have identical code and little
-private data, this approach is unnecessary and does not scale.
+The idea of implementing custom schedulers for lightweight thread abstractions
+is not new, and has been used to great effect in functional language runtimes
+like GHC \cite{MarlowEtAl}. Li and Zdancewic \cite{LiZdancewic} have developed
+monadic abstractions in Haskell for threaded and event-driven concurrency, for
+the same problem domain of scalable network applications; their implementation
+draws upon the concurrency and synchronization support in the GHC runtime.
 
-An improvement is to write multithreaded code in which each thread serves
-multiple clients. This is the most popular approach, and is utilized in the
-Apache HTTP server \cite{Apache}.
+We observe that, although concurrency is critical to network applications
+because I/O dominates computation time, core I/O routines comprise a small
+portion of an application's program logic, which is otherwise devoted to
+protocol implementation, caching, etc. We might expect that our concurrency
+abstractions can be similarly confined; however, the above solutions require
+writing the entire application in Haskell, or at least writing the core in
+Haskell and making liberal use of the FFI.
 
-Languages like Erlang and Haskell provide lightweight threads via user-space
-schedulers. This affords even better performance, but remains unnecessarily
-powerful for this problem space, imposing runtime overhead on applications which
-don't require preemptable, independently scheduled threads \cite{Vinoski}.
+Our contribution is to distill the Lauer-Needham duality to its essence, by
+compiling from a simple thread abstraction to self-contained event-driven C
+code. Our language, He-Man: Haskell Event Manager, Apropos Networking, is a
+compiled domain specific language embedded in Haskell, which offers a small
+monadic language supporting thread spawning, blocking events, and external
+function calls.
 
-By far the leanest approach is to write an event loop around kernel facilities
-for multiplexed, non-blocking I/O, such as \t{select}, \t{poll}, \t{epoll}, or
-\t{kqueue}. In this model, all blocking occurs at a single call site in the
-code. The programmer is responsible for explicitly maintaining per-client state
-and invoking the appropriate client on each I/O event.
+%XXX
 
-This requires the entire application to be restructured as a state machine with
-explicit per-client control blocks, a model which is very difficult for
-programmers to reason about. There exist numerous C/C++ libraries for event
-loops, but none of these avoid the state machine paradigm.
+The He-Man system consists of a compiler written in Haskell, and a lightweight
+runtime in C, which links with the compiler's output. Both are small and easily
+extensible. We target C because it is commonly used for network applications,
 
-Our observation is that the event loop model is simply trampolining
-continuation-passing-style, % TODO citation
-which can be obtained by mechanical translation from the straight-line threaded
-code. This suggests it is possible to obtain the best of both worlds---a
-threaded programming model with the performance of event loops.
+so it is trivial to link our
 
-To that end, we present He-Man: Haskell Event Manager, Apropos Networking, an
-embedded domain-specific language (EDSL) hosted in Haskell, which provides a
-threaded abstraction for network application programming and compiles to a C
-event loop based around Linux's \t{epoll}.
-
-He-Man is a straightforward monadic language with built-in network primitives,
-offering a simple thread-per-client programming model. It is easily extensible
-in Haskell, and also allows calls to foreign C functions; it outputs C code
-which links with an event loop runtime and any code a programmer might prefer to
-write directly in C. Our vision is that C programmers can use He-Man for their
+any code a programmer might prefer to write directly in C. 
+Our vision is that C programmers can use He-Man for their
 core application logic, and can write the rest of their application in C.
 
-\cite{*}
-
-% Contributions; probably should be elsewhere
-We identify our contributions as:
-\begin{enumerate}
-
-\item We present a case study of compiled embedded domain specific
-  language design and implementation.
-\end{enumerate}
+He-Man programs are compact and measurements indicate that it performs
+acceptably.
 
 \section{The event loop model}
-%% XXX: it isn't really clear what's going on in this section; it
-%% doesn't /really/ talk about event loop stuff in general. It is
-%% talking about one possible way to write event based things (the way
-%% He-Man generates).
-
 Network applications require a very constrained form of
@@
 backend.
+
+\cite{*}
 
@@
 
+% XXX taken from old intro
+
+Languages like Erlang and Haskell provide lightweight threads via user-space
+schedulers. This affords even better performance, but remains unnecessarily
+powerful for this problem space, imposing runtime overhead on applications which
+don't require preemptable, independently scheduled threads \cite{Vinoski}.
+
+There exist C/C++ libraries for event loops\cite{libevent}, but none of these
+avoid the state machine paradigm.
+
 %% XXX: needs serious work
@@
 
-\pagebreak
+% XXX
+
+We identify our contributions as:
+* We present the idea of generating code for network facing logic from a
+description written in a DSL while building the rest of the application in
+C. In a way, the DSL could be considered as a method for programmatically
+generating event driven frameworks customized for the application.
+* We observe that converting a traditional block-structured imperative
+program to a form amenable for use in an event-driven system is analogous
+to CPS conversion and thus easily done automatically. (Is this actually a
+contribution, or is it obvious and been said a million times before?)
+* We present a case study of compiled embedded domain specific language
+design and implementation
+* We demonstrate that it is practical to build efficient network
+applications using our system
+
 \bibliography{citations}{}

### 2012-03-11 paper/paper.tex
Finished introduction.

@@
 
-% Related work:
-%\cite{VanderwaartCrary} re bounded time
-%\cite{vonBehrenEtAl} re Capriccio
-%\cite{vonBehrenConditBrewer} re cooperative multithreading
-%\cite{LiZdancewic}
-%\cite{LauerNeedham} also intro
-%\cite{MarlowEtAl} also intro
-
-% Functional programming:
-%\cite{Vinoski}
-%\cite{Hellstrom}
-%\cite{BiagioniEtAl}
-%\cite{LiEtAl}
-%\cite{GraunkeEtAl}
-%\cite{Queinnec}
-
-% Haskell EDSLs:
-%\cite{Gill}
-%\cite{ElliottEtAl}
-%\cite{LeijenMeijer}
-
 \section{Introduction}
@@
 compiling from a simple thread abstraction to self-contained event-driven C
-code. Our language, He-Man: Haskell Event Manager, Apropos Networking, is a
+code. This abstraction, He-Man: Haskell Event Manager, Apropos Networking, is a
 compiled domain specific language embedded in Haskell, which offers a small
@@
 
-%XXX
+The He-Man system consists of a compiler written in Haskell, and a lightweight
+runtime in C which links with the compiled code. He-Man allows programmers to
+develop scalable network applications by expressing their I/O logic in He-Man,
+and linking the resulting event loop to the rest of their application, which can
+be written in C or any language which interfaces with C. We target C because it
+is commonly used in this domain, and well-suited to performance-critical
+applications.
 
-The He-Man system consists of a compiler written in Haskell, and a lightweight
-runtime in C, which links with the compiler's output. Both are small and easily
-extensible. We target C because it is commonly used for network applications,
+In the remainder of the paper, we discuss the event-driven programming model,
+our implementation of the He-Man system, and verify that its performance is
+competitive with handwritten event loops.
 
-so it is trivial to link our
+\section{The event-driven model}
 
-any code a programmer might prefer to write directly in C. 
-Our vision is that C programmers can use He-Man for their
-core application logic, and can write the rest of their application in C.
-
-He-Man programs are compact and measurements indicate that it performs
-acceptably.
-
-\section{The event loop model}
 Network applications require a very constrained form of
-concurrency. Because in most network applications, I/O (both network
-and disk) latency dominates the computation time, each client's thread
+concurrency. In most network applications, I/O (both network
+and disk) latency dominates computation time, so each client's thread
 of execution spends most of its time waiting on I/O. In particular, we
@@
 centralized in a single call to a multiplexing system call such as
-\t{select} or \t{epoll}. The multiplexing call will block until there
-is some I/O ready to be performed and indicate to the application
-which file descriptors are ready. From this, the application then
-needs to decide how handle these events. This requires, for each
-connection, having some sort of state machine to track what state the
-connection is in and what needs to happen next, as well as other local
-state.
+\t{select} or \t{epoll}. The multiplexing call will block until some 
+I/O is ready to be performed, and indicates to the application
+which file descriptors are ready. To decide how to handle these 
+events, an application must, for each connection, have some sort of 
+state machine to track what state the connection is in and what 
+needs to happen next, as well as other local state.
 
@@
 
-The event loop model is then, essentially, built around closure
+The event loop model is then, essentially, built around closure-
 converted trampolining continuation passing style. These are well
@@
 backend.
-
-\cite{*}
 
@@
 \centering
-\includegraphics[width=0.9\textwidth]{128kb_graph.png}
+\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
 \caption{Web server benchmark, 128 KB file.}
@@
 \centering
-\includegraphics[width=0.9\textwidth]{16kb_graph.png}
+\includegraphics[width=0.9\textwidth]{graph_16K.pdf}
 \caption{Web server benchmark, 16 KB file.}
 \label{fig:graph16}
+\end{figure*}
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=0.9\textwidth]{graph_100.pdf}
+\caption{Web server benchmark, 100 B file.}
+\label{fig:graph100}
 \end{figure*}
@@
 % see how many threads we can spawn with Memory.hs
+
+% Related work:
+%\cite{VanderwaartCrary} re bounded time
+%\cite{vonBehrenEtAl} re Capriccio
+%\cite{vonBehrenConditBrewer} re cooperative multithreading
+%\cite{LiZdancewic}
+%\cite{LauerNeedham} also intro
+%\cite{MarlowEtAl} also intro
+
+% Functional programming:
+%\cite{Vinoski}
+%\cite{Hellstrom}
+%\cite{BiagioniEtAl}
+%\cite{LiEtAl}
+%\cite{GraunkeEtAl}
+%\cite{Queinnec}
+
+% Haskell EDSLs:
+%\cite{Gill}
+%\cite{ElliottEtAl}
+%\cite{LeijenMeijer}
 
@@
 
+\cite{*}
+
 \bibliography{citations}{}

### 2012-03-11 paper/paper.tex
Write a new HTTP server section.

@@
 and to demonstrate separating network logic and application logic, we
-built a simple (and bad) HTTP 1.0 server. The network logic is written
-in He-Man and request parsing is written in C. The C side is
-implemented as a function that takes a buffer and returns whether it
-forms a complete HTTP request and, if so, fills a buffer with the
-file name to serve. The entire server is 54 lines of He-Man Haskell and
-27 lines of C. It suffers from a number of major deficiencies, but is
-complete enough to benchmark.
+built a very simple HTTP server. The network logic is written in
+He-Man and request parsing and response header building is written in
+C. The C request parsing is implemented as a function that takes a
+buffer and returns whether it forms a complete HTTP request and, if
+so, fills a buffer with the file name to serve. The entire server is
+68 lines of He-Man Haskell and 39 lines of C. It suffers from a number
+of major deficiencies, but is complete enough to benchmark. To make
+sure that it does not get away with sending less data than the servers
+it is being compared against, it sends some ``padding'' headers with
+useless data.
 
-Our benchmark measurements are taken using ab, the Apache HTTP server
-benchmarking tool \cite{ApacheAB}. Measurements were collected by
-having a variable number of client threads request a file 100 times;
-the experiment was repeated with a 16 kilobyte file and a 128 kilobyte
-file. We ran the tests against both our He-Man HTTP server and Apache
-2.2.20, installed with the default Ubuntu 11.10 settings, except for
-client, thread, and process limits being adjusted upwards. Since the
-He-Man runtime is currently only single-threaded, we ran the Apache
-web server pinned to a single processor. Since the test repeatedly
-requests the same file, it will always be in the kernel's disk cache,
-and thus this test is not disk bound.
+Our benchmark measurements are taken using weighttp, an http
+benchmarking tool associated with lighttpd
+\cite{weighttp}. Measurements were collected by requesting a file many
+times with a variable number of concurrent connections; the experiment
+was repeated with a 100 byte file (80000 requests), a 16 kilobyte file
+(30000 requests), and a 128 kilobyte file (8000 requests).
+weighttp was run with 4 threads making requests.
+Since the test repeatedly requests the same file, it should always be in
+the kernel's disk cache.
 
-For our experiments, the server ran Linux kernel 3.0.0 on a Lenovo
-Thinkpad W520 with a quad core 2.3 GHz Intel Core i7-2820QM and 8
-gigabytes of RAM, and the client ran on a Lenovo Thinkpad T500 with a
-Intel Core 2 Duo and 8 gigabytes of RAM.  % XXX: client specs
-The machines were connected by a gigabit Ethernet link.
-Unfortunately, ab is not particularly scalable, so we had trouble
-testing with more than 500 clients.
+For our experiments, both the client and server ran Linux kernel
+2.6.32 on quad core 3.1 GHz Intel Core i5-2400 processors with 8
+gigabytes of RAM. The machines were connected over a 100 megabit
+Ethernet network.
+We ran the tests against our He-Man HTTP server, Apache 2.2.14 (installed
+with the default Ubuntu 10.04 LTS settings, except for client, thread,
+and process limits being adjusted upwards) and Lighttpd 1.4.26
+(installed with the default Ubuntu 10.04 LTS settings).
 
-Figure \ref{fig:graph128} compares our server to Apache for a 128 KB
-file. Our server performs nearly identically to Apache on this load.
-Figure \ref{fig:graph16} compares our server to Apache for a 16 KB
-file. In this scenario, we substantially outperform Apache. We
-conjecture this is because with a smaller file size, overhead for
-connection setup begins to dominate, and connection setup in He-Man
-consists of a few memory allocations whereas connection setup under
-Apache may require spawning new OS level threads.
+Each trial was run 10 times; the plots show the arithmetic mean of the
+number of requests per second the servers handled; error bars show the
+standard deviation.
+
+Figure \ref{fig:graph100} compares the three servers while serving a
+very small (100 byte) file. Measurements with small files test the
+overhead of the web server more clearly than measurements with large
+files, which wind up testing how efficiently the kernel can transfer
+data. In this test, He-Man outperforms Apache and lighttpd across all
+of our measurements. This can be attributed to factors both in He-Man
+and in our server. He-Man imposes very low overhead and very low
+connection creation cost, which makes it easy to quickly handle many
+new connections. Our server is also much less featureful that Apache
+and lighttpd, which means it needs to do less. It is difficult to
+separate out these factors.
+
+Figure \ref{fig:graph128} and Figure \ref{fig:graph16} compares our
+server to Apache and lighttpd for 128 KB and 16 KB files. On these
+loads, our server performs nearly identically to Apache while being
+outperformed by the highly-turned lighttpd. In these tests, both
+Apache and He-Man begin to handle fewer requests per second above some
+number of concurrent requests (about 400 for the 16 KB file and 300
+for the 128 KB file). We investigated the He-Man httpd server's
+behavior in these tests, and found that the measured slowdown in
+requests/second occured when, after rapidly processing most of the
+requests in a test batch, a few requests would spend a very long time
+waiting to connect, causing a pause at the end of the run. During
+these pauses, both the client and the server were blocked in the
+kernel, waiting for network events. This strongly suggests to us that
+the problem is not with the He-Man compilation phase but that our web
+server would benefit from more careful tuning of network interactions
+(which may require improvements to the runtime).
 
@@
 \centering
-\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
-\caption{Web server benchmark, 128 KB file.}
-\label{fig:graph128}
+\includegraphics[width=0.9\textwidth]{graph_100.pdf}
+\caption{Web server benchmark, 100 byte file.}
+\label{fig:graph100}
 \end{figure*}
@@
 \centering
-\includegraphics[width=0.9\textwidth]{graph_100.pdf}
-\caption{Web server benchmark, 100 B file.}
-\label{fig:graph100}
+\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
+\caption{Web server benchmark, 128 KB file.}
+\label{fig:graph128}
 \end{figure*}

### 2012-03-11 src/Language/HeMan/Lib.hs

@@ -52,6 +52,11 @@ sock_write :: FdE -> BufferE -> IntE -> Prog IntE
 sock_write fd buf len =
   callName "amt_written" (CFn "write") Int (fd, buf, len)
 
+
+sendfile :: FdE -> FdE -> IntE -> Prog IntE
+sendfile out_fd in_fd len =
+  callName "amt_written" (CFn "sendfile") Int (out_fd, in_fd, num 0, len)
+
 close :: FdE -> Prog IntE
 close fd =
   call (CFn "close") Int (fd)

### 2012-03-11 paper/paper.tex
Add note about HTTP server.

@@
 long and is a very clear representation of the control flow of each
-individual thread. (Additionally, the He-Man library has 17 lines of
+individual thread. (Additionally, the He-Man library has 25 lines of
 code implementing generic routines for simulating blocking IO using
@@
 although it is unlikely to be any slower in practice. It has 10 states
-and is 240 lines of C.
-%This could be reduced substantially by allowing
-%conditionals outside of block tail position.
+and is 240 lines of C. The echo server was able to handle 10,000
+concurrent, occasionally transmitting, connections with no
+problem. With 10,000 clients connected the echo server used about 40
+MB of memory, which is almost entirely accounted for 4 KB buffer that
+each connection maintained for I/O.
 

### 2012-03-11 paper/paper.tex
Add a note in future work

@@
 
-We also plan to expand the language with the ability to declare functions and
-make non-external function calls within a thread. This will make code simpler
-and shorter by allowing the programmer to abstract away common functionality
-like read and write loops.
+XXX: I'm not really sure what we want to put here. I think maybe we
+shouldn't have a future work section. If we just want to talk about
+the deficencies of the thing we built, that should go somewhere else,
+maybe? It isn't really clear we have obvious future research
+directions to talk about.
 
-In the runtime, our current thread \t{struct}s are larger than necessary.
+We also plan to expand the language with the ability to declare
+functions and make non-external function calls within a thread. This
+will make the generated code smaller by allowing the programmer to
+abstract away common functionality like read and write loops.
+
+The runtime could be much more carefully tuned.
 Because the runtime allocates a \t{struct} on every thread spawn, we should

### 2012-03-11 paper/paper.tex
Work on related work section.

--- 2012-03-11 paper/paper.tex
@@
 and to demonstrate separating network logic and application logic, we
-built a very simple HTTP server. The network logic is written in
-He-Man and request parsing and response header building is written in
-C. The C request parsing is implemented as a function that takes a
-buffer and returns whether it forms a complete HTTP request and, if
-so, fills a buffer with the file name to serve. The entire server is
-68 lines of He-Man Haskell and 39 lines of C. It suffers from a number
-of major deficiencies, but is complete enough to benchmark. To make
-sure that it does not get away with sending less data than the servers
-it is being compared against, it sends some ``padding'' headers with
-useless data.
+built a simple (and bad) HTTP 1.0 server. The network logic is written
+in He-Man and request parsing is written in C. The C side is
+implemented as a function that takes a buffer and returns whether it
+forms a complete HTTP request and, if so, fills a buffer with the
+file name to serve. The entire server is 54 lines of He-Man Haskell and
+27 lines of C. It suffers from a number of major deficiencies, but is
+complete enough to benchmark.
 
-Our benchmark measurements are taken using weighttp, an http
-benchmarking tool associated with lighttpd
-\cite{weighttp}. Measurements were collected by requesting a file many
-times with a variable number of concurrent connections; the experiment
-was repeated with a 100 byte file (80000 requests), a 16 kilobyte file
-(30000 requests), and a 128 kilobyte file (8000 requests).
-weighttp was run with 4 threads making requests.
-Since the test repeatedly requests the same file, it should always be in
-the kernel's disk cache.
+Our benchmark measurements are taken using ab, the Apache HTTP server
+benchmarking tool \cite{ApacheAB}. Measurements were collected by
+having a variable number of client threads request a file 100 times;
+the experiment was repeated with a 16 kilobyte file and a 128 kilobyte
+file. We ran the tests against both our He-Man HTTP server and Apache
+2.2.20, installed with the default Ubuntu 11.10 settings, except for
+client, thread, and process limits being adjusted upwards. Since the
+He-Man runtime is currently only single-threaded, we ran the Apache
+web server pinned to a single processor. Since the test repeatedly
+requests the same file, it will always be in the kernel's disk cache,
+and thus this test is not disk bound.
 
-For our experiments, both the client and server ran Linux kernel
-2.6.32 on quad core 3.1 GHz Intel Core i5-2400 processors with 8
-gigabytes of RAM. The machines were connected over a 100 megabit
-Ethernet network.
-We ran the tests against our He-Man HTTP server, Apache 2.2.14 (installed
-with the default Ubuntu 10.04 LTS settings, except for client, thread,
-and process limits being adjusted upwards) and Lighttpd 1.4.26
-(installed with the default Ubuntu 10.04 LTS settings).
+For our experiments, the server ran Linux kernel 3.0.0 on a Lenovo
+Thinkpad W520 with a quad core 2.3 GHz Intel Core i7-2820QM and 8
+gigabytes of RAM, and the client ran on a Lenovo Thinkpad T500 with a
+Intel Core 2 Duo and 8 gigabytes of RAM.  % XXX: client specs
+The machines were connected by a gigabit Ethernet link.
+Unfortunately, ab is not particularly scalable, so we had trouble
+testing with more than 500 clients.
 
-Each trial was run 10 times; the plots show the arithmetic mean of the
-number of requests per second the servers handled; error bars show the
-standard deviation.
-
-Figure \ref{fig:graph100} compares the three servers while serving a
-very small (100 byte) file. Measurements with small files test the
-overhead of the web server more clearly than measurements with large
-files, which wind up testing how efficiently the kernel can transfer
-data. In this test, He-Man outperforms Apache and lighttpd across all
-of our measurements. This can be attributed to factors both in He-Man
-and in our server. He-Man imposes very low overhead and very low
-connection creation cost, which makes it easy to quickly handle many
-new connections. Our server is also much less featureful that Apache
-and lighttpd, which means it needs to do less. It is difficult to
-separate out these factors.
-
-Figure \ref{fig:graph128} and Figure \ref{fig:graph16} compares our
-server to Apache and lighttpd for 128 KB and 16 KB files. On these
-loads, our server performs nearly identically to Apache while being
-outperformed by the highly-turned lighttpd. In these tests, both
-Apache and He-Man begin to handle fewer requests per second above some
-number of concurrent requests (about 400 for the 16 KB file and 300
-for the 128 KB file). We investigated the He-Man httpd server's
-behavior in these tests, and found that the measured slowdown in
-requests/second occured when, after rapidly processing most of the
-requests in a test batch, a few requests would spend a very long time
-waiting to connect, causing a pause at the end of the run. During
-these pauses, both the client and the server were blocked in the
-kernel, waiting for network events. This strongly suggests to us that
-the problem is not with the He-Man compilation phase but that our web
-server would benefit from more careful tuning of network interactions
-(which may require improvements to the runtime).
+Figure \ref{fig:graph128} compares our server to Apache for a 128 KB
+file. Our server performs nearly identically to Apache on this load.
+Figure \ref{fig:graph16} compares our server to Apache for a 16 KB
+file. In this scenario, we substantially outperform Apache. We
+conjecture this is because with a smaller file size, overhead for
+connection setup begins to dominate, and connection setup in He-Man
+consists of a few memory allocations whereas connection setup under
+Apache may require spawning new OS level threads.
 
@@
 \centering
-\includegraphics[width=0.9\textwidth]{graph_100.pdf}
-\caption{Web server benchmark, 100 byte file.}
-\label{fig:graph100}
+\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
+\caption{Web server benchmark, 128 KB file.}
+\label{fig:graph128}
 \end{figure*}
@@
 \centering
-\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
-\caption{Web server benchmark, 128 KB file.}
-\label{fig:graph128}
+\includegraphics[width=0.9\textwidth]{graph_100.pdf}
+\caption{Web server benchmark, 100 B file.}
+\label{fig:graph100}
 \end{figure*}
@@
 %\cite{VanderwaartCrary} re bounded time
-%\cite{vonBehrenEtAl} re Capriccio
 %\cite{vonBehrenConditBrewer} re cooperative multithreading
 %\cite{LiZdancewic}
-%\cite{LauerNeedham} also intro
-%\cite{MarlowEtAl} also intro
-
-% Functional programming:
-%\cite{Vinoski}
-%\cite{Hellstrom}
-%\cite{BiagioniEtAl}
-%\cite{LiEtAl}
-%\cite{GraunkeEtAl}
-%\cite{Queinnec}
 
@@
 \section{Related work}
+
+\subsection{Hybrid concurrency models}
+
+The Lauer-Needham duality \cite{LauerNeedham} between threaded and event-driven
+concurrency has given rise to a number of programming models attempting to
+bridge the gap between these abstractions.
+
+Capriccio \cite{vonBehrenEtAl} is a user-level threading library which
+implements cooperative (non-preemptive) multithreading, and replaces blocking
+I/O primitives in \t{libc} with a wrapper for non-blocking I/O. Capriccio allows
+programmers to write C code in a threaded style, and performs a compiler
+analysis to statically allocate per-thread stack space. Capriccio's goals are
+similar to ours, but we feel that He-Man's approach of compiling to source is
+more portable and much simpler.
+
+Flux \cite{BurnsEtAl} is a domain-specific language intended, like He-Man, to
+handle concurrency in network applications separately from the rest of the
+application's logic. Flux allows programmers to declaratively specify how data
+flows through the server's components, and compiles to a threaded or
+event-driven runtime which implements this behavior concurrently. Flux source
+code cannot express loops, except for an implicit outermost loop, as a way of
+preventing deadlocks; He-Man allows more expressive control flow while making
+the same guarantee.
+
+Many functional programming languages present a native lightweight thread
+abstraction. Erlang \cite{VirdingEtAl} was designed for highly concurrent
+telecommunications applications, and easily scales to 20,000 clients
+\cite{Hellstrom}. Concurrent ML \cite{ReppyEtAl} adds concurrency primitives to
+Standard ML, implemented with front-end continuations, and a back-end
+synchronization protocol. The GHC Haskell runtime supports concurrency
+primitives \cite{LiEtAl} by scheduling lightweight threads onto a small number
+of kernel threads \cite{MarlowEtAl}.
+
+% XXX: remains unnecessarily general for us?
+
+He-Man is most similar to Li and Zdancewic's Haskell implementation of
+a hybrid event/thread programming model for network applications
+\cite{LiZdancewic}.  Their project has similar goals to ours, but is
+implemented entirely within Haskell. As a result, unlike He-Man, their
+framework requires entire applications to be written in Haskell, and
+incurs runtime overhead from the Haskell system.
+
+While He-Man's goals are similar to
+those of several other systems, we feel that it solves
+
+\subsection{Lightweight threads}
+
+% Functional programming:
+\cite{Vinoski}
+\cite{Hellstrom}
+FoxNet for Standard ML \cite{BiagioniEtAl}
+\cite{GraunkeEtAl}
+\cite{Queinnec}
 
@@
 
-A number of functional languages support lightweight threads with
-user-space scheduling, most notably Erlang, which was designed for
-highly concurrent telecommunications applications, and easily scales
-to 20,000 clients \cite{Hellstrom}. The Haskell runtime supports a
-variety of concurrency primitives which permit a lightweight threaded
-programming model \cite{LiEtAl}.
-
 %A number of custom network stack implementations exist for functional
 %languages, such as FoxNet for Standard ML \cite{BiagioniEtAl}, and
-
-He-Man is most similar to Li and Zdancewic's Haskell implementation of
-a hybrid event/thread programming model for network applications
-\cite{LiZdancewic}.  Their project has similar goals to ours, but is
-implemented entirely within Haskell. As a result, unlike He-Man, their
-framework requires entire applications to be written in Haskell, and
-incurs runtime overhead from the Haskell system.
-
-% XXX taken from old intro
-
-Languages like Erlang and Haskell provide lightweight threads via user-space
-schedulers. This affords even better performance, but remains unnecessarily
-powerful for this problem space, imposing runtime overhead on applications which
-don't require preemptable, independently scheduled threads \cite{Vinoski}.
 

### 2012-03-11 paper/paper.tex
A lot of editing.

@@
 \usepackage{verbatim}
+\usepackage{listings}
+\usepackage[usenames,dvipsnames]{color}
 
-\usepackage{listings}
 \lstset{
@@
   basewidth={0.5em,0.45em},
-%  commentstyle=\ttfamily\color{ForestGreen},
+  commentstyle=\ttfamily\color{ForestGreen},
   literate=*{+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
@@
 }
+\lstnewenvironment{code}{\lstset{frame=}}{}
 
-\lstnewenvironment{egcode}
-  {\lstset{frame=TRBL}}
-  {\lstset{frame=trbl}}
-
-\lstnewenvironment{code}
-  {\lstset{frame=}}
-  {}
-
+\newcommand{\prettybox}[1]{\fbox{\parbox{\linewidth}{#1}}}
 \newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
@@
 
-\newcommand{\prettybox}[1]{\fbox{\parbox{\linewidth}{#1}}}
-
-% A bunch of crazy BNF TeX I wrote once
+% For BNF grammars:
 \def\bnf#1{\prettybox{\halign{##\hfil& \hfil ##\hfil& \ ##\hfil\cr #1}}}
@@
 
-% XXX: We need to cite ElliottEtAl somewhere!
 % XXX: standardize on IO vs I/O
+% XXX: standardize on continuation-passing vs continuation passing
 % XXX: standardize on domain-specific vs domain specific
+% TODO: talk about type system, runtime improvements, optimizations
 
@@
 \maketitle
-
-% TODO: talk about type system, runtime improvements, optimizations
 
@@
 Network applications require a very constrained form of
-concurrency. In most network applications, I/O (both network
-and disk) latency dominates computation time, so each client's thread
+concurrency. In most network applications, network and disk
+I/O latency dominate computation time, so each client's thread
 of execution spends most of its time waiting on I/O. In particular, we
@@
 
-In the event loop model, network I/O is done using non-blocking
-\t{read} and \t{write} calls and all of the blocking for I/O is
-centralized in a single call to a multiplexing system call such as
-\t{select} or \t{epoll}. The multiplexing call will block until some 
+In the event-driven model, network I/O is performed with non-blocking
+\t{read} and \t{write} calls, and all blocking on I/O is consolidated
+in a single, centralized call to a multiplexing system call such as
+\t{select} or \t{epoll}. This multiplexing call blocks until some 
 I/O is ready to be performed, and indicates to the application
 which file descriptors are ready. To decide how to handle these 
-events, an application must, for each connection, have some sort of 
-state machine to track what state the connection is in and what 
-needs to happen next, as well as other local state.
+events, an application must, for each connection, maintain a
+state machine to track the connection's current and next states, as 
+well as any local data.
 
-One common scheme for doing this involves registering a function
-pointer with the event loop to be called on an event. Typically each
-connection is also associated with a structure of connection local
-data. Instead of making a blocking system call, then, an application
-registers an event with the event loop along with the function to call
-when it occurs, and then returns back to the loop.
+One common solution involves registering a function
+pointer with the event loop to be called on an event. Typically, each
+connection is also associated with a structure of connection-local
+data. Instead of making a blocking system call, an application
+registers an event with a function pointer to call once the event 
+occurs, and then returns back to the loop.
 
 The advantage of this scheme is that it is highly scalable and can be
-efficient: handling a new connection typically just involves
-allocating a new structure and registering it with the event loop. The
-disadvantage is that now network facing code needs to be scattered
-through many small functions, with all access to connection state
-mediated through a shared structure.
+efficient: handling a new connection requires only allocating a new 
+structure and registering it with the event loop. The disadvantage is 
+that network-facing code must be scattered across many small functions 
+which call each other only through function pointers, and all connection 
+state is accessed through a shared structure.
 
-The event loop model is then, essentially, built around closure-
-converted trampolining continuation passing style. These are well
-studied problems and can easily be done mechanically in the He-Man
-backend.
+However, as we discuss in the next section, these code transformations are
+familiar to functional programmers as closure conversion (to access connection
+state across function boundaries), continuation-passing style (via function
+pointers) with trampolining (yielding to the event loop after registering I/O).
+These are well-studied conversions which can be performed mechanically by the
+He-Man backend.
 
@@
 
-It is important to provide some sort of type discipline for He-Man
-programs. Providing a type system in the front end allows us to
-guarantee that the C we output does not contain type errors, vastly
-simplifying development. We also take advantage of this to provide
-type abstractions above and beyond C's limited type system. % XXX: bad wording
-For example, He-Man's type system contains an abstract
-$\b{FD}$ type that is distinct from the type of integers.
+We provide a simple type system for He-Man programs which guarantees that we
+generate C code which is free of type errors. We take advantage of this to
+provide abstractions not present in C, e.g., the \b{FD} type of file descriptors
+is distinct from integers in He-Man, while C draws no distinction between them.
 
-We offload the work of type checking He-Man programs to the Haskell type
-checker. Our method for doing this is fairly standard, following Leijen and
-Meijer \cite{LeijenMeijer} in using phantom types to enforce typing.
+We type-check He-Man programs with the Haskell type checker, following Leijen
+and Meijer \cite{LeijenMeijer} in using phantom types. A fragment of the code
+implementing the He-Man type system is shown in Figure \ref{fig:types}. Types
+and expressions are internally represented by the datatypes \t{IExpr} and
+\t{IType}, and typed expressions are represented with the \t{Expr} data type
+constructor, which is a trivial wrapper around \t{IExpr} parameterized
+over a phantom type. Operations on typed expressions have type signatures that
+restrict them to the proper types, and discard the wrapper to act on the
+\t{IExpr} underneath.
 
-A fragment of the code implementing the He-Man type system is shown in
-Figure \ref{fig:types}. The internal type and expression
-representations are the datatypes \t{IExpr} and \t{IType}, and
-typed expressions are represented with the \t{Expr} data type
-constructor, which is a trivial wrapper around \t{IExpr} parameterized
-over a phantom type. Operations on typed expressions have type
-signatures that restrict them to the proper types, and then peel away
-the wrapper to work on the \t{IExpr} underneath.
-
-% XXX: say something about GADTs??
-Types are represented by the \t{Type} GADT that is also parameterized
-over a phantom type representing the underlying type. Expression
-builder functions will convert \t{Type} values are to \t{Itype} using
-the \t{toItype} function, which does a simple pass-through conversion.
-The decision to make \t{Type} a GADT instead of a simple wrapper with
-constructor functions in the same manner as \t{Expr} is for purely
-aesthetic reasons: we felt that type annotations in He-Man code were
-more clearly readable if they were capitalized, which required that
-they be datatype constructors.
-\footnote{The authors believe that this may be the worst reason to
-  choose to use GADTs to date.}
+Types are represented by the \t{Type} GADT, which is parameterized over a
+phantom type representing the underlying type. When building expressions, we
+convert \t{Type} values to \t{IType} with the \t{toItype} function. The decision
+to make \t{Type} a GADT instead of a simple wrapper with constructor functions
+in the same manner as \t{Expr} is purely aesthetic: we felt that type
+annotations in He-Man code were more clearly readable if they were capitalized,
+which required them to be datatype constructors.
+%\footnote{The authors believe that this may be the worst reason to
+%  choose to use GADTs to date.}
 
@@
 
-Type checking foreign function calls and thread declaration and
-spawning is somewhat more interesting.  For foreign function calls,
-our main concern was being able to write a single wrapper for call
-that handles a varying number of typed arguments. We want more than
-this, however: we want to ensure that both the code for a declared
-thread uses its arguments properly at the declared types code and that
-spawning threads provides arguments of the correct types.
-% and maybe a bit novel? I dunno.
+Type-checking foreign function calls and thread declaration and spawning is
+somewhat more interesting. We wanted a single wrapper for foreign function calls
+to handle a varying number of typed arguments; we also wanted to ensure that
+thread spawn sites pass in values of the correct types, and declared threads use
+their arguments in type-safe ways.
 
@@
 class \t{ArgPacket}. \t{ArgPacket} is parameterized over related
-tuples \footnote{Technically speaking, the unary \t{ArgPacket}s are
+tuples\footnote{Technically speaking, the unary \t{ArgPacket}s are
 not actually tuples.} of expressions and types. The idea is that
@@
 \t{exps} and \t{tys} are tuples of \t{Expr}s and \t{Type}s,
-respectively, the tuples have the same number of elements, and for
+respectively; the tuples have the same number of elements; and for
 each pair of corresponding elements in the two tuples, the \t{Type}
 represents the type of the \t{Expr} (that is, they have the same
-phantom type argument). To enforce this one-to-one correspondence and
-to allow the Haskell inferencer to use the information it provides, we
-use functional dependencies % XXX: cite, or something?
-to declare that both \t{exps} and \t{tys} are uniquely determined by
-the other.
+phantom type argument).
 
-Then, by parameterizing the wrapper functions for foreign function
-calls and thread declaration and spawning over \t{ArgPacket exps tys},
-we get the desired typing guarantees.
-% XXX: is this explained fully?
+We use functional dependencies to declare that both \t{exps} and \t{tys} are
+uniquely determined by the other, which enforces this one-to-one correspondence
+and allows the Haskell inferencer to use provided typing information. We obtain
+our desired typing guarantees by parameterizing the wrapper functions for
+foreign function calls and thread declaration/spawning over \t{ArgPacket exps tys}.
 
@@
 
-The He-Man runtime is written in C and provides a state machine based
-thread scheduling model. In the view of the runtime, a thread consists
-of some bookkeeping state, some private state, and the next
-continuation function the thread is to run. The main part of the
-runtime's state consists of the runqueue, which contains a list of all
-currently runnable threads.
+The He-Man runtime is a multithreaded C program which provides a generic state
+machine thread scheduling model. To the runtime, a thread consists of some
+bookkeeping state, some private state, and the thread's next continuation
+function. The runtime's state is primarily in the runqueue, which contains a
+list of all currently runnable threads.
 
-Continuation functions take a pointer to the thread information,
-perform some computation (such as making non-blocking I/O calls),
-update the continuation field in the thread data, and return whether
-the thread is still runnable; to block, a continuation function
+Each continuation function takes a pointer to the thread information,
+performs some computation (e.g., making non-blocking I/O calls),
+updates the continuation field in the thread data, and returns whether
+the thread is still runnable. To block, a continuation function
 registers itself as blocking on an event and returns false; to exit,
@@
 function until it returns false. To prevent a thread that does not
-need to block from starving the rest of the system, after some number
-of invocations of the continuation (currently 50), the thread is
-paused and put on the back of the runqueue.
+need to block from starving the rest of the system, after a continuation is
+invoked some number of times (currently 50), the thread is paused and moved to
+the back of the runqueue.
 
-The He-Man runtime main loop is very simple, then: each worker thread
+The He-Man runtime main loop is very simple: each worker thread
 loops indefinitely, and on each iteration it removes the thread at the
 head of the runqueue (if it exists) and runs it, and polls for I/O
-events. A global runqueue is shared between threads; a more clever
-runtime could have per-worker-thread runqueues and do some sort of
+events. A global runqueue is shared between threads; a fancier 
+implementation could maintain per-worker-thread runqueues and use
 work stealing.
@@
 channel). Events are level-triggered: blocking on an event will return
-whenever the event ``is ready'', as opposed to when the event
-``becomes ready'' as in edge-triggered (using edge-triggered semantics
-causes extra difficulties in an environment with multiple threads
-processing events).
+whenever the event ``is ready'', not when the event ``becomes ready''
+(edge-triggered). We make this design decision because edge-triggered semantics
+cause difficulties in an environment where multiple threads are processing
+events.
 
@@
 
-The actual polling for I/O events is done by calling \t{epoll}
-(although it would not be difficult to adapt it to use BSD's
-\t{kqueue} or the more standard \t{poll} or \t{select}).  For each
-event that has been returned by \t{epoll}, if a thread is waiting on
-the event, the thread is added to the runqueue. If the the runqueue is
-empty, the call to \t{epoll} blocks, otherwise it is called with a
-timeout of zero seconds, to simply poll for events.
+The actual polling for I/O events is done by calling \t{epoll}, and could be
+easily adapted to BSD's \t{kqueue}, or the more standard \t{poll} or \t{select}.
+For each event that \t{epoll} returns, if a thread is waiting on that event, the
+thread is added to the runqueue. If the the runqueue is empty, the call to
+\t{epoll} blocks; otherwise it is called with a timeout of zero seconds, to
+simply poll for events.
 
@@
 To evaluate He-Man, we built several network applications with He-Man
-and compared them to similar network applications built in C.  We also
-discuss He-Man's memory overhead.
+and compared them to similar network applications built in C.
 
@@
 
-To compare the ease of writing event based network applications, we
-built simple echo servers using both He-Man and C. The C version was
-written in as an event loop driven explicit state machine using the
-He-Man runtime. The C version, which is fairly tightly written and
-which uses the He-Man runtime in its capacity as a somewhat generic
-event based IO framework, is 98 lines of source, according to
-SLOCCount. It is fairly readable, but its structure is obscured by
-being split into states and the need to explicitly handle read and
-write returning \t{EAGAIN} and write returning short. The state
-machine has a total of five states. The He-Man version is 33 lines
-long and is a very clear representation of the control flow of each
-individual thread. (Additionally, the He-Man library has 25 lines of
-code implementing generic routines for simulating blocking IO using
-nonblocking IO and wait.) The total line counts, however, somewhat
-understate the difference between the versions, since the boilerplate
-for setting up connections is about the same for both versions. When
-just looking at the core read/write loop, the difference is much more
-apparent: the C version is 38 lines of code, and the He-Man version is
-5.
+To compare the ease of writing event-driven network applications, we wrote
+simple echo servers in both He-Man and C. The C version was written as an
+explicit state machine using the He-Man runtime in its capacity as a generic
+event-driven I/O framework.  This code is fairly tightly written, and consists
+of 98 lines of source code, according to SLOCCount.  It is fairly readable, but
+its structure is obscured by being split into states, and the need to explicitly
+handle read and write returning \t{EAGAIN}, or write returning short. The state
+machine has a total of five states.
 
-The generated code is substantially less tight than the C version,
-although it is unlikely to be any slower in practice. It has 10 states
-and is 240 lines of C. The echo server was able to handle 10,000
-concurrent, occasionally transmitting, connections with no
-problem. With 10,000 clients connected the echo server used about 40
-MB of memory, which is almost entirely accounted for 4 KB buffer that
-each connection maintained for I/O.
+The He-Man version is 33 lines long and clearly represents the control flow of
+each individual thread. (This does not include the He-Man library, which has 25
+lines of code implementing generic routines for simulating blocking IO using
+nonblocking IO and wait.) The total line counts, somewhat understate the
+difference between the versions, since the boilerplate for setting up
+connections is about the same for both versions. When just looking at the core
+read/write loop, the difference is much more apparent: the C version is 38 lines
+of code, and the He-Man version is 5.
+
+The generated code is substantially longer than the C version, although not any
+slower in practice; it has 10 states in 240 lines of C. This echo server easily
+handles 10,000 concurrent, occasionally transmitting, connections.  With 10,000
+clients, the echo server uses about 40 MB of memory, almost entirely consisting
+of a 4 KB I/O buffer for each connection.
 
@@
 
-To test the suitability of our system for more complicated programs
-and to demonstrate separating network logic and application logic, we
-built a very simple HTTP server. The network logic is written in
-He-Man and request parsing and response header building is written in
-C. The C request parsing is implemented as a function that takes a
-buffer and returns whether it forms a complete HTTP request and, if
-so, fills a buffer with the file name to serve. The entire server is
-68 lines of He-Man Haskell and 39 lines of C. It suffers from a number
-of major deficiencies, but is complete enough to benchmark. To make
-sure that it does not get away with sending less data than the servers
-it is being compared against, it sends some ``padding'' headers with
-useless data.
+To test the suitability of our system for more complicated programs and to
+demonstrate separating network logic from application logic, we also built a
+very simple HTTP server. The network logic is written in He-Man; request parsing
+and response header building are implemented in C. Request parsing is handled by
+function which takes a buffer and returns whether it forms a complete HTTP
+request; if so, it fills a buffer with the file name to serve. The entire server
+is 68 lines of He-Man and 39 lines of C. Although it is very incomplete as an
+HTTP server, it is sufficient for benchmarking purposes. We add some useless
+headers for padding, to ensure that it sends the same amount of data as the
+complete servers we benchmark against.
 
-Our benchmark measurements are taken using weighttp, an http
-benchmarking tool associated with lighttpd
-\cite{weighttp}. Measurements were collected by requesting a file many
-times with a variable number of concurrent connections; the experiment
-was repeated with a 100 byte file (80000 requests), a 16 kilobyte file
-(30000 requests), and a 128 kilobyte file (8000 requests).
-weighttp was run with 4 threads making requests.
-Since the test repeatedly requests the same file, it should always be in
-the kernel's disk cache.
+Our benchmark measurements are taken using weighttp \cite{weighttp}, an http
+benchmarking tool associated with lighttpd.  Measurements were collected by
+requesting a file many times with a variable number of concurrent connections;
+the experiment was repeated with a 100 byte file (80,000 requests), a 16 KB file
+(30,000 requests), and a 128 KB file (8,000 requests). weighttp was run with 4
+threads making requests. Since the test repeatedly requests the same file, it
+should always be in the kernel's disk cache.
 
-For our experiments, both the client and server ran Linux kernel
-2.6.32 on quad core 3.1 GHz Intel Core i5-2400 processors with 8
-gigabytes of RAM. The machines were connected over a 100 megabit
-Ethernet network.
-We ran the tests against our He-Man HTTP server, Apache 2.2.14 (installed
-with the default Ubuntu 10.04 LTS settings, except for client, thread,
-and process limits being adjusted upwards) and Lighttpd 1.4.26
-(installed with the default Ubuntu 10.04 LTS settings).
+For our experiments, both the client and server ran Linux kernel 2.6.32 on quad
+core 3.1 GHz Intel Core i5-2400 processors with 8 gigabytes of RAM. The machines
+were connected over a 100 megabit Ethernet network. We ran the tests on our
+He-Man HTTP server, Apache 2.2.14 (installed with the default Ubuntu 10.04 LTS
+settings, except for client, thread, and process limits being adjusted upwards),
+and Lighttpd 1.4.26 (installed with the default Ubuntu 10.04 LTS settings).
 
-Each trial was run 10 times; the plots show the arithmetic mean of the
-number of requests per second the servers handled; error bars show the
-standard deviation.
+Each trial was run 10 times. The plots show the arithmetic mean of the number of
+requests per second handled by each server; error bars show the standard
+deviation.
+
+% XXX work on the below paragraphs
 
@@
 
-The per-thread memory overhead imposed by He-Man is very small: it
-consists of a thread id, list pointers for the scheduling queue, the
-next state to enter, a pointer to the most recently received event,
-and list heads and tails for lists of events owned by the thread and
-buffers owned by the thread (the latter two are so that we can
-automatically reclaim these resources when the thread exits). On a
-64-bit system, the total overhead comes out to 64 bytes, and could
-easily be reduced further (e.g., by using singly-linked lists).  For
-comparison, the default thread stack size under NPTL on Linux is 2
-megabytes.
-
-% see how many threads we can spawn with Memory.hs
-
-% Related work:
-%\cite{VanderwaartCrary} re bounded time
-%\cite{vonBehrenConditBrewer} re cooperative multithreading
-%\cite{LiZdancewic}
-
-% Haskell EDSLs:
-%\cite{Gill}
-%\cite{ElliottEtAl}
-%\cite{LeijenMeijer}
+The per-thread memory overhead imposed by He-Man is very small: it consists of a
+thread id, list pointers for the scheduling queue, the next state to enter, a
+pointer to the most recently received event, and list heads and tails for lists
+of events owned by the thread and buffers owned by the thread.  (The event lists
+allow us to automatically reclaim resources when a thread exits). On a 64-bit
+system, the total overhead comes out to 64 bytes, and could be reduced further.
+For comparison, the default thread stack size under NPTL on Linux is 2 MB.
 
@@
 
+% Related work:
+%\cite{VanderwaartCrary} re bounded time
+%\cite{vonBehrenConditBrewer} re cooperative multithreading
+%\cite{LiZdancewic}
+
+% Haskell EDSLs:
+%\cite{Gill}
+%\cite{ElliottEtAl} CITE THIS
+%\cite{LeijenMeijer}
+
 % Functional programming:
@@
 \section{Future work}
-
-XXX: I'm not really sure what we want to put here. I think maybe we
-shouldn't have a future work section. If we just want to talk about
-the deficencies of the thing we built, that should go somewhere else,
-maybe? It isn't really clear we have obvious future research
-directions to talk about.
 

### 2012-03-11 paper/paper.tex
Finished related work, conclusion.

@@
 
-% XXX: standardize on IO vs I/O
-% XXX: standardize on continuation-passing vs continuation passing
-% XXX: standardize on domain-specific vs domain specific
-% TODO: talk about type system, runtime improvements, optimizations
+%There exist C/C++ libraries for event loops\cite{libevent}, but none of these
+%avoid the state machine paradigm.
 
 %XXX: new title?
-\title{He-Man: A domain specific language for lightweight concurrency}
+\title{He-Man: A domain-specific language for lightweight concurrency}
 %\subtitle{I said hey, what's goin' on}
@@
 
-This paper describes the design and Haskell implementation of He-Man, a domain
-specific language for network application programming which presents a
+This paper describes the design and Haskell implementation of He-Man, a
+domain-specific language for network application programming which presents a
 sequential thread-like abstraction, but compiles to performant event-driven C
@@
 the same problem domain of scalable network applications; their implementation
-draws upon the concurrency and synchronization support in the GHC runtime.
+draws upon concurrency and synchronization support in the GHC runtime.
 
@@
 code. This abstraction, He-Man: Haskell Event Manager, Apropos Networking, is a
-compiled domain specific language embedded in Haskell, which offers a small
+compiled domain-specific language embedded in Haskell, which offers a small
 monadic language supporting thread spawning, blocking events, and external
@@
 ``ready''. Note that \b{Call} is for calling foreign functions; He-Man
-does not have its own notion of function.
+does not have its own notion of function.\footnote{Adding functions or
+observable sharing to He-Man could reduce output code size, but would have
+little impact on performance.}
 
@@
 \end{code}
-\caption{He-Man expression typing} % XXX
+\caption{Implementation of He-Man type system.}
 \label{fig:types}
@@
 \end{code}
-\caption{He-Man argument packets} % XXX
+\caption{Implementation of He-Man \t{ArgPacket}s.}
 \label{fig:arg-packets}
@@
 blocks. Threads are identified with their first block.
-% TODO discuss the transformations
+
+TODO here we will discuss the transformations that occur
 
@@
 
-Once our 
-% TODO fuseBlocks pass
+TODO here we discuss the fuseBlocks pass
 
@@
 events: socket/pipe file I/O and asynchronous He-Man channels.  The
-runtime has partial support for Linux Kernel Asynchronous IO for file
-IO, using a Linux \t{eventfd} to allow \t{epoll} to wait on AIO
+runtime has partial support for Linux Kernel Asynchronous I/O for file
+I/O, using a Linux \t{eventfd} to allow \t{epoll} to wait on asynchronous I/O
 events. Adding new types of events that an application requires would
@@
 each individual thread. (This does not include the He-Man library, which has 25
-lines of code implementing generic routines for simulating blocking IO using
-nonblocking IO and wait.) The total line counts, somewhat understate the
+lines of code implementing generic routines for simulating blocking I/O using
+nonblocking I/O and wait.) The total line counts, somewhat understate the
 difference between the versions, since the boilerplate for setting up
@@
 
-\begin{figure*}[!ht]
+\begin{figure*}[htbp]
 \centering
@@
 \end{figure*}
-\begin{figure*}[!ht]
+\begin{figure*}[htbp]
 \centering
@@
 
-% XXX: remains unnecessarily general for us?
+Li and Zdancewic \cite{LiZdancewic} have developed a hybrid concurrency monad in
+Haskell for building scalable network applications. Their implementation uses
+the CPS monad to represent computations as continuations in a ready queue, and
+gives the illusion that I/O is blocking in each thread, while it is actually
+handled with foreign calls to \t{epoll}. This concurrency monad is executed in
+Haskell, unlike He-Man programs, which are compiled for speed and portability
+\cite{ElliottEtAl}.
 
-He-Man is most similar to Li and Zdancewic's Haskell implementation of
-a hybrid event/thread programming model for network applications
-\cite{LiZdancewic}.  Their project has similar goals to ours, but is
-implemented entirely within Haskell. As a result, unlike He-Man, their
-framework requires entire applications to be written in Haskell, and
-incurs runtime overhead from the Haskell system.
+\subsection{Functional network programming}
 
-While He-Man's goals are similar to
-those of several other systems, we feel that it solves
+In addition to the concurrency models mentioned above, a number of functional
+languages have network libraries specifically tailored to concurrency issues,
+like Erlang's TCP/IP suite \cite{ParisEtAl}, and the FoxNet library for Standard
+ML \cite{BiagioniEtAl}, which uses continuations to implement coroutines for
+cooperative multithreading.
 
-\subsection{Lightweight threads}
+It is important to note that in environments providing cooperative
+multithreading, like Capriccio, FoxNet, and He-Man, while the runtime guarantees
+that no thread monopolizes execution while waiting for I/O, it is certainly
+possible for one thread to monopolize execution by running indefinitely.
+Bounding execution time is a difficult problem which has been studied in the
+context of grid computing \cite{VanderwaartCrary}. The He-Man compiler
+guarantees that no code block contains an infinite loop; unbounded execution can
+only occur in calls to external functions.
 
-% Related work:
-%\cite{VanderwaartCrary} re bounded time
-%\cite{vonBehrenConditBrewer} re cooperative multithreading
-%\cite{LiZdancewic}
+The Scheme community often uses continuations to develop web applications which
+robustly handle client state despite the statelessness of HTTP. This can be
+accomplished either by CPSing a direct-style program \cite{GraunkeEtAl} or by
+modifying a HTTP server to explicitly store continuations \cite{Queinnec}.
 
-% Haskell EDSLs:
-%\cite{Gill}
-%\cite{ElliottEtAl} CITE THIS
-%\cite{LeijenMeijer}
+\section{Conclusion}
 
-% Functional programming:
-\cite{Vinoski}
-\cite{Hellstrom}
-FoxNet for Standard ML \cite{BiagioniEtAl}
-\cite{GraunkeEtAl}
-\cite{Queinnec}
+The Lauer-Needham duality suggests that the performance of event-driven
+concurrent code should be achievable with simpler threaded semantics. We propose
+a model of network application programming which completely separates I/O logic,
+which requires a tuned concurrent implementation, from largely single-threaded
+protocol and implementation details.
 
-The C10K problem is particularly evident in the context of HTTP
-servers, which routinely maintain many concurrent connections. As we
-discussed in Section \ref{sec:evaluation}, Apache does not scale
-particularly well. The popular Lighttpd and Nginx HTTP server projects
-were both started with the explicit goal of outperforming Apache with
-respect to the C10K problem \cite{Lighttpd,Nginx}, and both use
-event-based architectures similar to He-Man.
-
-%A number of custom network stack implementations exist for functional
-%languages, such as FoxNet for Standard ML \cite{BiagioniEtAl}, and
-
-There exist C/C++ libraries for event loops\cite{libevent}, but none of these
-avoid the state machine paradigm.
-
-%% XXX: needs serious work
-\section{Future work}
-
-We also plan to expand the language with the ability to declare
-functions and make non-external function calls within a thread. This
-will make the generated code smaller by allowing the programmer to
-abstract away common functionality like read and write loops.
-
-The runtime could be much more carefully tuned.
-Because the runtime allocates a \t{struct} on every thread spawn, we should
-replace our \t{malloc} calls with calls to a custom slab allocator.
-
-We also hope to implement observable sharing, as in \cite{Gill}, to allow the
-same thread to have distinct spawn sites without code duplication. This does not
-add inefficiency, but does increase the size of our output code.
-
-\section{Conclusions}
-Expressing network application logic as a compiled, embedded, domain
-specific language is an elegant and practical way to build scalable
-network applications without having to manually write state-machine
-based event-loop code. Embedding the language in Haskell allows
-powerful metaprogramming and allowed the language to be developed with
-a fraction of the effort of building a full language. Compiling to C
-allows C programmers to write parts of their system (for example, the
-caching layer) in C while writing the network logic in
-straightforward, straight-line He-Man code.
-
-% XXX
-
-We identify our contributions as:
-* We present the idea of generating code for network facing logic from a
-description written in a DSL while building the rest of the application in
-C. In a way, the DSL could be considered as a method for programmatically
-generating event driven frameworks customized for the application.
-* We observe that converting a traditional block-structured imperative
-program to a form amenable for use in an event-driven system is analogous
-to CPS conversion and thus easily done automatically. (Is this actually a
-contribution, or is it obvious and been said a million times before?)
-* We present a case study of compiled embedded domain specific language
-design and implementation
-* We demonstrate that it is practical to build efficient network
-applications using our system
-
-\cite{*}
+We present He-Man, a domain-specific language for expressing network application
+I/O in a straightforward threaded abstraction, which compiles to efficient
+event-driven C code. We demonstrate that it is practical to build efficient
+network applications with a combination of He-Man and C code, and unlike other
+hybrid concurrency systems, the He-Man runtime is very compact and easily
+modified.
 

### 2014-04-02 paper/paper.tex

@@
 \usepackage[usenames,dvipsnames]{color}
+
+\newcommand{\prettybox}[1]{\fbox{\parbox{\linewidth}{#1}}}
+\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
+\renewcommand{\t}{\texttt}
+\renewcommand{\b}{\textbf}
+\renewcommand{\i}{\textit}
 
@@
 }
-\lstnewenvironment{code}{\lstset{frame=}}{}
-
-\newcommand{\prettybox}[1]{\fbox{\parbox{\linewidth}{#1}}}
-\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
-\renewcommand{\t}{\texttt}
-\renewcommand{\b}{\textbf}
-\renewcommand{\i}{\textit}
+%\lstnewenvironment{code}{\lstset{frame=}}{}
+\newenvironment{code}{\endgraf\small\verbatim}{\endverbatim}
 
@@
 
-%XXX: new title?
 \title{He-Man: A domain-specific language for lightweight concurrency}
@@
 about, it is considerably less scalable. We observe that these paradigms are not
-only interchangeable, but primarily confined to an application's core control
-flow; we propose a hybrid model in which programmers express this core logic at
-a high level of abstraction, yet are free to implement the remainder of the
-application at a low level.
+only interchangeable, but primarily confined to an application's core I/O
+control flow logic. We propose a hybrid model in which programmers express this
+core logic in a high-level concurrency abstraction, yet are free to implement
+the remainder of the application in a generic fashion.
 
@@
 
-Modern network servers, like HTTP, IRC, and multiplayer games, need to support
+Modern network servers, as for HTTP, IRC, and multiplayer games, need to support
 many simultaneous, possibly idle, connections---this is known as the C10K
@@
 These are well-studied conversions which can be performed mechanically by the
-He-Man backend.
+He-Man back-end.
 
@@
 create our concrete wrappers. (\t{.=.} is a wrapper for \b{Assign} 
-that takes a \t{Prog Expr} as its second argument and the \t{sock\_*} 
+that takes a \t{Prog Expr} as its second argument, and the \t{sock\_*} 
 functions invoke \b{Call} wrappers on the appropriate network calls.)
@@
 
-TODO here we will discuss the transformations that occur
+\b{If} statements are the most interesting case. We place the subsequent
+statements in a separate block, and process the two bodies of the \b{If}
+statement, placing each in a separate block, and terminating each with a tail
+\b{Goto} to the subsequent statements. We then process the statements preceding
+the \b{If}, terminating their block with a tail-position \b{If} which jumps to
+either branch. \b{While} statements are handled in a similar fashion.
 
-%As an illustrative example, consider encountering an \b{If} statement in our
-%backwards pass, where we have access to the translation of the later
-%statements, as well as the earlier statements. Morally, we would like to
-%translate this to a conditional jump to the translation of either the first or
-%second arm, each of which is terminated by a jump to the later statements.
-%(Note that \t{flattenPrgm} leaves all optimization to the next pass.)
-
-%We package up the later statements in a new block, and then translate the
-%preceding statements, terminated by a tail-position conditional \b{If}. We also
-%translate each arm of the original \b{If}, terminated by a tail-position
-%\b{Goto} to the block containing the later statements. Importantly, all three
-%of these recursive translations may generate new blocks, which will be
-%collected automatically by the \t{Writer} monad.
-
-%Because blocking I/O may only happen in the main loop, we also terminate a
-%block after any blocking I/O call, by inserting a \b{GotoWait} to the block
-%containing the later statements. \b{GotoWait} is identical to \b{Goto} except
-%to signal to later passes that the thread \emph{must} yield to the main loop at
-%this location, and thus the goto must not be optimized away.
-% TODO for concurrency-related reasons
+We translate thread-blocking operations ($\b{Wait}~e$ in the front-end language)
+to a blocking jump ($\b{GotoWait}~e$) to a new code block containing the next
+statements to execute. \b{GotoWait} is similar to \b{Goto}, except that it
+contains an event to wait on, and is guaranteed not to be optimized away by
+later passes. Thread spawns are translated into a \b{Spawn} statement which
+takes a list of arguments and a destination label; the thread's code, which is
+part of the front-end \b{Spawn} statement, is processed separately.
 
@@
 
-At this point, we need to eliminate spurious control flow edges from our state
-machine in order to enable further block elimination. To do this, we perform
-block-local constant propagation and folding, and dead code elimination. At face
-value, it is not important for us to optimize straight-line code: our output is
-further optimized by a C compiler, and execution time is dominated by I/O, not
-computation. However, simple constant folding eliminates many \t{if} statements
-%TODO examples, justification
+At this point, we would like to eliminate spurious control flow from our state
+machine in order to enable further block elimination. Our \t{optimize} pass
+performs block-local constant propagation and folding, symbolic constraint
+propagation, and dead code elimination. At face value, it is not important for
+us to optimize straight-line code: our output is further optimized by a C
+compiler, and execution time is dominated by I/O, not computation. However,
+these simple optimizations can statically resolve many \b{If} tails generated by
+the previous pass, resulting in substantially less control flow.
 
-TODO here we discuss the fuseBlocks pass
+Consider, for example, the following code snippet from our library function
+\t{full\_write}:
+\begin{smallverbatim}
+full_write (fd, e) buf size = do
+  amt_written <- var "total_written" Int 0
+  failed <- var "failed" Int 0
+  while (amt_written .< size .&& failed .== 0) $ do
+    -- write amt bytes to buffer
+    ifE (amt .<= 0) (failed .= 1) $ -- amt_written += amt
+\end{smallverbatim}
+After only \t{simplifyJumps}, this function compiles to eight blocks, including
+the following:
+\begin{smallverbatim}
+Block 9 {
+  IInt total_written_16 = 0
+  IInt failed_17 = 0
+  if (total_written_16 < file_size_10 && failed_17 == 0)
+    goto 8
+  else
+    goto 7
+}
+Block 40 {
+  failed_17 := 1
+  if (total_written_16 < file_size_10 && failed_17 == 0)
+    goto 8
+  else
+    goto 7
+}
+\end{smallverbatim}
+Block 9 is executed the first time the \t{while} loop is encountered; block 40
+is executed on the \t{amt .<= 0} failure condition, and jumps back to the top
+of the \t{while} loop. But constant folding shows that the loop is always
+executed at least once, so long as the file size is positive, and is always
+exited when the failure condition is tripped. After optimization, we are left
+with:
+\begin{smallverbatim}
+Block 9 {
+  IInt total_written_16 = 0
+  IInt failed_17 = 0
+  if (0 < file_size_10)
+    goto 8
+  else
+    goto 7
+}
+Block 40 {
+  failed_17 := 1
+  goto 7
+}
+\end{smallverbatim}
+While \t{optimize} does not result in much smaller code, in practice it does
+remove many edges from the control flow graph.
+
+For performance reasons, we would like to yield to the event loop sparingly,
+which amounts to increasing the length of each block as much as possible. Our
+final optimization pass, \t{fuseBlocks}, hoists certain blocks into their
+callers to eliminate unnecessary yields. To facilitate hoisting, we generalize
+the grammar to allow statements in each branch of an \b{If} tail.
+
+We hoist blocks which are invoked in a unique location, and those which end in a
+thread \b{Exit} or blocking I/O (\b{GotoWait}). The former eliminates yields
+without increasing code size; the latter allows threads to register I/O or free
+resources immediately. In the \t{full\_write} example, block 40 is only reachable
+from the failure \t{if}, and is eliminated entirely, while the thread exit
+located after the \t{while} loop is hoisted into block 9.
+\begin{smallverbatim}
+Block 9 {
+  IInt total_written_16 = 0
+  IInt failed_17 = 0
+  if (0 < file_size_10)
+    goto 8
+  else
+    IInt tmp_19 = close(tmp_9)
+    exit
+}
+\end{smallverbatim}
+After optimization, \t{full\_write} is reduced from eight to four blocks.
 
@@
 lines of code implementing generic routines for simulating blocking I/O using
-nonblocking I/O and wait.) The total line counts, somewhat understate the
+nonblocking I/O and wait.) The total line counts understate the
 difference between the versions, since the boilerplate for setting up
@@
 
-% XXX work on the below paragraphs
+% XXX work on the below paragraphs?
 
@@
 
-It is important to note that in environments providing cooperative
-multithreading, like Capriccio, FoxNet, and He-Man, while the runtime guarantees
-that no thread monopolizes execution while waiting for I/O, it is certainly
-possible for one thread to monopolize execution by running indefinitely.
-Bounding execution time is a difficult problem which has been studied in the
-context of grid computing \cite{VanderwaartCrary}. The He-Man compiler
-guarantees that no code block contains an infinite loop; unbounded execution can
-only occur in calls to external functions.
+Cooperative multithreading (as in Capriccio, FoxNet, and He-Man) is more
+performant in this domain than preemption, because it avoids unnecessary context
+switches and executes only unblocked threads. On the other hand, the scheduler
+is only invoked when a thread voluntarily yields, so a thread may monopolize
+execution time if it never yields. Placing bounds on execution time is a
+difficult problem in general, and has been studied in the context of grid
+computing \cite{VanderwaartCrary}. The He-Man compiler guarantees that no code
+block contains an infinite loop; unbounded execution can only occur in calls to
+external functions.
 
